### 2015-08-24 flux-entropy-paper/flux-entropy-paper.tex
committed a paper template

\documentclass{sigchi}

%% EXAMPLE BEGIN -- HOW TO OVERRIDE THE DEFAULT COPYRIGHT STRIP -- (July 22, 2013 - Paul Baumann)
% \toappear{Permission to make digital or hard copies of all or part of this work for personal or classroom use is      granted without fee provided that copies are not made or distributed for profit or commercial advantage and that copies bear this notice and the full citation on the first page. Copyrights for components of this work owned by others than ACM must be honored. Abstracting with credit is permitted. To copy otherwise, or republish, to post on servers or to redistribute to lists, requires prior specific permission and/or a fee. Request permissions from permissions@acm.org. \\
% {\emph{CHI'14}}, April 26--May 1, 2014, Toronto, Canada. \\
% Copyright \copyright~2014 ACM ISBN/14/04...\$15.00. \\
% DOI string from ACM form confirmation}
%% EXAMPLE END -- HOW TO OVERRIDE THE DEFAULT COPYRIGHT STRIP -- (July 22, 2013 - Paul Baumann)

% Arabic page numbers for submission.  Remove this line to eliminate
% page numbers for the camera ready copy

\pagenumbering{arabic}

% Load basic packages
\usepackage{balance}  % to better equalize the last page
\usepackage{graphics} % for EPS, load graphicx instead 
%\usepackage[T1]{fontenc}
\usepackage{txfonts}
\usepackage{times}    % comment if you want LaTeX's default font
\usepackage[pdftex]{hyperref}
% \usepackage{url}      % llt: nicely formatted URLs
\usepackage{color}
\usepackage{textcomp}
\usepackage{booktabs}
\usepackage{ccicons}
\usepackage{todonotes}

% llt: Define a global style for URLs, rather that the default one
\makeatletter
\def\url@leostyle{%
  \@ifundefined{selectfont}{\def\UrlFont{\sf}}{\def\UrlFont{\small\bf\ttfamily}}}
\makeatother
\urlstyle{leo}

% To make various LaTeX processors do the right thing with page size.
\def\pprw{8.5in}
\def\pprh{11in}
\special{papersize=\pprw,\pprh}
\setlength{\paperwidth}{\pprw}
\setlength{\paperheight}{\pprh}
\setlength{\pdfpagewidth}{\pprw}
\setlength{\pdfpageheight}{\pprh}

% Make sure hyperref comes last of your loaded packages, to give it a
% fighting chance of not being over-written, since its job is to
% redefine many LaTeX commands.
\definecolor{linkColor}{RGB}{6,125,233}
\hypersetup{%
  pdftitle={SIGCHI Conference Proceedings Format},
  pdfauthor={LaTeX},
  pdfkeywords={SIGCHI, proceedings, archival format},
  bookmarksnumbered,
  pdfstartview={FitH},
  colorlinks,
  citecolor=black,
  filecolor=black,
  linkcolor=black,
  urlcolor=linkColor,
  breaklinks=true,
}

% create a shortcut to typeset table headings
% \newcommand\tabhead[1]{\small\textbf{#1}}

% End of preamble. Here it comes the document.
\begin{document}

\title{SIGCHI Conference Proceedings Format}

\numberofauthors{3}
\author{%
  \alignauthor{1st Author Name\\
    \affaddr{Affiliation}\\
    \affaddr{City, Country}\\
    \email{e-mail address}}\\
  \alignauthor{2nd Author Name\\
    \affaddr{Affiliation}\\
    \affaddr{City, Country}\\
    \email{e-mail address}}\\
  \alignauthor{3rd Author Name\\
    \affaddr{Affiliation}\\
    \affaddr{City, Country}\\
    \email{e-mail address}}\\
}

\maketitle

\begin{abstract}
  UPDATED---\today. This sample paper describes the
  formatting requirements for SIGCHI conference proceedings, and
  offers recommendations on writing for the worldwide SIGCHI
  readership. Please review this document even if you have submitted
  to SIGCHI conferences before, as some format details have changed
  relative to previous years. Abstracts should be about 150 words and
  are required.
\end{abstract}

\keywords{Authors' choice; of terms; separated; by semi\-colons;
  commas, within terms only; this section is required.}

\category{H.5.m.}{Information Interfaces and Presentation
  (e.g. HCI)}{Miscellaneous} \category{See
  \url{http://acm.org/about/class/1998/} for the full list of ACM
  classifiers. This section is required.}{}{}

\section{Introduction}

This format is to be used for submissions that are published in the
conference proceedings. We wish to give this volume a consistent,
high-quality appearance. We therefore ask that authors follow some
simple guidelines. You should format your paper exactly like this
document. The easiest way to do this is to replace the content with
your own material.  This document describes how to prepare your
submissions using \LaTeX.

\section{Page Size and Columns}
On each page your material should fit within a rectangle of 7 $\times$
9.25 inches (18 $\times$ 23.5 cm), centered on a US Letter page (8.5
$\times$ 11 inches), beginning 0.75 inches (1.9 cm) from the top of
the page, with a 0.33 inches (0.85 cm) space between two 3.3 inches
(8.4 cm) columns. Right margins should be justified, not
ragged. Please be sure your document and PDF are US letter and not A4.

\section{Typeset Text}
The styles contained in this document have been modified from the
default styles to reflect ACM formatting conventions. For example,
content paragraphs like this one are formatted using the Normal style.

\LaTeX\ sometimes will create overfull lines that extend into columns.
To attempt to combat this, the \texttt{.cls} file has a command,
\texttt{{\textbackslash}sloppy}, that essentially asks \LaTeX\ to
prefer underfull lines with extra whitespace.  For more details on
this, and info on how to control it more finely, check out
{\url{http://www.economics.utoronto.ca/osborne/latex/PMAKEUP.HTM}}.

\subsection{Title and Authors}

Your paper's title, authors and affiliations should run across the
full width of the page in a single column 17.8 cm (7 in.) wide.  The
title should be in Helvetica or Arial 18-point bold.  Authors' names
should be in Times New Roman or Times Roman 12-point bold, and
affiliations in 12-point regular.

See \texttt{{\textbackslash}author} section of this template for
instructions on how to format the authors. For more than three
authors, you may have to place some address information in a footnote,
or in a named section at the end of your paper. Leave one 10-point
line of white space below the last line of affiliations.

\subsection{Abstract and Keywords}

Every submission should begin with an abstract of about 150 words,
followed by a set of Author Keywords and ACM Classification
Keywords. The abstract and keywords should be placed in the left
column of the first page under the left half of the title. The
abstract should be a concise statement of the problem, approach, and
conclusions of the work described. It should clearly state the paper's
contribution to the field of HCI\@.

\subsection{Normal or Body Text}

Please use a 10-point Times New Roman or Times Roman font or, if this
is unavailable, another proportional font with serifs, as close as
possible in appearance to Times Roman 10-point. Other than Helvetica
or Arial headings, please use sans-serif or non-proportional fonts
only for special purposes, such as source code text.

\subsection{First Page Copyright Notice}
This template include a sample ACM copyright notice at the bottom of
page 1, column 1.  Upon acceptance, you will be provided with the
appropriate copyright statement and unique DOI string for publication.
Accepted papers will be distributed in the conference
publications. They will also be placed in the ACM Digital Library,
where they will remain accessible to thousands of researchers and
practitioners worldwide. See
\url{http://acm.org/publications/policies/copyright_policy} for the
ACM’s copyright and permissions policy.

%%%

\subsection{Subsequent Pages}

On pages beyond the first, start at the top of the page and continue
in double-column format.  The two columns on the last page should be
of equal length.

\begin{figure}
\centering
  \includegraphics[width=0.9\columnwidth]{figures/sigchi-logo}
  \caption{Insert a caption below each figure. Do not alter the
    Caption style.}~\label{fig:figure1}
\end{figure}

\subsection{References and Citations}

Use a numbered list of references at the end of the article, ordered
alphabetically by last name of first author, and referenced by numbers
in
brackets~\cite{acm_categories,ethics,Klemmer:2002:WSC:503376.503378}.
Your references should be published materials accessible to the
public. Internal technical reports may be cited only if they are
easily accessible (i.e., you provide the address for obtaining the
report within your citation) and may be obtained by any reader for a
nominal fee. Proprietary information may not be cited. Private
communications should be acknowledged in the main text, not referenced
(e.g., ``[Borriello, personal communication]'').

References should be in ACM citation format:
\url{http://acm.org/publications/submissions/latex_style}. This
includes citations to internet
resources~\cite{acm_categories,cavender:writing,CHINOSAUR:venue,psy:gangnam}
according to ACM format, although it is often appropriate to include
URLs directly in the text, as above.

% Use a numbered list of references at the end of the article, ordered
% alphabetically by first author, and referenced by numbers in
% brackets~\cite{ethics, Klemmer:2002:WSC:503376.503378,
%   Mather:2000:MUT, Zellweger:2001:FAO:504216.504224}. For papers from
% conference proceedings, include the title of the paper and an
% abbreviated name of the conference (e.g., for Interact 2003
% proceedings, use \textit{Proc. Interact 2003}). Do not include the
% location of the conference or the exact date; do include the page
% numbers if available. See the examples of citations at the end of this
% document. Within this template file, use the \texttt{References} style
% for the text of your citation.

% Your references should be published materials accessible to the
% public.  Internal technical reports may be cited only if they are
% easily accessible (i.e., you provide the address for obtaining the
% report within your citation) and may be obtained by any reader for a
% nominal fee.  Proprietary information may not be cited. Private
% communications should be acknowledged in the main text, not referenced
% (e.g., ``[Robertson, personal communication]'').

\begin{table}
  \centering
  \begin{tabular}{r c c}
    \toprule
    & \multicolumn{2}{c}{\small{\textbf{Caption}}} \\
    \cmidrule(r){2-3}
    {\small\textbf{Objects}}
    & {\small \textit{Pre-2002}}
    & {\small \textit{Current}} \\
    \midrule
    Tables & Above & Below \\
    Figures & Below & Below \\
    \bottomrule
  \end{tabular}
  \caption{Table captions should be placed below the table. We
    recommend table lines be 1 point, 25\% black. Minimize use of
    unnecessary table lines.}~\label{tab:table1}
\end{table}

\section{Sections}

The heading of a section should be in Helvetica or Arial 9-point bold,
all in capitals. Sections should \textit{not} be numbered.

\subsection{Subsections}

Headings of subsections should be in Helvetica or Arial 9-point bold
with initial letters capitalized.  For sub-sections and
sub-subsections, a word like \emph{the} or \emph{of} is not
capitalized unless it is the first word of the heading.

\subsubsection{Sub-subsections}

Headings for sub-subsections should be in Helvetica or Arial 9-point
italic with initial letters capitalized.  Standard
\texttt{{\textbackslash}section}, \texttt{{\textbackslash}subsection},
and \texttt{{\textbackslash}subsubsection} commands will work fine in
this template.

\section{Figures/Captions}

Place figures and tables at the top or bottom of the appropriate
column or columns, on the same page as the relevant text (see
Figure~\ref{fig:figure1}). A figure or table may extend across both
columns to a maximum width of 17.78 cm (7 in.).

\begin{figure*}
  \centering
  \includegraphics[width=2\columnwidth]{figures/map}
  \caption{In this image, the map maximizes use of space. You can make
    figures as wide as you need, up to a maximum of the full width of
    both columns. Note that \LaTeX\ tends to render large figures on a
    dedicated page. Image: \ccbynd~ayman on
    Flickr.}~\label{fig:figure2}
\end{figure*}

Captions should be Times New Roman or Times Roman 9-point bold.  They
should be numbered (e.g., ``Table~\ref{tab:table1}'' or
``Figure~\ref{fig:figure1}''), centered and placed beneath the figure
or table.  Please note that the words ``Figure'' and ``Table'' should
be spelled out (e.g., ``Figure'' rather than ``Fig.'') wherever they
occur. Figures, like Figure~\ref{fig:figure2}, may span columns and
all figures should also include alt text for improved accessibility.
Papers and notes may use color figures, which are included in the page
limit; the figures must be usable when printed in black-and-white in
the proceedings.

The paper may be accompanied by a short video figure up to five
minutes in length. However, the paper should stand on its own without
the video figure, as the video may not be available to everyone who
reads the paper.

\subsection{Inserting Images}
When possible, include a vector formatted graphic (i.e. PDF or EPS).
When including bitmaps,  use an image editing tool to resize the image
at the appropriate printing resolution (usually 300 dpi).

\section{Language, Style and Content}

The written and spoken language of SIGCHI is English. Spelling and
punctuation may use any dialect of English (e.g., British, Canadian,
US, etc.) provided this is done consis- tently. Hyphenation is
optional. To ensure suitability for an international audience, please
pay attention to the following:

\begin{itemize}
\item Write in a straightforward style.
\item Try to avoid long or complex sentence structures.
\item Briefly define or explain all technical terms that may be
  unfamiliar to readers.
\item Explain all acronyms the first time they are used in your
  text---e.g., ``Digital Signal Processing (DSP)''.
\item Explain local references (e.g., not everyone knows all city
  names in a particular country).
\item Explain ``insider'' comments. Ensure that your whole audience
  understands any reference whose meaning you do not describe (e.g.,
  do not assume that everyone has used a Macintosh or a particular
  application).
\item Explain colloquial language and puns. Understanding phrases like
  ``red herring'' may require a local knowledge of English.  Humor and
  irony are difficult to translate.
\item Use unambiguous forms for culturally localized concepts, such as
  times, dates, currencies, and numbers (e.g., ``1--5--97'' or
  ``5/1/97'' may mean 5 January or 1 May, and ``seven o'clock'' may
  mean 7:00 am or 19:00). For currencies, indicate equivalences:
  ``Participants were paid {\fontfamily{txr}\selectfont \textwon}
  25,000, or roughly US \$22.''
\item Be careful with the use of gender-specific pronouns (he, she)
  and other gendered words (chairman, manpower, man-months). Use
  inclusive language that is gender-neutral (e.g., she or he, they,
  s/he, chair, staff, staff-hours, person-years). See the
  \textit{Guidelines for Bias-Free Writing} for further advice and
  examples regarding gender and other personal
  attributes~\cite{Schwartz:1995:GBF}. Be particularly aware of
  considerations around writing about people with disabilities.
\item If possible, use the full (extended) alphabetic character set
  for names of persons, institutions, and places (e.g.,
  Gr{\o}nb{\ae}k, Lafreni\'ere, S\'anchez, Nguy\~{\^{e}}n,
  Universit{\"a}t, Wei{\ss}enbach, Z{\"u}llighoven, \r{A}rhus, etc.).
  These characters are already included in most versions and variants
  of Times, Helvetica, and Arial fonts.
\end{itemize}

\section{Accessibility}
The Executive Council of SIGCHI has committed to making SIGCHI
conferences more inclusive for researchers, practitioners, and
educators with disabilities. As a part of this goal, the all authors
are asked to work on improving the accessibility of their
submissions. Specifically, we encourage authors to carry out the
following five steps:
\begin{enumerate}
\item Add alternative text to all figures
\item Mark table headings
\item Add tags to the PDF
\item Verify the default language
\item Set the tab order to ``Use Document Structure''
\end{enumerate}
For more information and links to instructions and resources, please
see: \url{http://chi2016.acm.org/accessibility}.  The
\texttt{{\textbackslash}hyperref} package allows you to create well tagged PDF files,
please see the preamble of this template for an example.

\section{Page Numbering, Headers and Footers}
Your final submission should not contain footer or header information
at the top or bottom of each page. Specifically, your final submission
should not include page numbers. Initial submissions may include page
numbers, but these must be removed for camera-ready. Page numbers will
be added to the PDF when the proceedings are assembled.

\section{Producing and Testing PDF Files}

We recommend that you produce a PDF version of your submission well
before the final deadline.  Your PDF file must be ACM DL
Compliant. The requirements for an ACM Compliant PDF are available at:
{\url{http://www.sheridanprinting.com/typedept/ACM-distilling-settings.htm}}.

Test your PDF file by viewing or printing it with the same software we
will use when we receive it, Adobe Acrobat Reader Version 10. This is
widely available at no cost. Note that most
reviewers will use a North American/European version of Acrobat
reader, so please check your PDF accordingly.

When creating your PDF from Word, ensure that you generate a tagged
PDF from improved accessibility. This can be done by using the Adobe
PDF add-in, also called PDFMaker. Select Acrobat | Preferences from
the ribbon and ensure that ``Enable Accessibility and Reflow with
tagged Adobe PDF'' is selected. You can then generate a tagged PDF by
selecting ``Create PDF'' from the Acrobat ribbon.

\section{Conclusion}

It is important that you write for the SIGCHI audience. Please read
previous years’ proceedings to understand the writing style and
conventions that successful authors have used. It is particularly
important that you state clearly what you have done, not merely what
you plan to do, and explain how your work is different from previously
published work, i.e., the unique contribution that your work makes to
the field. Please consider what the reader will learn from your
submission, and how they will find your work useful. If you write with
these questions in mind, your work is more likely to be successful,
both in being accepted into the conference, and in influencing the
work of our field.

\section{Acknowledgments}

Sample text: We thank all the volunteers, and all publications support
and staff, who wrote and provided helpful comments on previous
versions of this document. Authors 1, 2, and 3 gratefully acknowledge
the grant from NSF (\#1234--2012--ABC). \textit{This whole paragraph is
  just an example.}

% Balancing columns in a ref list is a bit of a pain because you
% either use a hack like flushend or balance, or manually insert
% a column break.  http://www.tex.ac.uk/cgi-bin/texfaq2html?label=balance
% multicols doesn't work because we're already in two-column mode,
% and flushend isn't awesome, so I choose balance.  See this
% for more info: http://cs.brown.edu/system/software/latex/doc/balance.pdf
%
% Note that in a perfect world balance wants to be in the first
% column of the last page.
%
% If balance doesn't work for you, you can remove that and
% hard-code a column break into the bbl file right before you
% submit:
%
% http://stackoverflow.com/questions/2149854/how-to-manually-equalize-columns-
% in-an-ieee-paper-if-using-bibtex
%
% Or, just remove \balance and give up on balancing the last page.
%
\balance{}

\section{References Format}
Your references should be published materials accessible to the
public. Internal technical reports may be cited only if they are
easily accessible (i.e., you provide the address for obtaining the
report within your citation) and may be obtained by any reader for a
nominal fee. Proprietary information may not be cited. Private
communications should be acknowledged in the main text, not referenced
(e.g., ``[Golovchinsky, personal communication]'').

Use a numbered list of references at the end of the article, ordered
alphabetically by first author, and referenced by numbers in
brackets~\cite{ethics,Klemmer:2002:WSC:503376.503378}. For papers from
conference proceedings, include the title of the paper and an
abbreviated name of the conference (e.g., for Interact 2003
proceedings, use Proc.\ Interact 2003). Do not include the location of
the conference or the exact date; do include the page numbers if
available. See the examples of citations at the end of this document
and in the accompanying \texttt{BibTeX} document.

References \textit{must be the same font size as other body
  text}. References should be in alphabetical order by last name of
first author. Example reference formatting for individual journal
articles~\cite{ethics}, articles in conference
proceedings~\cite{Klemmer:2002:WSC:503376.503378},
books~\cite{Schwartz:1995:GBF}, theses~\cite{sutherland:sketchpad},
book chapters~\cite{winner:politics}, a journal issue~\cite{kaye:puc},
websites~\cite{acm_categories,cavender:writing},
tweets~\cite{CHINOSAUR:venue}, patents~\cite{heilig:sensorama}, and
online videos~\cite{psy:gangnam} is given here. This formatting is a
slightly abbreviated version of the format automatically generated by
the ACM Digital Library (\url{http://dl.acm.org}) as ``ACM Ref''. More
details of reference formatting are available at:
\url{http://www.acm.org/publications/submissions/latex_style}.

% REFERENCES FORMAT
% References must be the same font size as other body text.
\bibliographystyle{SIGCHI-Reference-Format}
\bibliography{sample}

\end{document}

%%% Local Variables:
%%% mode: latex
%%% TeX-master: t
%%% End:

### 2015-08-24 flux-entropy-paper/flux-entropy-paper.tex
test name

@@
 
-\title{SIGCHI Conference Proceedings Format}
+\title{Quantitative Evaluation of Touch-Screen Performance with
+  Transition Matrix Measures}
 
@@
 \begin{abstract}
-  UPDATED---\today. This sample paper describes the
-  formatting requirements for SIGCHI conference proceedings, and
-  offers recommendations on writing for the worldwide SIGCHI
-  readership. Please review this document even if you have submitted
-  to SIGCHI conferences before, as some format details have changed
-  relative to previous years. Abstracts should be about 150 words and
-  are required.
+
 \end{abstract}
@@
 
-This format is to be used for submissions that are published in the
-conference proceedings. We wish to give this volume a consistent,
-high-quality appearance. We therefore ask that authors follow some
-simple guidelines. You should format your paper exactly like this
-document. The easiest way to do this is to replace the content with
-your own material.  This document describes how to prepare your
-submissions using \LaTeX.
-
-\section{Page Size and Columns}
-On each page your material should fit within a rectangle of 7 $\times$
-9.25 inches (18 $\times$ 23.5 cm), centered on a US Letter page (8.5
-$\times$ 11 inches), beginning 0.75 inches (1.9 cm) from the top of
-the page, with a 0.33 inches (0.85 cm) space between two 3.3 inches
-(8.4 cm) columns. Right margins should be justified, not
-ragged. Please be sure your document and PDF are US letter and not A4.
-
-\section{Typeset Text}
-The styles contained in this document have been modified from the
-default styles to reflect ACM formatting conventions. For example,
-content paragraphs like this one are formatted using the Normal style.
-
-\LaTeX\ sometimes will create overfull lines that extend into columns.
-To attempt to combat this, the \texttt{.cls} file has a command,
-\texttt{{\textbackslash}sloppy}, that essentially asks \LaTeX\ to
-prefer underfull lines with extra whitespace.  For more details on
-this, and info on how to control it more finely, check out
-{\url{http://www.economics.utoronto.ca/osborne/latex/PMAKEUP.HTM}}.
-
-\subsection{Title and Authors}
-
-Your paper's title, authors and affiliations should run across the
-full width of the page in a single column 17.8 cm (7 in.) wide.  The
-title should be in Helvetica or Arial 18-point bold.  Authors' names
-should be in Times New Roman or Times Roman 12-point bold, and
-affiliations in 12-point regular.
-
-See \texttt{{\textbackslash}author} section of this template for
-instructions on how to format the authors. For more than three
-authors, you may have to place some address information in a footnote,
-or in a named section at the end of your paper. Leave one 10-point
-line of white space below the last line of affiliations.
-
-\subsection{Abstract and Keywords}
-
-Every submission should begin with an abstract of about 150 words,
-followed by a set of Author Keywords and ACM Classification
-Keywords. The abstract and keywords should be placed in the left
-column of the first page under the left half of the title. The
-abstract should be a concise statement of the problem, approach, and
-conclusions of the work described. It should clearly state the paper's
-contribution to the field of HCI\@.
-
-\subsection{Normal or Body Text}
-
-Please use a 10-point Times New Roman or Times Roman font or, if this
-is unavailable, another proportional font with serifs, as close as
-possible in appearance to Times Roman 10-point. Other than Helvetica
-or Arial headings, please use sans-serif or non-proportional fonts
-only for special purposes, such as source code text.
-
-\subsection{First Page Copyright Notice}
-This template include a sample ACM copyright notice at the bottom of
-page 1, column 1.  Upon acceptance, you will be provided with the
-appropriate copyright statement and unique DOI string for publication.
-Accepted papers will be distributed in the conference
-publications. They will also be placed in the ACM Digital Library,
-where they will remain accessible to thousands of researchers and
-practitioners worldwide. See
-\url{http://acm.org/publications/policies/copyright_policy} for the
-ACM’s copyright and permissions policy.
-
-%%%
-
-\subsection{Subsequent Pages}
-
-On pages beyond the first, start at the top of the page and continue
-in double-column format.  The two columns on the last page should be
-of equal length.
-
-\begin{figure}
-\centering
-  \includegraphics[width=0.9\columnwidth]{figures/sigchi-logo}
-  \caption{Insert a caption below each figure. Do not alter the
-    Caption style.}~\label{fig:figure1}
-\end{figure}
-
-\subsection{References and Citations}
-
-Use a numbered list of references at the end of the article, ordered
-alphabetically by last name of first author, and referenced by numbers
-in
-brackets~\cite{acm_categories,ethics,Klemmer:2002:WSC:503376.503378}.
-Your references should be published materials accessible to the
-public. Internal technical reports may be cited only if they are
-easily accessible (i.e., you provide the address for obtaining the
-report within your citation) and may be obtained by any reader for a
-nominal fee. Proprietary information may not be cited. Private
-communications should be acknowledged in the main text, not referenced
-(e.g., ``[Borriello, personal communication]'').
-
-References should be in ACM citation format:
-\url{http://acm.org/publications/submissions/latex_style}. This
-includes citations to internet
-resources~\cite{acm_categories,cavender:writing,CHINOSAUR:venue,psy:gangnam}
-according to ACM format, although it is often appropriate to include
-URLs directly in the text, as above.
-
-% Use a numbered list of references at the end of the article, ordered
-% alphabetically by first author, and referenced by numbers in
-% brackets~\cite{ethics, Klemmer:2002:WSC:503376.503378,
-%   Mather:2000:MUT, Zellweger:2001:FAO:504216.504224}. For papers from
-% conference proceedings, include the title of the paper and an
-% abbreviated name of the conference (e.g., for Interact 2003
-% proceedings, use \textit{Proc. Interact 2003}). Do not include the
-% location of the conference or the exact date; do include the page
-% numbers if available. See the examples of citations at the end of this
-% document. Within this template file, use the \texttt{References} style
-% for the text of your citation.
-
-% Your references should be published materials accessible to the
-% public.  Internal technical reports may be cited only if they are
-% easily accessible (i.e., you provide the address for obtaining the
-% report within your citation) and may be obtained by any reader for a
-% nominal fee.  Proprietary information may not be cited. Private
-% communications should be acknowledged in the main text, not referenced
-% (e.g., ``[Robertson, personal communication]'').
-
-\begin{table}
-  \centering
-  \begin{tabular}{r c c}
-    \toprule
-    & \multicolumn{2}{c}{\small{\textbf{Caption}}} \\
-    \cmidrule(r){2-3}
-    {\small\textbf{Objects}}
-    & {\small \textit{Pre-2002}}
-    & {\small \textit{Current}} \\
-    \midrule
-    Tables & Above & Below \\
-    Figures & Below & Below \\
-    \bottomrule
-  \end{tabular}
-  \caption{Table captions should be placed below the table. We
-    recommend table lines be 1 point, 25\% black. Minimize use of
-    unnecessary table lines.}~\label{tab:table1}
-\end{table}
-
-\section{Sections}
-
-The heading of a section should be in Helvetica or Arial 9-point bold,
-all in capitals. Sections should \textit{not} be numbered.
-
-\subsection{Subsections}
-
-Headings of subsections should be in Helvetica or Arial 9-point bold
-with initial letters capitalized.  For sub-sections and
-sub-subsections, a word like \emph{the} or \emph{of} is not
-capitalized unless it is the first word of the heading.
-
-\subsubsection{Sub-subsections}
-
-Headings for sub-subsections should be in Helvetica or Arial 9-point
-italic with initial letters capitalized.  Standard
-\texttt{{\textbackslash}section}, \texttt{{\textbackslash}subsection},
-and \texttt{{\textbackslash}subsubsection} commands will work fine in
-this template.
-
-\section{Figures/Captions}
-
-Place figures and tables at the top or bottom of the appropriate
-column or columns, on the same page as the relevant text (see
-Figure~\ref{fig:figure1}). A figure or table may extend across both
-columns to a maximum width of 17.78 cm (7 in.).
-
-\begin{figure*}
-  \centering
-  \includegraphics[width=2\columnwidth]{figures/map}
-  \caption{In this image, the map maximizes use of space. You can make
-    figures as wide as you need, up to a maximum of the full width of
-    both columns. Note that \LaTeX\ tends to render large figures on a
-    dedicated page. Image: \ccbynd~ayman on
-    Flickr.}~\label{fig:figure2}
-\end{figure*}
-
-Captions should be Times New Roman or Times Roman 9-point bold.  They
-should be numbered (e.g., ``Table~\ref{tab:table1}'' or
-``Figure~\ref{fig:figure1}''), centered and placed beneath the figure
-or table.  Please note that the words ``Figure'' and ``Table'' should
-be spelled out (e.g., ``Figure'' rather than ``Fig.'') wherever they
-occur. Figures, like Figure~\ref{fig:figure2}, may span columns and
-all figures should also include alt text for improved accessibility.
-Papers and notes may use color figures, which are included in the page
-limit; the figures must be usable when printed in black-and-white in
-the proceedings.
-
-The paper may be accompanied by a short video figure up to five
-minutes in length. However, the paper should stand on its own without
-the video figure, as the video may not be available to everyone who
-reads the paper.
-
-\subsection{Inserting Images}
-When possible, include a vector formatted graphic (i.e. PDF or EPS).
-When including bitmaps,  use an image editing tool to resize the image
-at the appropriate printing resolution (usually 300 dpi).
-
-\section{Language, Style and Content}
-
-The written and spoken language of SIGCHI is English. Spelling and
-punctuation may use any dialect of English (e.g., British, Canadian,
-US, etc.) provided this is done consis- tently. Hyphenation is
-optional. To ensure suitability for an international audience, please
-pay attention to the following:
-
-\begin{itemize}
-\item Write in a straightforward style.
-\item Try to avoid long or complex sentence structures.
-\item Briefly define or explain all technical terms that may be
-  unfamiliar to readers.
-\item Explain all acronyms the first time they are used in your
-  text---e.g., ``Digital Signal Processing (DSP)''.
-\item Explain local references (e.g., not everyone knows all city
-  names in a particular country).
-\item Explain ``insider'' comments. Ensure that your whole audience
-  understands any reference whose meaning you do not describe (e.g.,
-  do not assume that everyone has used a Macintosh or a particular
-  application).
-\item Explain colloquial language and puns. Understanding phrases like
-  ``red herring'' may require a local knowledge of English.  Humor and
-  irony are difficult to translate.
-\item Use unambiguous forms for culturally localized concepts, such as
-  times, dates, currencies, and numbers (e.g., ``1--5--97'' or
-  ``5/1/97'' may mean 5 January or 1 May, and ``seven o'clock'' may
-  mean 7:00 am or 19:00). For currencies, indicate equivalences:
-  ``Participants were paid {\fontfamily{txr}\selectfont \textwon}
-  25,000, or roughly US \$22.''
-\item Be careful with the use of gender-specific pronouns (he, she)
-  and other gendered words (chairman, manpower, man-months). Use
-  inclusive language that is gender-neutral (e.g., she or he, they,
-  s/he, chair, staff, staff-hours, person-years). See the
-  \textit{Guidelines for Bias-Free Writing} for further advice and
-  examples regarding gender and other personal
-  attributes~\cite{Schwartz:1995:GBF}. Be particularly aware of
-  considerations around writing about people with disabilities.
-\item If possible, use the full (extended) alphabetic character set
-  for names of persons, institutions, and places (e.g.,
-  Gr{\o}nb{\ae}k, Lafreni\'ere, S\'anchez, Nguy\~{\^{e}}n,
-  Universit{\"a}t, Wei{\ss}enbach, Z{\"u}llighoven, \r{A}rhus, etc.).
-  These characters are already included in most versions and variants
-  of Times, Helvetica, and Arial fonts.
-\end{itemize}
-
-\section{Accessibility}
-The Executive Council of SIGCHI has committed to making SIGCHI
-conferences more inclusive for researchers, practitioners, and
-educators with disabilities. As a part of this goal, the all authors
-are asked to work on improving the accessibility of their
-submissions. Specifically, we encourage authors to carry out the
-following five steps:
-\begin{enumerate}
-\item Add alternative text to all figures
-\item Mark table headings
-\item Add tags to the PDF
-\item Verify the default language
-\item Set the tab order to ``Use Document Structure''
-\end{enumerate}
-For more information and links to instructions and resources, please
-see: \url{http://chi2016.acm.org/accessibility}.  The
-\texttt{{\textbackslash}hyperref} package allows you to create well tagged PDF files,
-please see the preamble of this template for an example.
-
-\section{Page Numbering, Headers and Footers}
-Your final submission should not contain footer or header information
-at the top or bottom of each page. Specifically, your final submission
-should not include page numbers. Initial submissions may include page
-numbers, but these must be removed for camera-ready. Page numbers will
-be added to the PDF when the proceedings are assembled.
-
-\section{Producing and Testing PDF Files}
-
-We recommend that you produce a PDF version of your submission well
-before the final deadline.  Your PDF file must be ACM DL
-Compliant. The requirements for an ACM Compliant PDF are available at:
-{\url{http://www.sheridanprinting.com/typedept/ACM-distilling-settings.htm}}.
-
-Test your PDF file by viewing or printing it with the same software we
-will use when we receive it, Adobe Acrobat Reader Version 10. This is
-widely available at no cost. Note that most
-reviewers will use a North American/European version of Acrobat
-reader, so please check your PDF accordingly.
-
-When creating your PDF from Word, ensure that you generate a tagged
-PDF from improved accessibility. This can be done by using the Adobe
-PDF add-in, also called PDFMaker. Select Acrobat | Preferences from
-the ribbon and ensure that ``Enable Accessibility and Reflow with
-tagged Adobe PDF'' is selected. You can then generate a tagged PDF by
-selecting ``Create PDF'' from the Acrobat ribbon.
-
-\section{Conclusion}
-
-It is important that you write for the SIGCHI audience. Please read
-previous years’ proceedings to understand the writing style and
-conventions that successful authors have used. It is particularly
-important that you state clearly what you have done, not merely what
-you plan to do, and explain how your work is different from previously
-published work, i.e., the unique contribution that your work makes to
-the field. Please consider what the reader will learn from your
-submission, and how they will find your work useful. If you write with
-these questions in mind, your work is more likely to be successful,
-both in being accepted into the conference, and in influencing the
-work of our field.
-
-\section{Acknowledgments}
-
-Sample text: We thank all the volunteers, and all publications support
-and staff, who wrote and provided helpful comments on previous
-versions of this document. Authors 1, 2, and 3 gratefully acknowledge
-the grant from NSF (\#1234--2012--ABC). \textit{This whole paragraph is
-  just an example.}
-
-% Balancing columns in a ref list is a bit of a pain because you
-% either use a hack like flushend or balance, or manually insert
-% a column break.  http://www.tex.ac.uk/cgi-bin/texfaq2html?label=balance
-% multicols doesn't work because we're already in two-column mode,
-% and flushend isn't awesome, so I choose balance.  See this
-% for more info: http://cs.brown.edu/system/software/latex/doc/balance.pdf
-%
-% Note that in a perfect world balance wants to be in the first
-% column of the last page.
-%
-% If balance doesn't work for you, you can remove that and
-% hard-code a column break into the bbl file right before you
-% submit:
-%
-% http://stackoverflow.com/questions/2149854/how-to-manually-equalize-columns-
-% in-an-ieee-paper-if-using-bibtex
-%
-% Or, just remove \balance and give up on balancing the last page.
-%
-\balance{}
-
-\section{References Format}
-Your references should be published materials accessible to the
-public. Internal technical reports may be cited only if they are
-easily accessible (i.e., you provide the address for obtaining the
-report within your citation) and may be obtained by any reader for a
-nominal fee. Proprietary information may not be cited. Private
-communications should be acknowledged in the main text, not referenced
-(e.g., ``[Golovchinsky, personal communication]'').
-
-Use a numbered list of references at the end of the article, ordered
-alphabetically by first author, and referenced by numbers in
-brackets~\cite{ethics,Klemmer:2002:WSC:503376.503378}. For papers from
-conference proceedings, include the title of the paper and an
-abbreviated name of the conference (e.g., for Interact 2003
-proceedings, use Proc.\ Interact 2003). Do not include the location of
-the conference or the exact date; do include the page numbers if
-available. See the examples of citations at the end of this document
-and in the accompanying \texttt{BibTeX} document.
-
-References \textit{must be the same font size as other body
-  text}. References should be in alphabetical order by last name of
-first author. Example reference formatting for individual journal
-articles~\cite{ethics}, articles in conference
-proceedings~\cite{Klemmer:2002:WSC:503376.503378},
-books~\cite{Schwartz:1995:GBF}, theses~\cite{sutherland:sketchpad},
-book chapters~\cite{winner:politics}, a journal issue~\cite{kaye:puc},
-websites~\cite{acm_categories,cavender:writing},
-tweets~\cite{CHINOSAUR:venue}, patents~\cite{heilig:sensorama}, and
-online videos~\cite{psy:gangnam} is given here. This formatting is a
-slightly abbreviated version of the format automatically generated by
-the ACM Digital Library (\url{http://dl.acm.org}) as ``ACM Ref''. More
-details of reference formatting are available at:
-\url{http://www.acm.org/publications/submissions/latex_style}.
-
-% REFERENCES FORMAT
-% References must be the same font size as other body text.
 \bibliographystyle{SIGCHI-Reference-Format}
-\bibliography{sample}
+\bibliography{references}
 

### 2015-08-25 flux-entropy-paper/flux-entropy-paper.tex
added maths to motivate transition matrices, flux and entropy

@@
 \begin{abstract}
-
+  HCI systems for supporting collaborative creativity are often
+  evaluated with user ratings and interviews. However, such data can
+  be impractical to collect in-the-wild, such as during musical
+  performances. In this paper we consider measures of gestural
+  transitions during collaborative interactions with musical touch
+  screen interfaces. The performers' interactions are characterised as
+  sequences of discrete gestures and summarised in a transition matrix
+  for each performance. Using a corpus of more than 140 performances,
+  rehearsals, and demonstrations, we find that two matrix measures,
+  flux and entropy, vary significantly with different kinds of
+  in-the-wild interactions. Further, using performers' ratings
+  collected during a lab-study we find that these measures have a
+  significant relationship with the perceived effect of the interface
+  on performance. Thus transition analysis, flux and entropy, can be
+  used as a proxy for performer ratings during in-the-wild interaction.
 \end{abstract}
 
-\keywords{Authors' choice; of terms; separated; by semi\-colons;
-  commas, within terms only; this section is required.}
+\keywords{  Creativity Support Tools;
+  Agent; 
+  Design;
+  Methodology}
 
-\category{H.5.m.}{Information Interfaces and Presentation
-  (e.g. HCI)}{Miscellaneous} \category{See
-  \url{http://acm.org/about/class/1998/} for the full list of ACM
-  classifiers. This section is required.}{}{}
+\category{H.5.5.}{Information Interfaces and Presentation
+  (e.g. HCI)}{Sound and Music Computing}{Systems}
+
+% \category{See
+%   \url{http://acm.org/about/class/1998/} for the full list of ACM
+%   classifiers. This section is required.}{}{}
 
 \section{Introduction}
+
+Finally, we report a quantitative analysis of the transition matrices
+from our experimental protocols, and describe the quantities
+\emph{flux} and \emph{entropy} which can be used to characterise the
+gestural activity of the musicians during a performance. We found a
+significant effect of the app on these measures, allowing us to
+observe quantitatively the performers' dissatisfaction with a
+particular interface. We also investigate the significant relationship
+between both of these measures and the survey responses, which
+suggests that performers rate interaction with our agent and apps more
+highly when engaged in adventurous, exploratory performance.
+
+\section{Related Work}
+
+\section{Transition Matrices and Matrix Measures}
+
+It has been recognised that collaborative interactions such as
+improvised musical performances can be modelled as a sequence of
+states. In the following section we present a method for analysing the
+transitions between adjacent states in such a sequence and derive two
+high-level quantities that we use to compare musical performances
+under different circumstances. Suppose that an collaborative
+interaction can be characterised by a set of interaction states $S$
+with $m$ members.
+Each participant's activity can then be represented as a sequence:
+\begin{equation}
+ X_n \hskip 2em n = 1, \ldots, N
+\end{equation}
+Where each $X_i$ is a member of $G$. The transitions between different
+interaction states can be summarised in a transition matrix as if the
+sequence $X_n$ were a first-order Markov chain. Let $N_{ij}$ be the
+number of times that state $j$ follows state $i$. Each participant's
+transition matrix, $P$, will then be an $m \times m$ matrix with
+entries calculated thus
+\begin{equation}
+  p_{ij} = \frac{N_{ij}}{\sum_j N_{ij}}
+\end{equation}
+Each entry of $P$ will be such that the transition probability from
+state $i$ to state $j$ is given by the entry in the $i^{th}$ row and
+the $j^{th}$ column. The transition activity of the whole ensemble can
+be calculated by averaging the transition matrix for each performer.
+
+To compare the transition matrices of different interactions we use
+two high-level quantities, ``flux'' and ``entropy''. The flux of a
+transition matrix $P$ is given by
+\begin{equation}
+  \mathrm{flux}(P) = \frac{\|P\|_1-\|\mathrm{diag}(P)\|}{\|P\|_1}
+\end{equation}
+Where $\|P\|_1 = \sum_{i,j}|p_{ij}|$ is the element-wise 1-norm of $P$
+and $\mathrm{diag}(P)$ is the vector of the main diagonal entries of
+$P$.
+
+Flux measures the ratio of transitions that change
+state to self-transitions from a certain state back to itself.
+Intuitively, it is a measure of how frequently a group changes state. Flux
+returns a value in the interval $[0,1]$ and will return 0 when
+participants never change state, and 1 when participants never stay on
+the same state for two measurements in a row.
+
+A related (but different) high-level derived quantity for the
+transition matrix is its entropy, defined in the
+information-theoretic~\cite{Shannon:1948rt} sense:
+\begin{equation}
+  H(P) = -\sum_{i,j}p_{ij}\log_2(p_{ij})
+\end{equation}
+This entropy measure is small when the matrix is sparse, and largest
+when each matrix element is equal. It offers a different perspective
+on collaborative interactions than the flux measure by capturing the
+breadth of the gestural space explored by the participants throughout
+the course of an interaction. Consider the degenerate case of a
+participant who alternates between two states for a whole performance:
+the flux in this case will be maximal ($\mathrm{flux}(P) = 1$) since
+there are no self-transitions, (only $A \rightarrow B$ or
+$ B \rightarrow A$) even though the participant has only used a small
+subset of the state space. The entropy of this interaction,
+on the other hand, will be low. Entropy, therefore, is a measure of
+how broad the participant's exploration of the state space is in a given
+interaction.
+
+\section{Modelling Performer Ratings with Flux and Entropy}
+
+\begin{figure}
+  \centering
+  \includegraphics[width=\linewidth]{figures/stat-response-fit.pdf}
+  \caption{Transition matrix statistics (flux above and entropy below) vs Likert responses for
+    questions 3, 6 and 7, all of which exhibited significant
+    relationships under a proportional odds logistic regression model.
+    A least-squares linear model is shown with a dotted line.
+    \label{fig:stat-response-fit}}
+\end{figure}
+
+One final question to explore is how these flux and entropy measures
+in the interaction data relate to the performers' perceptions of their
+performance?
+
+To answer this question, we construct models with the flux and entropy
+measures as independent variables and the Likert responses (1 to 5) as
+the dependent variable. These measures exhibited clear relationships
+for three of the survey questions (Questions 3, 6 and 7), which are
+plotted in Figure~\ref{fig:stat-response-fit} where the least-squares
+line of best fit has a positive gradient for both flux and entropy.
+
+Determining the strengths of these positive relationships is
+difficult, both due to the problem of treating ordinal Likert
+responses as numeric~\cite{Gardner:2007fj} and also
+due to the unknown distribution of the entropy and flux statistics. As
+a result, we use a conservative proportional odds logistic regression
+(POLR)~(using the \texttt{polr} function in \texttt{R}, from the MASS
+package \cite{Venables:2002qv}) to treat the Likert responses as an
+ordered factor response.
+
+The POLR results indicate a significant ($p<0.05$) relationship
+between the user-reported Likert response and the flux statistic for
+Question 3 ($t = 2.77, df = 4, p = 0.025$), Question 6
+($t = 2.41, df = 4, p = 0.037$), and Question 7
+($t = 2.49, df = 4, p = 0.034$). The same questions displayed a
+significant relationship for the entropy statistic: Question 3
+($t = 2.93, df = 4, p = 0.022$), Question 6
+($t = 2.69, df = 4, p = 0.028$), and Question 7
+($t = 2.89, df = 4, p = 0.022$). These relationships are shown in
+Figure~\ref{fig:stat-response-fit}.
+
+It is notable that these questions (3, 6, and 7) all pertain to the
+\emph{perceived effect of the agent} on the performers. As the agent
+uses the flux measure to trigger new-idea messages, it appears that
+performers appreciated its intervention more in particularly creative
+performances where flux and entropy were high.
 

### 2015-08-25 metatone_post_hoc_analysis.py
tried to reduce ordinal logistic regression to Q6 from 2014 study

@@ -166,6 +166,9 @@ def main():
     flux_ratings_frame = perf_frame['flux']
     flux_ratings_frame = pd.concat([flux_ratings_frame,ratings],axis =1)
 
+    ratings = performance_surveys_2014["Q6"]
+    flux_entropy_ratings = flux_entropy_ratings.dropna()
+
     flux_entropy_ratings = pd.DataFrame({"rating":ratings})
     flux_entropy_ratings["flux"] = perf_frame["flux"]
     flux_entropy_ratings["entropy"] = perf_frame["entropy"]

### 2015-08-25 R/analysis.R
add time-chunking R code

@@ -62,3 +62,29 @@ message(paste("($t = ", format(summary(mod)$coefficients[1,3], digits = 3), ", d
 
 ggplot(flux.entropy.ratings, aes(as.factor(rating),flux)) + geom_point() + geom_boxplot()
 ggplot(flux.entropy.ratings, aes(flux,rating)) + geom_smooth(method=lm)
+
+## time-window chunking (beginning-middle-end)
+
+library("plyr")
+library("reshape2")
+
+filenames <- llply(list.files("../data"),
+                  function(x) c(x, names(read.csv(paste("../data/", x, sep = "")))))
+
+process_session <- function(filename, session_df){
+    ## assume first column is filename, second is time
+    t <- as.numeric(strptime(as.character(session_df$time), "%Y-%m-%d %H:%M:%S"))
+    ## normalise the time variable into [0,1]
+    session_df$time <- (t-min(t))/(max(t)-min(t))
+    melt(cbind(filename, session_df), id.vars = 1:2, variable.name = "musician", value.name = "gesture")
+}
+
+df <- ldply(list.files("../data"),
+            function(x) {
+                filename <- paste("../data/", x, sep = "")
+                process_session(x, read.csv(filename))
+            })
+## get all the column classes right
+df$gesture <- factor(df$gesture)
+## add time-windowed chunk section
+df$section <- ordered(c("beginning", "middle", "end"), levels = c("beginning", "middle", "end"))[as.integer(df$time*3)+1]

### 2015-08-25 metatone_post_hoc_analysis.py
updates to split up some monolithic files

@@ -17,6 +17,9 @@ import metatone_classifier
 import transitions
 import PlotMetatonePerformanceAndTransitions
 import generate_posthoc_gesture_score
+from minirank import *
+from sklearn import datasets, metrics, cross_validation
+import matplotlib.pyplot as plt
 
 class MetatoneTouchLog:
     """
@@ -136,8 +139,8 @@ def main():
     perf_frame.to_csv("metatone-performance-data.csv")
 
     #print("Creating Gesture Scores.")
-    #for perf in performances:
-    #    perf.print_gesture_score() ## Prints out a gesture-score pdf for reference.
+    for perf in performances:
+        perf.print_gesture_score() ## Prints out a gesture-score pdf for reference.
 
     # print("Finding the lengths.")
     # performer_length_dict = {}
@@ -160,27 +163,25 @@ def main():
 
     # a.apply(lambda x: LIKERT_MAPPING[x])
     LIKERT_MAPPING = {1:1,2:1,3:2,4:2,5:3,6:4,7:4,8:5,9:5}
-    ratings = performance_surveys_2014["Q1"]
-    ratings = c.append(performance_surveys_2015_runthrough["Q26"].apply(lambda x: LIKERT_MAPPING[x]))
-    ratings = ratings.append(performance_surveys_2015_study["Q23"].apply(lambda x: LIKERT_MAPPING[x]))
-    flux_ratings_frame = perf_frame['flux']
-    flux_ratings_frame = pd.concat([flux_ratings_frame,ratings],axis =1)
+    #ratings = performance_surveys_2014["Q1"]
+    #ratings = ratings.append(performance_surveys_2015_runthrough["Q26"].apply(lambda x: LIKERT_MAPPING[x]))
+    #ratings = ratings.append(performance_surveys_2015_study["Q23"].apply(lambda x: LIKERT_MAPPING[x]))
+    #flux_ratings_frame = perf_frame['flux']
+    #flux_ratings_frame = pd.concat([flux_ratings_frame,ratings],axis =1)
 
-    ratings = performance_surveys_2014["Q6"]
-    flux_entropy_ratings = flux_entropy_ratings.dropna()
+    #ratings = performance_surveys_2014["Q6"]
+    #flux_entropy_ratings = flux_entropy_ratings.dropna()
 
-    flux_entropy_ratings = pd.DataFrame({"rating":ratings})
-    flux_entropy_ratings["flux"] = perf_frame["flux"]
[…]

### 2015-08-25 flux-entropy-paper/flux-entropy-paper.tex
a few extra words

@@
 
+\section{A Corpus of Musical Touch-Screen Interactions}
+
+In the following sections, we will use transition matrices, flux, and
+entropy to analyse a corpus of data collected as part of our research
+into collaborative interactions with musical touch-screen tablet
+interfaces. In a period of more than two years (April 2013 - June 2015), we have
+collected data from $135$ sessions where various participants used
+musical touch-screen interfaces on Apple iPad devices. These sessions
+had a number of contexts including rehearsals, performances, lab study
+sessions, classes, and casual demonstrations. Participants performed
+free improvisations as well as written compositions, performed on
+iPads alone and in setups of iPads and acoustic percussion
+instruments. The median length of these sessions was 7:32 and the
+median number of participants was four.
+
 \section{Modelling Performer Ratings with Flux and Entropy}

### 2015-08-25 R/analysis.R
trying to make a better plot for entropy

@@ -1,7 +1,7 @@
 library(ggplot2)
-
 library(MASS)
-
+chifig.3colours <- c("#e41a1c", "#377eb8", "#4daf4a")
+chifig.2colours <- c("#984ea3", "#ff7f00")
 ## time,filename,number_performers,performance_context,performance_type,instruments,notes,video_location,flux,number_performers,length_seconds,entropy,filename
 df <- read.csv("../metatone-performance-data.csv")
 df$time <- strptime(as.character(df$time), "%Y-%m-%d %H:%M:%S.%OS")
@@ -16,22 +16,48 @@ summary(reh.perf.studies.df)
 
 ggplot(reh.perf.studies.df, aes(performance_type)) + geom_bar(aes(fill = performance_context))
 ggplot(reh.perf.studies.df, aes(performance_type, flux)) + geom_boxplot(aes(fill = performance_context))
-ggplot(reh.perf.studies.df, aes(performance_type, entropy)) + geom_boxplot(aes(fill = performance_context))
+ggplot(reh.perf.studies.df, aes(performance_type, entropy)) + geom_point(aes(colour = performance_context))
+
+    geom_boxplot(aes(fill = performance_context))
 
 
 ggplot(df, aes(flux, entropy)) + geom_point(aes(colour = instruments)) + facet_wrap(~instruments)
 ggplot(df, aes(flux, entropy)) + geom_point(aes(size = number_performers, colour = performance_type))
+
 ggplot(df, aes(flux, number_performers)) + geom_point(aes(colour = performance_type))
-ggplot(df, aes(flux, exp(entropy))) + geom_point(aes(size = number_performers, colour = performance_context))
+
+ggplot(df, aes(flux, exp(entropy))) + geom_point(aes(size = number_performers, colour = performance_context), alpha = 0.6)
+
+
 
 + facet_wrap(~performance_context)
+
+
 ggplot(df, aes(time, flux)) + geom_point(aes(size = number_performers, colour = performance_type))
 
 ## stats
+head(df)
 
 ## Question: Do Performance Type and Context have a significant effect on Flux and Entropy?
 summary(aov(flux~performance_context*performance_type*instruments, reh.perf.studies.df))
-summary(aov(entropy~performance_context*performance_type*instruments, reh.perf.studies.df))
+summary(aov(flux~performance_context*performance_type*instruments, df))
+
+
+summary(aov(exp(entropy) ~ performance_context * performance_type * instruments, df))
[…]

### 2015-08-25 R/analysis.R
add some more time-chunking code

@@ -94,6 +94,8 @@ ggplot(flux.entropy.ratings, aes(flux,rating)) + geom_smooth(method=lm)
 library("plyr")
 library("reshape2")
 
+source("tm.R")
+
 filenames <- llply(list.files("../data"),
                   function(x) c(x, names(read.csv(paste("../data/", x, sep = "")))))
 
@@ -111,6 +113,22 @@ df <- ldply(list.files("../data"),
                 process_session(x, read.csv(filename))
             })
 ## get all the column classes right
-df$gesture <- factor(df$gesture)
+df$gesture <- ordered(gestures[df$gesture+1], levels = gestures)
 ## add time-windowed chunk section
 df$section <- ordered(c("beginning", "middle", "end"), levels = c("beginning", "middle", "end"))[as.integer(df$time*3)+1]
+
+## Transition Matrices
+tm <- ddply(df, .(filename, musician, section), calculate.transitions)
+## add flux, entropy
+tm <- ddply(tm, .(filename, musician, section),
+            function(x){
+                data.frame(x,
+                           flux = transition.flux(x),
+                           entropy = transition.entropy(x))
+            })
+
+## plotting
+library("ggplot2")
+
+ggplot(tm, aes(section, flux)) + geom_violin(aes(fill=section))
+ggplot(tm, aes(section, entropy)) + geom_violin(aes(fill=section))

### 2015-08-25 R/tm.R

@@ -1,0 +1,142 @@
+gestures <- c("N", # 0: Nothing
+              "FT", # 1: Fast Taps
+              "ST", # 2: Slow Taps
+              "FS", # 3: Fast Swipes
+              "FSA", # 4: Fast Swipes Accelerating
+              "VSS", # 5: Very Slow Swirl
+              "BS", # 6: Big Swirl
+              "SS", # 7: Small Swirl
+              "C" # 8: Combination of Swirl and Taps
+              )
+
+jumble.df <- function(df, cols.to.jumble){
+    if(is.numeric(cols.to.jumble) && (TRUE %in% (cols.to.jumble < 0)))
+        cols.to.jumble <- (1:dim(df)[2])[cols.to.jumble]
+    for(c in cols.to.jumble){
+        df[,c] <- df[sample.int(dim(df)[1]),c]
+    }
+    df
+}
+
+## transition matrices
+
+## assumes single session-artist df
+calculate.transitions <- function(df, left.stochastic=FALSE){
+    if(dim(df)[1]<2)
+        return(data.frame())
+    trans <- data.frame(from = df$gesture[-length(df$gesture)],
+                        to= df$gesture[-1])
+    if(left.stochastic)
+        res <- ddply(trans, .(from), function(x) ddply(x, .(to), function(y) data.frame(prob = length(y$to)/length(x$from))))
+    else
+        res <- ddply(trans, .(from), function(x) ddply(x, .(to), function(y) data.frame(prob = length(y$to)/length(trans$from))))
+    data.frame(time = min(df$time), res)
+}
+
+## ## add together (and renormalise) TMs from different musicians
+## group.transitions <- function(df){
+##     ddply(df,
+##           .(session, app, agent, time, from, to),
+##           summarize,
+##           blah...)
+## }
+
+transition.flux <- function(df){
+    if(sum(df$prob)!=1)
+        message("warning: prob sum isn't 1")
+    on.diag.sum <- sum(df[df$from==df$to, "prob"])
+    data.frame(flux = 1-on.diag.sum)
+}
+
+transition.entropy <- function(df){
+    data.frame(entropy = -sum(df$prob*log2(df$prob)))
+}
+
+quantize.times <- function(time, delta.t){
+    delta.t * (time %/% delta.t)
+}
+
+timeslice.df <- function(df, delta.t){
+    mutate(df, time = quantize.times(time, delta.t))
+}
+
+timesliced.transitions <- function(df, delta.t){
[…]

### 2015-08-25 break_up_monolithic_record.py
split up a huge record with a new python script

@@ -1,0 +1,24 @@
+"""
+Split up a touch frame into chunks separated by 30s of inactivity.
+"""
+touches = pd.read_csv("data/2015-03-20T13-39-35-MetatoneOSCLog-6PlayerRehearsal-touches.csv", index_col='time', parse_dates=True)
+touches["timediff"] = touches.index
+touches["timediff"] = touches["timediff"] - touches["timediff"].shift()
+long_break = touches.ix[touches["timediff"] > pd.to_timedelta("00:00:30")]
+#touches.ix[touches.index.indexer_between_time(long_break.index[1],long_break.index[2])]
+first_touch = touches[:1].index[0]
+frame_list = []
+for last_touch in long_break.index:
+    part_frame = touches.ix[touches.index.indexer_between_time(first_touch,last_touch)]
+    first_touch = last_touch
+    frame_list.append(part_frame[:-1]) ## this makes it left closed, right open
+    
+# then add the rest:
+last_touch = touches.index[-1]
+part_frame =  touches.ix[touches.index.indexer_between_time(first_touch,last_touch)] #inclusive
+frame_list.append(part_frame) ## the last one is closed on both sides!
+
+# now write each one to csv. (forget about the time_diff column)
+for frame in frame_list:
+    frame_name = frame.index[0].strftime('%Y-%m-%dT%H-%M-%S') + "-MetatoneOSCLog-autosplit-touches.csv"
+    frame[['device_id','x_pos','y_pos','velocity']].to_csv(frame_name)

### 2015-08-26 metatone_post_hoc_analysis.py
stop making gesture plots

@@ -139,8 +139,8 @@ def main():
     perf_frame.to_csv("metatone-performance-data.csv")
 
     #print("Creating Gesture Scores.")
-    for perf in performances:
-        perf.print_gesture_score() ## Prints out a gesture-score pdf for reference.
+    #for perf in performances:
+    #    perf.print_gesture_score() ## Prints out a gesture-score pdf for reference.
 
     # print("Finding the lengths.")
     # performer_length_dict = {}

### 2015-08-26 break_up_monolithic_record.py
more data splits

@@ -1,10 +1,16 @@
 """
 Split up a touch frame into chunks separated by 30s of inactivity.
 """
-touches = pd.read_csv("data/2015-03-20T13-39-35-MetatoneOSCLog-6PlayerRehearsal-touches.csv", index_col='time', parse_dates=True)
+from __future__ import print_function
+import os
+import pandas as pd
+import numpy as np
+import time
+import datetime
+touches = pd.read_csv("data/2013-04-20T14-55-00-MetatoneOSCLog-touches-posthoc-gestures.csv", index_col='time', parse_dates=True)
 touches["timediff"] = touches.index
 touches["timediff"] = touches["timediff"] - touches["timediff"].shift()
-long_break = touches.ix[touches["timediff"] > pd.to_timedelta("00:00:30")]
+long_break = touches.ix[touches["timediff"] > pd.to_timedelta("00:02:00")]
 #touches.ix[touches.index.indexer_between_time(long_break.index[1],long_break.index[2])]
 first_touch = touches[:1].index[0]
 frame_list = []

### 2015-08-26 flux-entropy-paper/flux-entropy-paper.tex
added a few refs

--- 2015-08-25 flux-entropy-paper/flux-entropy-paper.tex
@@
 
+\begin{itemize}
+\item Using Markov chains and transition matrices for usability
+  modelling~\cite{Thimbleby:2001kq,Thimbleby:2004fj}
+\item coding livecoding~\cite{Swift:2014tya}
+\end{itemize}
+
 \section{Transition Matrices and Matrix Measures}
@@
 
+\section{A Corpus of Musical Touch-Screen Interactions}
+
+In the following sections, we will use transition matrices, flux, and
+entropy to analyse a corpus of data collected as part of our research
+into collaborative interactions with musical touch-screen tablet
+interfaces. In a period of more than two years (April 2013 - June 2015), we have
+collected data from $139$ sessions where various participants used
+musical touch-screen interfaces on Apple iPad devices. Each session
+took place in a number of contexts including rehearsals, performances, lab study
+sessions, classes, and casual demonstrations. Participants performed
+free improvisations as well as written compositions and performed on
+iPads alone or using setups of iPads and acoustic percussion
+instruments. The median length of these sessions was 7m32s (minimum was
+1m0s and maximum was 26m) and the
+median number of participants was four (minimum 1, maximum 9).
+
 \section{Modelling Performer Ratings with Flux and Entropy}

### 2015-08-26 metatone_post_hoc_analysis.py
prototyped longest break method

@@ -41,7 +41,7 @@ class MetatoneTouchLog:
             self.gestures.to_csv(gestures_path)
         self.ensemble_transition_matrix = transitions.calculate_full_group_transition_matrix(self.gestures)
         self.ensemble_transition_matrix = transitions.transition_matrix_to_stochastic_matrix(self.ensemble_transition_matrix)
-
+        self.longest_break = self.find_longest_breaks()
 
     def first_touch_timestamp(self):
         """
@@ -107,6 +107,33 @@ class MetatoneTouchLog:
         plot_title = time.strftime('%Y-%m-%dT%H-%M-%S', performance_time) + "-gesture-plot"
         PlotMetatonePerformanceAndTransitions.plot_gesture_only_score(plot_title, self.gestures)
         print("Saved gesture plot: " + plot_title)
+        
+    def find_long_breaks(self):
+        """
+        Finds breaks longer than 2 minutes in a touch log. Prints a warning!
+        """
+        self.touches["timediff"] = self.touches.index
+        self.touches["timediff"] = self.touches["timediff"] - self.touches["timediff"].shift()
+        long_break = self.touches.ix[self.touches["timediff"] > pd.to_timedelta("00:03:00")]
+        if not long_break.empty:
+            print("WARNING: Long Break in Performance: " + self.performance_title)
+            print("Printing autosplit touch files.")
+            print(str(long_break["timediff"]))
+            first_touch = self.touches[:1].index[0]
+            frame_list = []
+            for last_touch in long_break.index:
+                part_frame = self.touches.ix[self.touches.index.indexer_between_time(first_touch,last_touch)]
+                first_touch = last_touch
+                frame_list.append(part_frame[:-1]) ## this makes it left closed, right open
+            # then add the rest:
+            last_touch = self.touches.index[-1]
+            part_frame = self.touches.ix[self.touches.index.indexer_between_time(first_touch,last_touch)] #inclusive
+            frame_list.append(part_frame) ## the last one is closed on both sides!
+            # now write each one to csv. (forget about the time_diff column)
+            for frame in frame_list:
+                frame_name = frame.index[0].strftime('%Y-%m-%dT%H-%M-%S') + "-MetatoneOSCLog-autosplit-touches.csv"
+                frame[['device_id','x_pos','y_pos','velocity']].to_csv(frame_name)
+        return 0
 
 def main():
     """Load up all the performances and do some stats"""
@@ -138,9 +165,9 @@ def main():
[…]

### 2015-08-26 R/analysis.R
very strange plot

@@ -22,7 +22,8 @@ ggplot(reh.perf.studies.df, aes(performance_type, entropy)) + geom_point(aes(col
 
 
 ggplot(df, aes(flux, entropy)) + geom_point(aes(colour = instruments)) + facet_wrap(~instruments)
-ggplot(df, aes(flux, entropy)) + geom_point(aes(size = number_performers, colour = performance_type))
+
+ggplot(df, aes(flux, 2 ^ entropy)) + geom_point(aes(size = number_performers, colour = performance_type))
 
 ggplot(df, aes(flux, number_performers)) + geom_point(aes(colour = performance_type))
 
@@ -44,17 +45,10 @@ summary(aov(flux~performance_context*performance_type*instruments, df))
 
 
 summary(aov(exp(entropy) ~ performance_context * performance_type * instruments, df))
-summary(aov((2 ^ entropy)~performance_context*performance_type*instruments, df))
 
-ggplot(reh.perf.studies.df, aes(performance_context, entropy, colour = performance_type, shape = instruments, fill = instruments)) + geom_boxplot() + geom_point(alpha = 0.5, position = position_jitter(w = 0.1, h = 0))  + scale_fill_manual(values = chifig.3colours) +scale_color_manual(values = chifig.2colours) + scale_y_log()
+summary(aov( entropy ~ performance_context*performance_type*instruments, df))
 
-
-+ stat_summary(fun.data = "mean_cl_boot", position=position_jitter(w = 0.2, h=0))
-                                                                                                                                                                                                                                 
-
-                                                                                                                                                                                                                                        
-                                                                                                                                                                                                                               
-
+ggplot(reh.perf.studies.df, aes(performance_context, flux, colour = performance_type, shape = instruments, fill = instruments)) + geom_violin()+ geom_boxplot() + geom_point(alpha = 0.5, position = position_jitter(w = 0.1, h = 0))  + scale_fill_manual(values = chifig.3colours) +scale_color_manual(values = chifig.2colours) + stat_summary(fun.data = "mean_cl_boot", position=position_jitter(w = 0.2, h=0))
 
 summary(aov(entropy ~ performance_context * performance_type * instruments, reh.perf.studies.df))
 summary(aov(exp(entropy)~performance_context*performance_type*instruments, reh.perf.studies.df))

### 2015-08-26 R/analysis.R
yes! some more results in the aov

@@ -7,6 +7,9 @@ df <- read.csv("../metatone-performance-data.csv")
 df$time <- strptime(as.character(df$time), "%Y-%m-%d %H:%M:%S.%OS")
 summary(df)
 
+touch.sessions <- subset(df, performance_type %in% c("composition","improvisation"))
+summary(touch.sessions)
+
 ## rehearsals/performances/studies only
 reh.perf.studies.df <- subset(df, performance_context %in% c("rehearsal", "performance", "study"))
 reh.perf.studies.df <- subset(reh.perf.studies.df, performance_type %in% c("composition","improvisation"))
@@ -41,7 +44,14 @@ head(df)
 
 ## Question: Do Performance Type and Context have a significant effect on Flux and Entropy?
 summary(aov(flux~performance_context*performance_type*instruments, reh.perf.studies.df))
-summary(aov(flux~performance_context*performance_type*instruments, df))
+summary(aov(entropy~performance_context*performance_type*instruments, reh.perf.studies.df))
+                                        # yes! this is a result.
+# performance_type and performance_type:instruments both have significant effects
+
+
+summary(aov(flux~performance_context*performance_type*instruments, touch.sessions))
+summary(aov(entropy~performance_context*performance_type*instruments, touch.sessions))
+# entropy has another main effect for performance_context!
 
 
 summary(aov(exp(entropy) ~ performance_context * performance_type * instruments, df))

### 2015-08-27 metatone_post_hoc_analysis.py
fixed up error in instruments column

@@ -149,6 +149,7 @@ def main():
     ## Also load up the experiment design dataframe to merge with the data!
     print("Loading performance information frame.")
     performance_information = pd.read_csv("metatone-performance-information.csv", index_col='time', parse_dates=True)
+    performance_information = performance_information[['performance_context', 'performance_type', 'instruments', 'notes']]
 
     print("Generating the performance data frame.")
     perf_names = {}
@@ -162,12 +163,11 @@ def main():
             }})
     perf_frame = pd.DataFrame.from_dict(perf_names, orient="index")
     perf_frame = pd.concat([performance_information,perf_frame], axis = 1)
-    #perf_frame['performance_context','performance_type','instruments','notes','video_location'] = performance_information['performance_context','performance_type','instruments','notes','video_location']
     perf_frame.to_csv("metatone-performance-data.csv")
 
-    print("Creating Gesture Scores.")
-    for perf in performances:
-        perf.print_gesture_score() ## Prints out a gesture-score pdf for reference.
+    #print("Creating Gesture Scores.")
+    #for perf in performances:
+    #    perf.print_gesture_score() ## Prints out a gesture-score pdf for reference.
 
     # print("Finding the lengths.")
     # performer_length_dict = {}

### 2015-08-27 metatone_post_hoc_analysis.py
compute a few extra matrix measures

@@ -41,7 +41,7 @@ class MetatoneTouchLog:
             self.gestures.to_csv(gestures_path)
         self.ensemble_transition_matrix = transitions.calculate_full_group_transition_matrix(self.gestures)
         self.ensemble_transition_matrix = transitions.transition_matrix_to_stochastic_matrix(self.ensemble_transition_matrix)
-        self.longest_break = self.find_longest_breaks()
+        self.longest_break = self.find_long_breaks()
 
     def first_touch_timestamp(self):
         """
@@ -72,6 +72,15 @@ class MetatoneTouchLog:
         Returns the entropy of the whole ensemble transition matrix.
         """
         return transitions.entropy_measure(self.ensemble_transition_matrix)
+
+    def ensemble_determinant(self):
+        return np.linalg.det(self.ensemble_transition_matrix)
+
+    def ensemble_norm(self):
+        return np.linalg.norm(self.ensemble_transition_matrix, ord="fro")
+
+    def ensemble_trace(self):
+        return np.linalg.trace(self.ensemble_transition_matrix)
 
     def performance_length(self):
         """
@@ -159,7 +168,10 @@ def main():
             "number_performers": perf.number_performers(),
             "length_seconds": perf.performance_length(),
             "flux": perf.ensemble_flux(),
-            "entropy": perf.ensemble_entropy()
+            "entropy": perf.ensemble_entropy(),
+            "determinant": perf.ensemble_determinant(),
+            "norm": perf.ensemble_norm(),
+            "trace": perf.ensemble_trace()
             }})
     perf_frame = pd.DataFrame.from_dict(perf_names, orient="index")
     perf_frame = pd.concat([performance_information,perf_frame], axis = 1)

### 2015-08-27 flux-entropy-paper/flux-entropy-paper.tex
new calculations for other matrix measures

@@
 interfaces. In a period of more than two years (April 2013 - June 2015), we have
-collected data from $139$ sessions where various participants used
+collected data from $150$ sessions where various participants used
 musical touch-screen interfaces on Apple iPad devices. Each session
@@
 iPads alone or using setups of iPads and acoustic percussion
-instruments. The median length of these sessions was 7m32s (minimum was
+instruments. The median length of these sessions was 7m28s (minimum was
 1m0s and maximum was 26m) and the
-median number of participants was four (minimum 1, maximum 9).
+median number of participants was four (minimum 1, maximum 7).
+
+\begin{figure}
+  \centering
+  \includegraphics[width=\linewidth]{figures/sessions-count}
+  \caption{
+    \label{fig:count-data}}
+\end{figure}
+
+\begin{figure}
+  \centering
+  \includegraphics[width=\linewidth]{figures/flux-boxplot}
+  \caption{
+    \label{fig:flux-boxplot}}
+\end{figure}
+
+\begin{figure}
+  \centering
+  \includegraphics[width=\linewidth]{figures/entropy-boxplot}
+  \caption{
+    \label{fig:entropy-boxplot}}
+\end{figure}
+
+\begin{figure}
+  \centering
+  \includegraphics[width=\linewidth]{figures/flux-entropy-distribution}
+  \caption{
+    \label{fig:flux-entropy-distribution}}
+\end{figure}
+
+A three-way ANOVA procedure was performed on the data set to find
+whether performance context, type, and instrumentation could predict
+the entropy and flux values.
+
+Using performative sessions only (i.e. not the demonstration
+sessions). We found a significant main effect on flux due to
+performance type ($F(1,86) = 5.87, p < 0.05$), and interaction
+effect due to performance type and instrumentation ($F(1,86) = 12.37, p
+< 0.001$). Similar effects on entropy were found due to performance
+type ($F(1,86) = 13.62, p < 0.001$) and an interaction effect due to
+performance type and instrumentation ($F(1,86) = 4.98, p < 0.05$). The
+effect on entropy due to performance context (rehearsal,
+performance, or study) was not found to be significant at $\alpha =
+0.05$, ($F(2,86) = 2.11, p = 0.128$). When including demonstration
+sessions in the analysis, however, performance context was found to
+have a significant main effect on entropy ($F(4,138) = 5.07, p <
+0.001$).
 

### 2015-08-27 R/analysis.R

@@ -3,62 +3,65 @@ library(MASS)
 chifig.3colours <- c("#e41a1c", "#377eb8", "#4daf4a")
 chifig.2colours <- c("#984ea3", "#ff7f00")
 ## time,filename,number_performers,performance_context,performance_type,instruments,notes,video_location,flux,number_performers,length_seconds,entropy,filename
-df <- read.csv("../metatone-performance-data.csv")
-df$time <- strptime(as.character(df$time), "%Y-%m-%d %H:%M:%S.%OS")
-summary(df)
-
-touch.sessions <- subset(df, performance_type %in% c("composition","improvisation"))
-summary(touch.sessions)
-
+raw.performance.data <- read.csv("../metatone-performance-data.csv")
+raw.performance.data$time <- strptime(as.character(raw.performance.data$time), "%Y-%m-%d %H:%M:%S.%OS")
+valid.sessions <- subset(raw.performance.data, performance_type %in% c("composition","improvisation"))
+flux.entropy.ratings <- read.csv("../flux_entropy_ratings.csv")
+#summary(raw.performance.data)
+#summary(valid.sessions)
 ## rehearsals/performances/studies only
-reh.perf.studies.df <- subset(df, performance_context %in% c("rehearsal", "performance", "study"))
-reh.perf.studies.df <- subset(reh.perf.studies.df, performance_type %in% c("composition","improvisation"))
-summary(reh.perf.studies.df$performance_type)
-summary(reh.perf.studies.df)
+perf.contexts <- subset(raw.performance.data, performance_context %in% c("rehearsal", "performance", "study"))
+perf.contexts <- subset(perf.contexts, performance_type %in% c("composition","improvisation"))
+## 
+summary(perf.contexts$performance_type) # Composition: 23 - Improvisation: 72
+summary(perf.contexts)
 ## plots
+# Count of Session Data
+ggplot(valid.sessions, aes(performance_context)) + geom_bar(aes(fill = performance_type))
+#ggplot(perf.contexts, aes(performance_context)) + geom_bar(aes(fill = performance_type))
+#
+ggplot(valid.sessions, aes(performance_context, flux)) + geom_boxplot(aes(fill = performance_type))
+ggplot(valid.sessions, aes(performance_context, entropy)) +  geom_boxplot(aes(fill = performance_type))
 
-ggplot(reh.perf.studies.df, aes(performance_type)) + geom_bar(aes(fill = performance_context))
[…]

### 2015-08-27 metatone_post_hoc_analysis.py

@@ -80,7 +80,7 @@ class MetatoneTouchLog:
         return np.linalg.norm(self.ensemble_transition_matrix, ord="fro")
 
     def ensemble_trace(self):
-        return np.linalg.trace(self.ensemble_transition_matrix)
+        return np.trace(self.ensemble_transition_matrix)
 
     def performance_length(self):
         """

### 2015-08-27 R/analysis.R
trying out different TM methods

@@ -3,7 +3,9 @@ library(MASS)
 chifig.3colours <- c("#e41a1c", "#377eb8", "#4daf4a")
 chifig.2colours <- c("#984ea3", "#ff7f00")
 ## time,filename,number_performers,performance_context,performance_type,instruments,notes,video_location,flux,number_performers,length_seconds,entropy,filename
-raw.performance.data <- read.csv("../metatone-performance-data.csv")
+#raw.performance.data <- read.csv("../metatone-performance-data-stochasticmatrices.csv")
+raw.performance.data <- read.csv("../metatone-performance-data-normalmatrices.csv")
+#raw.performance.data <- read.csv("../metatone-performance-data.csv")
 raw.performance.data$time <- strptime(as.character(raw.performance.data$time), "%Y-%m-%d %H:%M:%S.%OS")
 valid.sessions <- subset(raw.performance.data, performance_type %in% c("composition","improvisation"))
 flux.entropy.ratings <- read.csv("../flux_entropy_ratings.csv")
@@ -24,7 +26,11 @@ ggplot(valid.sessions, aes(performance_context, flux)) + geom_boxplot(aes(fill =
 ggplot(valid.sessions, aes(performance_context, entropy)) +  geom_boxplot(aes(fill = performance_type))
 
 ggplot(valid.sessions, aes(performance_context, trace)) + geom_boxplot(aes(fill = performance_type))
-+ scale_y_log10()
+ggplot(valid.sessions, aes(performance_context, norm)) + geom_boxplot(aes(fill = performance_type))
+ggplot(valid.sessions, aes(performance_context, determinant)) + geom_boxplot(aes(fill = performance_type))
+ggplot(valid.sessions, aes(performance_context, number_performers)) + geom_boxplot(aes(fill = performance_type))
+ggplot(valid.sessions, aes(performance_context, length_seconds)) + geom_boxplot(aes(fill = performance_type))
+
 # distribution of flux and entropy
 ggplot(valid.sessions, aes(flux, entropy)) + geom_point(aes(colour = performance_type, shape = instruments)) + facet_wrap(~performance_context)
 #ggplot(df, aes(flux, 2 ^ entropy)) + geom_point(aes(size = number_performers, colour = performance_type))
@@ -34,18 +40,20 @@ ggplot(valid.sessions, aes(flux, entropy)) + geom_point(aes(colour = performance
 
 # Flux through time.
[…]

### 2015-08-27 metatone_post_hoc_analysis.py

@@ -41,6 +41,7 @@ class MetatoneTouchLog:
             self.gestures.to_csv(gestures_path)
         self.ensemble_transition_matrix = transitions.calculate_full_group_transition_matrix(self.gestures)
         self.ensemble_transition_matrix = transitions.transition_matrix_to_stochastic_matrix(self.ensemble_transition_matrix)
+        #self.ensemble_transition_matrix = transitions.transition_matrix_to_normal_transition_matrix(self.ensemble_transition_matrix)
         self.longest_break = self.find_long_breaks()
 
     def first_touch_timestamp(self):
@@ -81,6 +82,9 @@ class MetatoneTouchLog:
 
     def ensemble_trace(self):
         return np.trace(self.ensemble_transition_matrix)
+
+    def ensemble_one_norm(self):
+        return sum(sum(self.ensemble_transition_matrix))
 
     def performance_length(self):
         """

### 2015-08-27 R/analysis.R
merge session metadata into time-windowed stats df

@@ -108,11 +108,18 @@ ggplot(flux.entropy.ratings, aes(flux,rating)) + geom_smooth(method=lm) + geom_p
 
 library("plyr")
 library("reshape2")
-
+library("stringr")
 source("tm.R")
 
-filenames <- llply(list.files("../data"),
-                  function(x) c(x, names(read.csv(paste("../data/", x, sep = "")))))
+## TODO use metatone-performance-data.csv to get metadata, use
+## filename as a foreign key into the gestures data csvs (append
+## -gesture to filename)
+
+metadata <- read.csv("../metatone-performance-information.csv")[,1:10]
+metadata$time <- strptime(as.character(metadata$time), "%Y-%m-%d %H:%M:%S.%OS")
+metadata$number_performers <- as.numeric(metadata$number_performers)
+metadata$length_seconds <- as.numeric(metadata$length_seconds)
+metadata$notes <- as.character(metadata$notes)
 
 process_session <- function(filename, session_df){
     ## assume first column is filename, second is time
@@ -131,19 +138,26 @@ df <- ldply(list.files("../data"),
 df$gesture <- ordered(gestures[df$gesture+1], levels = gestures)
 ## add time-windowed chunk section
 df$section <- ordered(c("beginning", "middle", "end"), levels = c("beginning", "middle", "end"))[as.integer(df$time*3)+1]
+## return filename to "base" filename
+df$filename <- str_replace(df$filename, "-touches-posthoc-gestures.csv", "")
 
 ## Transition Matrices
 tm <- ddply(df, .(filename, musician, section), calculate.transitions)
 ## add flux, entropy
 tm <- ddply(tm, .(filename, musician, section),
             function(x){
-                data.frame(x,
-                           flux = transition.flux(x),
+                data.frame(flux = transition.flux(x),
                            entropy = transition.entropy(x))
             })
+## average flux/entropy over whole group
+tm <- ddply(tm, .(filename, section), summarise, flux = mean(flux), entropy = mean(entropy))
+tm <- merge(tm, metadata, by = "filename", all.x = TRUE)
 
 ## plotting
 library("ggplot2")
 
 ggplot(tm, aes(section, flux)) + geom_violin(aes(fill=section))
 ggplot(tm, aes(section, entropy)) + geom_violin(aes(fill=section))
+
+ggplot(tm, aes(section, flux)) + geom_boxplot(aes(fill=section))
+ggplot(tm, aes(section, entropy)) + geom_boxplot(aes(fill=section))

### 2015-08-27 R/analysis.R
a couple more plots

@@ -156,8 +156,13 @@ tm <- merge(tm, metadata, by = "filename", all.x = TRUE)
 ## plotting
 library("ggplot2")
 
-ggplot(tm, aes(section, flux)) + geom_violin(aes(fill=section))
-ggplot(tm, aes(section, entropy)) + geom_violin(aes(fill=section))
+ggplot(tm, aes(section, flux)) + geom_jitter(aes(color=performance_context, size=number_performers, shape=performance_type), alpha = 0.5)
+
+ggplot(tm, aes(performance_context, flux)) + geom_boxplot(aes(fill=section)) + geom_point(alpha=.05, size = 5)
+
+ggsave("../flux-by-section.pdf")
+
+ggplot(tm, aes(section, entropy)) + geom_violin(aes(fill=performance_context))
 
 ggplot(tm, aes(section, flux)) + geom_boxplot(aes(fill=section))
 ggplot(tm, aes(section, entropy)) + geom_boxplot(aes(fill=section))

### 2015-08-31 flux-entropy-paper/flux-entropy-paper.tex
lots of new content for the paper

--- 2015-08-26 flux-entropy-paper/flux-entropy-paper.tex
@@
 
+\begin{figure*}
+  \centering
+  \includegraphics[width=\linewidth]{figures/three-performance-contexts}
+  \caption{Collaborative musical interaction with our touch-screen
+    interfaces in three different contexts: (left to right) rehearsal
+    in a studio,
+    performance in front of an audience, and demonstration at a
+    research poster session. All of these session were improvised,
+    and in the performance session, the musicians used percussion
+    instruments as well as iPads.
+    \label{fig:three-performance-contexts}}
+\end{figure*}
+
 Finally, we report a quantitative analysis of the transition matrices
@@
 
-In the following sections, we will use transition matrices, flux, and
-entropy to analyse a corpus of data collected as part of our research
-into collaborative interactions with musical touch-screen tablet
-interfaces. In a period of more than two years (April 2013 - June 2015), we have
-collected data from $139$ sessions where various participants used
-musical touch-screen interfaces on Apple iPad devices. Each session
-took place in a number of contexts including rehearsals, performances, lab study
-sessions, classes, and casual demonstrations. Participants performed
-free improvisations as well as written compositions and performed on
-iPads alone or using setups of iPads and acoustic percussion
-instruments. The median length of these sessions was 7m32s (minimum was
-1m0s and maximum was 26m) and the
-median number of participants was four (minimum 1, maximum 9).
-
-\section{Modelling Performer Ratings with Flux and Entropy}
+\begin{table}
+%\centering
+\begin{tabular}{l|llll}
+\hline
+            & Total & Min  & Median   & Max     \\ 
+\hline
+N           & 150       &         &        &  \\
+Record size & 306MB     &         &        &   \\
+Length      & 20H34M22S & 1M0S    & 7M29S  & 26M0S\\
+Participants& 524       & 1       & 4      & 9    \\
+Flux        &           & 0.02239 & 0.1445 & 0.3246\\
+\hline
+\end{tabular}
+\caption{
+  Representative data about the corpus of musical touch-screen
+  interaction sessions used in this study. These sessions were
+  rehearsals, performances, studies, and demonstration sessions and the
+  participants improvised or performed written compositions. Each record
+  consisted of touch-data recorded in CSV format and was later analysed
+  with a gestural classification system. Some participants were
+  present in multiple sessions, however this data was not tracked in
+  the corpus.\label{corpus-table}}
+\end{table}
 
@@
   \centering
-  \includegraphics[width=\linewidth]{figures/stat-response-fit.pdf}
-  \caption{Transition matrix statistics (flux above and entropy below) vs Likert responses for
-    questions 3, 6 and 7, all of which exhibited significant
-    relationships under a proportional odds logistic regression model.
-    A least-squares linear model is shown with a dotted line.
-    \label{fig:stat-response-fit}}
+  \includegraphics[width=\linewidth]{figures/sessions-count}
+  \caption{Distribution of the sessions in our interaction corpus by
+    performance context. Improvisations or compositions were repeated
+    several times in rehearsals and study sessions but only performed
+    once in concerts.
+    \label{fig:count-data}}
 \end{figure}
 
-One final question to explore is how these flux and entropy measures
-in the interaction data relate to the performers' perceptions of their
-performance?
+In the following sections, we will use transition matrices and the
+flux measure to analyse a corpus of data collected as part of our research
+into collaborative interactions with musical touch-screen tablet
+interfaces.
 
-To answer this question, we construct models with the flux and entropy
-measures as independent variables and the Likert responses (1 to 5) as
-the dependent variable. These measures exhibited clear relationships
-for three of the survey questions (Questions 3, 6 and 7), which are
-plotted in Figure~\ref{fig:stat-response-fit} where the least-squares
-line of best fit has a positive gradient for both flux and entropy.
+Over a period of more than two years (April 2013 - June 2015), we have
+collected data from $150$ sessions where various participants used
+musical touch-screen interfaces on Apple iPad devices. These
+interfaces all shared a common mode of interaction where the majority
+of the screen could be used for free form touch interactions; tapping
+produced short sounds, while swiping or swirling produced long sounds
+with volume proportional to the velocity of the moving touch point.
+All touch interactions from each participant in a session was recorded
+by sending data to a central server application. This data was stored
+in CSV files recording the date and time, X, and Y locations,
+velocity, and a unique device identifier for each touch event detected
+by the iOS operating system.
 
-Determining the strengths of these positive relationships is
-difficult, both due to the problem of treating ordinal Likert
-responses as numeric~\cite{Gardner:2007fj} and also
-due to the unknown distribution of the entropy and flux statistics. As
-a result, we use a conservative proportional odds logistic regression
-(POLR)~(using the \texttt{polr} function in \texttt{R}, from the MASS
-package \cite{Venables:2002qv}) to treat the Likert responses as an
-ordered factor response.
+Each session took place in a number of contexts including rehearsals,
+performances, lab study sessions, classes, and casual demonstrations,
+some of which are shown in Figure
+\ref{fig:three-performance-contexts}. Participants performed free
+improvisations as well as written compositions and performed on iPads
+alone or using setups of iPads and acoustic percussion instruments.
+The median length of these sessions was 7m29s and the median number of
+participants was four, other descriptive statistics about the sessions
+is shown in Table \ref{corpus-table}. Each session was coded with
+metadata regarding the session context (whether it was a performance,
+rehearsal, study, demonstration, or data collection session), the
+session type (either improvisation or composition), and the
+instruments used (iPads alone, or iPads with acoustic percussion). The
+distribution of recorded sessions by context and type is shown in
+Figure \ref{fig:count-data}.
 
-The POLR results indicate a significant ($p<0.05$) relationship
-between the user-reported Likert response and the flux statistic for
-Question 3 ($t = 2.77, df = 4, p = 0.025$), Question 6
-($t = 2.41, df = 4, p = 0.037$), and Question 7
-($t = 2.49, df = 4, p = 0.034$). The same questions displayed a
-significant relationship for the entropy statistic: Question 3
-($t = 2.93, df = 4, p = 0.022$), Question 6
-($t = 2.69, df = 4, p = 0.028$), and Question 7
-($t = 2.89, df = 4, p = 0.022$). These relationships are shown in
-Figure~\ref{fig:stat-response-fit}.
+To analyse the gestural interactions present in each of these
+sessions, the sessions were automatically coded according to a
+vocabulary of nine simple, continuous, touch-screen gestures. Examples
+of these gestures was generated by an expert user in formal process
+where each gesture was performed for 1-minute and feature vectors of
+discriptive statistics on these examples was used to train a Random
+Forest~\cite{Breiman:2001kx} classifier\footnote{The classifier used
+  was from Python's Scikit-learn package~\cite{scikit-learn}.}. The
+classifier was applied to 5-second windows of the touch-data each
+second provide a chain of gesture-states for each participant in each
+session. These chains can be used to create $9 /times 9$
+gesture-transition matrices for individual performers, and the whole
+ensemble in each session.
 
-It is notable that these questions (3, 6, and 7) all pertain to the
-\emph{perceived effect of the agent} on the performers. As the agent
-uses the flux measure to trigger new-idea messages, it appears that
-performers appreciated its intervention more in particularly creative
-performances where flux and entropy were high.
+% While musical interactions such as those in
+% our corpus are often analysed using qualitative methods, it is
+% problematic to
+
+While the interaction sessions recorded in our corpus were performed
+by many different participants with varying experience levels, in
+different contexts and on different instrumental setups, the common
+interface and method of recording the data allows all the sessions to
+be transformed into gesture sequences. This common format allows us to
+ask if performance styles vary in between these different situations.
+The sequences can be plotted and inspected qualitatively, but
+transition matrices present a quantitative approach to analysing such
+interactions.
+
+In the rest of this paper we will use our matrix measure techniques to
+interrogate this dataset and to answer two questions. First, do the
+transition matrix measures allow us to distinguish between different
+performance contexts, types and instrumental setups? Secondly, do
+these performance measure allow us further insight into the corpus
+than the metadata can provide?
+
+\subsection{Differentiating interaction sessions}
+\label{differentiating-interaction-sessions}
+
+The distribution of flux and entropy with respect to session contexts,
+types and instrumentation are shown in Figures \ref{fig:flux-boxplot},
+\ref{fig:entropy-boxplot}, and \ref{fig:flux-entropy-distribution}. A
+three-way ANOVA procedure was performed on the data set to find
+whether the metatdata could significantly predict the entropy and flux
+values. In the following analysis, only the 95 performative sessions
+were considered (i.e. not the demonstration or data collection
+contexts).
+
+All main effects on flux were found to be significant: performance
+context ($F(2,86) = 6.28, p < 0.01$), performance type ($F(1,86) =
+7.21, p < 0.01$), and instrumentation ($F(1,86) = 20.06, p < 0.001$).
+Significant interaction effects were also found for performance
+context and type ($F(1,86) = 6.18, p < 0.05$), and performance context and
+instrumentation ($F(1,86) = 10.48, p < 0.01$).
+
+For entropy, significant main effects were found due to performance
+context ($F(2,86) = 12.10, p < 0.001$), and instrumentation ($F(1,86)
+= 25.86$). The interaction of performance type and instrumentation was
+also found to be significant ($F(1,86) = 9.93, p<0.01$).
+
+These tests show that the different interaction styles that might be
+present in performance contexts, types and
+instrumentations are reflected in the values of flux and entropy of
+their transition matrices.
+
+% need to add overall contrasts column to do pairwise t-tests.
+
+\begin{figure}
+  \centering
+  \includegraphics[width=\linewidth]{figures/flux-boxplot}
+  \caption{Distributions of flux values by performance context. Study
+    and rehearsal sessions had the highest flux for improvisations
+    which fell during performances. Compositions had lower flux in
+    rehearsals than performances.
+    \label{fig:flux-boxplot}}
+\end{figure}
+
+\begin{figure}
+  \centering
+  \includegraphics[width=\linewidth]{figures/entropy-boxplot}
+  \caption{Distribution of entropy by
+  performance context. Demonstration sessions had much lower entropy
+  indicating data unevenly spread in the transition matrices. This
+  could be because untrained
+  participants did not explore the full gestural potential of the
+  instruments. 
+  \label{fig:entropy-boxplot} }
+\end{figure}
+
+\begin{figure}
+  \centering
+  \includegraphics[width=\linewidth]{figures/flux-entropy-distribution}
+  \caption{Distribution of flux and entropy by performance type and
+    context. Flux and entropy tended to be linearly related in
+    improvisations, particularly in demonstrations and studies, while
+    compositions had a more scattered relationship.
+    \label{fig:flux-entropy-distribution}}
+\end{figure}
+
+\subsection{Breaking performances into sections}
+
+It is clear that flux and entropy can detect the variation in
+performance styles in between performances of different types, but can
+they be used to understand some of the internal structure of
+performances? A simple model for understanding temporal art-forms is
+the ``beginning, middle, end'' structure. To investigate how the
+structure fits session in our corpus, we can divide the
+gestural-sequences of each session into thirds and calculate the
+transition matrices given by each section.
+
+\subsection{Performance styles over time}
+
+\begin{figure}
+  \centering
+  \includegraphics[width=\linewidth]{figures/flux-through-time}
+  \caption{Distribution flux values of sessions through time.
+    Different performance styles are visible as clusters of similar
+    flux regardless of performance context.
+    \label{fig:flux-through-time}}
+\end{figure}
+
+\section{Conclusions}
+
+% \section{Modelling Performer Ratings with Flux and Entropy}
+
+% \begin{figure}
+%   \centering
+%   \includegraphics[width=\linewidth]{figures/stat-response-fit.pdf}
+%   \caption{Transition matrix statistics (flux above and entropy below) vs Likert responses for
+%     questions 3, 6 and 7, all of which exhibited significant
+%     relationships under a proportional odds logistic regression model.
+%     A least-squares linear model is shown with a dotted line.
+%     \label{fig:stat-response-fit}}
+% \end{figure}
+
+% One final question to explore is how these flux and entropy measures
+% in the interaction data relate to the performers' perceptions of their
+% performance?
+
+% To answer this question, we construct models with the flux and entropy
+% measures as independent variables and the Likert responses (1 to 5) as
+% the dependent variable. These measures exhibited clear relationships
+% for three of the survey questions (Questions 3, 6 and 7), which are
+% plotted in Figure~\ref{fig:stat-response-fit} where the least-squares
+% line of best fit has a positive gradient for both flux and entropy.
+
+% Determining the strengths of these positive relationships is
+% difficult, both due to the problem of treating ordinal Likert
+% responses as numeric~\cite{Gardner:2007fj} and also
+% due to the unknown distribution of the entropy and flux statistics. As
+% a result, we use a conservative proportional odds logistic regression
+% (POLR)~(using the \texttt{polr} function in \texttt{R}, from the MASS
+% package \cite{Venables:2002qv}) to treat the Likert responses as an
+% ordered factor response.
+
+% The POLR results indicate a significant ($p<0.05$) relationship
+% between the user-reported Likert response and the flux statistic for
+% Question 3 ($t = 2.77, df = 4, p = 0.025$), Question 6
+% ($t = 2.41, df = 4, p = 0.037$), and Question 7
+% ($t = 2.49, df = 4, p = 0.034$). The same questions displayed a
+% significant relationship for the entropy statistic: Question 3
+% ($t = 2.93, df = 4, p = 0.022$), Question 6
+% ($t = 2.69, df = 4, p = 0.028$), and Question 7
+% ($t = 2.89, df = 4, p = 0.022$). These relationships are shown in
+% Figure~\ref{fig:stat-response-fit}.
+
+% It is notable that these questions (3, 6, and 7) all pertain to the
+% \emph{perceived effect of the agent} on the performers. As the agent
+% uses the flux measure to trigger new-idea messages, it appears that
+% performers appreciated its intervention more in particularly creative
+% performances where flux and entropy were high.
 

### 2015-08-31 R/analysis.R
updates to produce publication plots

@@ -1,7 +1,9 @@
 library(ggplot2)
 library(MASS)
+library(grid)
 chifig.3colours <- c("#e41a1c", "#377eb8", "#4daf4a")
 chifig.2colours <- c("#984ea3", "#ff7f00")
+chifig.5colours <- c("#e41a1c", "#377eb8", "#4daf4a", "#984ea3", "#ff7f00")
 ## time,filename,number_performers,performance_context,performance_type,instruments,notes,video_location,flux,number_performers,length_seconds,entropy,filename
 #raw.performance.data <- read.csv("../metatone-performance-data-stochasticmatrices.csv")
 raw.performance.data <- read.csv("../metatone-performance-data-normalmatrices.csv")
@@ -10,37 +12,54 @@ raw.performance.data$time <- strptime(as.character(raw.performance.data$time), "
 valid.sessions <- subset(raw.performance.data, performance_type %in% c("composition","improvisation"))
 flux.entropy.ratings <- read.csv("../flux_entropy_ratings.csv")
 #summary(raw.performance.data)
-#summary(valid.sessions)
+summary(valid.sessions)
 ## rehearsals/performances/studies only
 perf.contexts <- subset(raw.performance.data, performance_context %in% c("rehearsal", "performance", "study"))
 perf.contexts <- subset(perf.contexts, performance_type %in% c("composition","improvisation"))
-## 
 summary(perf.contexts$performance_type) # Composition: 23 - Improvisation: 72
 summary(perf.contexts)
+                                        # Calculate total recorded time.
+total.seconds <- max(valid.sessions$length_seconds)
+message(paste(total.seconds %/% 3600, "H", (total.seconds %% 3600) %/% 60, "M", signif(total.seconds %% 60, 4), "S", sep = "" ))
+                                        # performer data
+message(paste(sum(valid.sessions$number_performers),min(valid.sessions$number_performers),median(valid.sessions$number_performers),max(valid.sessions$number_performers),sep = " & "))
+                                        #
+message(paste(signif(sum(valid.sessions$flux),4),signif(min(valid.sessions$flux),4),signif(median(valid.sessions$flux),4),signif(max(valid.sessions$flux),4),sep = " & "))
+
+# cut out data_collection
+#valid.sessions <- (subset(valid.sessions, performance_context != "data_collection"))
 ## plots
 # Count of Session Data
[…]

### 2015-08-31 R/analysis.R
ben updates: pushed at Chuckles' request

@@ -129,11 +129,8 @@ ggplot(flux.entropy.ratings, aes(flux,rating)) + geom_smooth(method=lm) + geom_p
 library("plyr")
 library("reshape2")
 library("stringr")
+library("ggplot2")
 source("tm.R")
-
-## TODO use metatone-performance-data.csv to get metadata, use
-## filename as a foreign key into the gestures data csvs (append
-## -gesture to filename)
 
 metadata <- read.csv("../metatone-performance-information.csv")[,1:10]
 metadata$time <- strptime(as.character(metadata$time), "%Y-%m-%d %H:%M:%S.%OS")
@@ -172,13 +169,21 @@ tm <- ddply(tm, .(filename, musician, section),
 ## average flux/entropy over whole group
 tm <- ddply(tm, .(filename, section), summarise, flux = mean(flux), entropy = mean(entropy))
 tm <- merge(tm, metadata, by = "filename", all.x = TRUE)
+## remove data which doesn't make sense in a  beginning-middle-end sense
+tm <- subset(tm, performance_type!="fail")
+tm <- subset(tm, performance_context %in% c("performance", "rehearsal", "study"))
 
 ## plotting
-library("ggplot2")
 
 ggplot(tm, aes(section, flux)) + geom_jitter(aes(color=performance_context, size=number_performers, shape=performance_type), alpha = 0.5)
 
-ggplot(tm, aes(performance_context, flux)) + geom_boxplot(aes(fill=section)) + geom_point(alpha=.05, size = 5)
+ggplot(tm, aes(performance_type, flux)) + geom_boxplot(aes(fill=section)) + geom_point(alpha=.05, size = 5)
+
+summary(aov(flux~performance_type*section, tm))
+
+ggplot(tm, aes(as.numeric(section), flux)) + geom_boxplot(aes(fill=section)) + geom_point(alpha=.05, size = 5) + stat_smooth(method="lm") + facet_wrap(~performance_context)
+
+ggplot(tm, aes(1, flux)) + geom_boxplot(aes(fill=section)) + geom_point(alpha=.05, size = 5)
 
 ggsave("../flux-by-section.pdf")
 
@@ -186,3 +191,5 @@ ggplot(tm, aes(section, entropy)) + geom_violin(aes(fill=performance_context))
 
 ggplot(tm, aes(section, flux)) + geom_boxplot(aes(fill=section))
 ggplot(tm, aes(section, entropy)) + geom_boxplot(aes(fill=section))
+
+## things!

### 2015-08-31 R/tm.R

@@ -42,8 +42,9 @@ calculate.transitions <- function(df, left.stochastic=FALSE){
 ## }
 
 transition.flux <- function(df){
-    if(sum(df$prob)!=1)
-        message("warning: prob sum isn't 1")
+    if(sum(df$prob)!=1){
+        message("warning: prob sum isn't 1 - sum = ", sum(df$prob))
+    }
     on.diag.sum <- sum(df[df$from==df$to, "prob"])
     data.frame(flux = 1-on.diag.sum)
 }

### 2015-08-31 flux-entropy-paper/flux-entropy-paper.tex
added last two analysis sections

@@
 
+While the interaction session in our corpus have many different
+contexts, they are all part of a program of artistic and HCI research
+that has evolved over time. Figures \ref{fig:flux-through-time} and
+\ref{fig:entropy-through-time} show the values of flux and entropy for
+each session in the corpus. Over the course of this research program,
+different ensembles and different activities have been emphasised,
+some of these differences can be seen as clusters of performances with
+similar flux and entropy in the two plots. For instance, the earliest
+series of rehearsals and performances had flux between 0.05 and 0.15,
+while the period of activity after July 2014 had flux between 0.1 and
+0.25. This change corresponded with an intensive series of study
+sessions and rehearsals where the participants developed a more
+fluctuating style of performance. The entropy between these two
+periods, however, remains at a similiar level, showing the the space
+of available gestures that the performers explored was more constant.
+
 \begin{figure}
@@
   \includegraphics[width=\linewidth]{figures/flux-through-time}
-  \caption{Distribution flux values of sessions through time.
+  \caption{Distribution of flux values of sessions through time.
     Different performance styles are visible as clusters of similar
@@
     \label{fig:flux-through-time}}
+\end{figure}
+
+\begin{figure}
+  \centering
+  \includegraphics[width=\linewidth]{figures/entropy-through-time}
+  \caption{Distribution of entropy values of sessions through time.
+    \label{fig:entropy-through-time}}
 \end{figure}

### 2015-08-31 R/analysis.R
plot of entropy over time

@@ -58,6 +58,9 @@ quartz.save("../flux-entropy-paper/figures/flux-entropy-distribution.pdf", type 
 ggplot(valid.sessions, aes(time, flux)) + geom_point(aes(colour = performance_context,shape = performance_type), alpha = 0.8) + theme(plot.margin=unit(rep(0,4), "cm"), legend.position = "right", legend.box = "vertical") + scale_colour_manual(values=chifig.5colours)
 quartz.save("../flux-entropy-paper/figures/flux-through-time.pdf", type = "pdf")
 
+ggplot(valid.sessions, aes(time, entropy)) + geom_point(aes(colour = performance_context,shape = performance_type), alpha = 0.8) + theme(plot.margin=unit(rep(0,4), "cm"), legend.position = "right", legend.box = "vertical") + scale_colour_manual(values=chifig.5colours)
+quartz.save("../flux-entropy-paper/figures/entropy-through-time.pdf", type = "pdf")
+
 
                                         # Entropy through time
 ggplot(raw.performance.data, aes(time, entropy)) + geom_point(aes(size = number_performers, colour = performance_context,shape = performance_type), alpha = 0.6)

### 2015-08-31 R/analysis.R
cleaned up some crap

@@ -26,6 +26,8 @@ message(paste(sum(valid.sessions$number_performers),min(valid.sessions$number_pe
                                         #
 message(paste(signif(sum(valid.sessions$flux),4),signif(min(valid.sessions$flux),4),signif(median(valid.sessions$flux),4),signif(max(valid.sessions$flux),4),sep = " & "))
 
+
+## Initial Plots
 # cut out data_collection
 #valid.sessions <- (subset(valid.sessions, performance_context != "data_collection"))
 ## plots
@@ -45,7 +47,6 @@ ggplot(valid.sessions, aes(performance_context, norm)) + geom_boxplot(aes(fill =
 ggplot(perf.contexts, aes(performance_context, determinant)) + geom_boxplot(aes(fill = performance_type))
 ggplot(valid.sessions, aes(performance_context, number_performers)) + geom_boxplot(aes(fill = performance_type))
 ggplot(valid.sessions, aes(performance_context, length_seconds)) + geom_boxplot(aes(fill = performance_type))
-
 # distribution of flux and entropy
 ggplot((subset(valid.sessions, performance_context != "data_collection")), aes(flux, entropy)) + geom_point(aes(colour = performance_type, shape = instruments), alpha = 0.8)  + facet_wrap(~performance_context) + theme(plot.margin=unit(rep(0,4), "cm"), legend.position = "right", legend.box = "vertical") + scale_colour_manual(values=chifig.2colours)
 quartz.save("../flux-entropy-paper/figures/flux-entropy-distribution.pdf", type = "pdf")
@@ -57,13 +58,9 @@ quartz.save("../flux-entropy-paper/figures/flux-entropy-distribution.pdf", type 
 # Flux through time.
 ggplot(valid.sessions, aes(time, flux)) + geom_point(aes(colour = performance_context,shape = performance_type), alpha = 0.8) + theme(plot.margin=unit(rep(0,4), "cm"), legend.position = "right", legend.box = "vertical") + scale_colour_manual(values=chifig.5colours)
 quartz.save("../flux-entropy-paper/figures/flux-through-time.pdf", type = "pdf")
-
+# Entropy through time.
[…]

### 2015-08-31 R/analysis.R
add some figures

@@ -105,6 +105,7 @@ library("plyr")
 library("reshape2")
 library("stringr")
 library("ggplot2")
+library("grid")
 source("tm.R")
 
 metadata <- read.csv("../metatone-performance-information.csv")[,1:10]
@@ -156,11 +157,17 @@ ggplot(tm, aes(performance_type, flux)) + geom_boxplot(aes(fill=section)) + geom
 
 summary(aov(flux~performance_type*section, tm))
 
-ggplot(tm, aes(as.numeric(section), flux)) + geom_boxplot(aes(fill=section)) + geom_point(alpha=.05, size = 5) + stat_smooth(method="lm") + facet_wrap(~performance_context)
-
-ggplot(tm, aes(1, flux)) + geom_boxplot(aes(fill=section)) + geom_point(alpha=.05, size = 5)
+ggplot(tm, aes(section, flux)) + geom_boxplot(aes(fill=section)) + facet_wrap(~performance_context) + scale_fill_manual(values=chifig.3colours) + theme(plot.margin=unit(rep(0,4), "cm"), legend.position = "top", legend.box = "horizontal")
 
 ggsave("../flux-by-section.pdf")
+
+ggplot(tm, aes(section, entropy)) + geom_boxplot(aes(fill=section)) + facet_wrap(~performance_context) + scale_fill_manual(values=chifig.3colours) + theme(plot.margin=unit(rep(0,4), "cm"), legend.position = "top", legend.box = "horizontal")
+
+ggsave("../entropy-by-section.pdf")
+
+
+
+ggplot(tm, aes(1, flux)) + geom_boxplot(aes(fill=section)) + geom_point(alpha=.05, size = 5)
 
 ggplot(tm, aes(section, entropy)) + geom_violin(aes(fill=performance_context))
 

### 2015-08-31 R/analysis.R
more better plots for section analysis

@@ -85,22 +85,21 @@ summary(aov(entropy~performance_context, valid.sessions))
 ggplot(perf.contexts, aes(performance_context, flux, colour = performance_type, shape = instruments, fill = instruments)) + geom_violin()+ geom_boxplot() + geom_point(alpha = 0.5, position = position_jitter(w = 0.1, h = 0))  + scale_fill_manual(values = chifig.3colours) +scale_color_manual(values = chifig.2colours) + stat_summary(fun.data = "mean_cl_boot", position=position_jitter(w = 0.2, h=0))
 
 # Potential POLR stuff.
-#flux.entropy.ratings <- read.csv("../flux_entropy_ratings.csv")
-#flux.entropy.ratings
-
+flux.entropy.ratings <- read.csv("../flux_entropy_ratings.csv")
+head(flux.entropy.ratings)
 summary(aov(rating ~ flux, data = flux.entropy.ratings))
 summary(aov(rating ~ entropy, data = flux.entropy.ratings))
-
 mod <- polr(factor(rating) ~ flux, data = flux.entropy.ratings)
-
 ## get a p-value. this is filthy, but MM said it was ok.  so blame him.
 message(paste("($t = ", format(summary(mod)$coefficients[1,3], digits = 3), ", df = ", format(mod$edf, digits = 3), ", p = ", format(pt(summary(mod)$coefficients[1,3], mod$edf, lower.tail = FALSE), digits = 2), "$)", sep = ""))
-
 ggplot(flux.entropy.ratings, aes(as.factor(rating),flux)) + geom_point() + geom_boxplot()
 ggplot(flux.entropy.ratings, aes(flux,rating)) + geom_smooth(method=lm) + geom_point()
 
-## time-window chunking (beginning-middle-end)
 
+
+#######
+## time-window chunking (beginning-middle-end)
+#######
 library("plyr")
 library("reshape2")
 library("stringr")
@@ -122,7 +121,7 @@ process_session <- function(filename, session_df){
     melt(cbind(filename, session_df), id.vars = 1:2, variable.name = "musician", value.name = "gesture")
 }
 
-df <- ldply(list.files("../data"),
+df <- ldply(list.files("../data", pattern = "-posthoc-gestures\\.csv$"),
             function(x) {
                 filename <- paste("../data/", x, sep = "")
                 process_session(x, read.csv(filename))
[…]

### 2015-08-31 metatone_post_hoc_analysis.py
adding lots of plots

@@ -206,18 +206,18 @@ def main():
 
     # a.apply(lambda x: LIKERT_MAPPING[x])
     LIKERT_MAPPING = {1:1,2:1,3:2,4:2,5:3,6:4,7:4,8:5,9:5}
-    #ratings = performance_surveys_2014["Q1"]
+    #ratings = performance_surveys_2014["Q6"]
     #ratings = ratings.append(performance_surveys_2015_runthrough["Q26"].apply(lambda x: LIKERT_MAPPING[x]))
     #ratings = ratings.append(performance_surveys_2015_study["Q23"].apply(lambda x: LIKERT_MAPPING[x]))
     #flux_ratings_frame = perf_frame['flux']
     #flux_ratings_frame = pd.concat([flux_ratings_frame,ratings],axis =1)
 
-    #ratings = performance_surveys_2014["Q6"]
-    #flux_entropy_ratings = flux_entropy_ratings.dropna()
+    ratings = performance_surveys_2014["Q6"]
 
-    #flux_entropy_ratings = pd.DataFrame({"rating":ratings})
-    #flux_entropy_ratings["flux"] = perf_frame["flux"]
-    #flux_entropy_ratings["entropy"] = perf_frame["entropy"]
+    flux_entropy_ratings = pd.DataFrame({"rating":ratings})
+    flux_entropy_ratings = flux_entropy_ratings.dropna()
+    flux_entropy_ratings["flux"] = perf_frame["flux"]
+    flux_entropy_ratings["entropy"] = perf_frame["entropy"]
     #flux_entropy_ratings.to_csv("flux_entropy_ratings.csv")
 
 def test_regression(df):

### 2015-09-01 flux-entropy-paper/flux-entropy-paper.tex
added todos for the rest of the day
adding lots of plots
updates for the results section

@@
 interaction can be characterised by a set of interaction states $S$
-with $m$ members.
-Each participant's activity can then be represented as a sequence:
+with $m$ members. Each participant's activity can then be represented
+as a sequence:
 \begin{equation}
@@
 two high-level quantities, ``flux'' and ``entropy''. The flux of a
-transition matrix $P$ is given by
+sequence is defined as the ratio of changed transitions (e.g. state $x$
+to state $y$), to self transitions (e.g. state $x$ to state $x$). The
+the flux of a transition matrix $P$ can be calculated as:
 \begin{equation}
-  \mathrm{flux}(P) = \frac{\|P\|_1-\|\mathrm{diag}(P)\|}{\|P\|_1}
+  \mathrm{flux}(P) = 1 - tr(P)
 \end{equation}
-Where $\|P\|_1 = \sum_{i,j}|p_{ij}|$ is the element-wise 1-norm of $P$
-and $\mathrm{diag}(P)$ is the vector of the main diagonal entries of
-$P$.
-
-Flux measures the ratio of transitions that change
-state to self-transitions from a certain state back to itself.
-Intuitively, it is a measure of how frequently a group changes state. Flux
+Where the trace of P, $tr(P)$ is given by the vector 1-norm of the
+diagonal entries of $P$. 
+% Where $\|P\|_1 = \sum_{i,j}|p_{ij}|$ is the element-wise 1-norm of $P$
+% and $\mathrm{diag}(P)$ is the vector of the main diagonal entries of
+% $P$.
+Intuitively, flux is a measure of how frequently a group changes state. Flux
 returns a value in the interval $[0,1]$ and will return 0 when
@@
 
-% While musical interactions such as those in
-% our corpus are often analysed using qualitative methods, it is
-% problematic to
-
 While the interaction sessions recorded in our corpus were performed
@@
 
-% need to add overall contrasts column to do pairwise t-tests.
+% todo need to add overall contrasts column to do pairwise t-tests.
 
@@
 
+Figures \ref{fig:type-section-flux} and \ref{fig:type-section-entropy}
+show that the flux and entropy of compositions are much more stable
+with respect to section than for improvisations. The stability of
+compositions is to be expected since performers are expected to use
+the same gestures in each rehearsal and performance of a composition,
+it may be that different kinds of compositions could result in
+different structural differences. Both measures have a downward trend
+through sections in improvisations with Kruskal-Wallis tests showing a
+significant effect of section on flux ($\chi^2(2)=5.8, p=0.05$). This
+could be explained by performers entering an exploration stage at the
+start of an improvisation, where they change gesture frequently to
+experiment with different ideas. At the end of an improvisation,
+performers may change gesture less frequently while ``winding up'' the
+performance.
+
+The gestural change in improvisation can be further explored by
+dividing the sessions by performance context. Figures
+\ref{fig:context-section-flux}, and \ref{fig:context-section-entropy} show
+this division for improvisations only. In rehearsals and studies the
+distribution of flux and entropy over sections has a similar shape,
+both measures drop in the ending section of studies and after the
+beginning section in rehearsals. In performances however, flux drops
+after the beginning while entropy drops substantially in the ending
+section. This indicates a more constrained performance style in this
+closing section, where performers transition between fewer gestures in
+the final section of performances.
+
+% todo add the plots here.
+
+\begin{figure}
+  \centering
+  \includegraphics[width=\linewidth]{figures/type-section-flux}
+  \caption{The flux of compositions and improvisations divided into
+    beginning, middle, and end. Compositions remain stable throughout,
+    but improvisations show a downward trend.
+    \label{fig:type-section-flux}}
+\end{figure}
+
+\begin{figure}
+  \centering
+  \includegraphics[width=\linewidth]{figures/type-section-entropy}
+  \caption{The entropy of compositions and improvisations divided by
+    beginning, middle, and end. Composition entropy tends to move up
+    at the end, while flux moves down.
+    \label{fig:type-section-entropy}}
+\end{figure}
+
+% todo add the results from the plots.
+
+\begin{figure}
+  \centering
+  \includegraphics[width=\linewidth]{figures/context-section-flux}
+  \caption{The distribution in performances divided by ``beginnings,
+    middles, and ends''.
+    \label{fig:context-section-flux}}
+\end{figure}
+
+\begin{figure}
+  \centering
+  \includegraphics[width=\linewidth]{figures/context-section-entropy}
+  \caption{The distribution of entropy in performances divided into
+    sections and by performance context.
+    \label{fig:context-section-entropy}}
+\end{figure}
+
 \subsection{Performance styles over time}
@@
 
+% todo add anymore results from these plots
+
 \section{Conclusions}
+
+In this paper we have extended the transition matrix methodology
+previously presented at CHI with two simple matrix measures, flux and
+entropy, both of which have been demonstrated to have significant
+discriminatory power with regards to a corpus of 150 collaborative
+touch-screen interactions. Not only do these measures differentiate
+between different types of sessions and the internal structure of
+these sessions, but they have explanatory power that has thrown new
+light on the musical interactions under examination. HCI researchers
+are increasingly concerned with interactions that happen in-the-wild,
+where the interactions of many users may be collected without the
+opportunity to survey or interview users. We suggest that analysing
+interaction-state transitions using matrices and simple measures can
+lead to new insights into HCI interactions outside the lab.
+
+% todo blah de blah
 

### 2015-09-01 R/analysis.R

@@ -148,17 +148,39 @@ tm <- merge(tm, metadata, by = "filename", all.x = TRUE)
 tm <- subset(tm, performance_type!="fail")
 tm <- subset(tm, performance_context %in% c("performance", "rehearsal", "study"))
 
-improvisation.sections <- subset(tm, performance_type = "improvisation")
-                                 
+improvisation.sections <- subset(tm, performance_type == "improvisation")
+composition.sections <- subset(tm, performance_type == "composition")
+
+head(improvisation.sections)
 ## Models
-# ANOVA and t-test
-summary(aov(flux~performance_type*section, improvisation.sections))
-pairwise.t.test(improvisation.sections$flux,improvisation.sections$section,p.adjust.method="bonferroni")
-summary(aov(entropy~performance_type*section, tm))
-pairwise.t.test(tm$entropy,tm$section,p.adjust.method="bonferroni")
+
+# Kruskal Wallis tests on performance type sections.
+kruskal.test(improvisation.sections$flux~improvisation.sections$section)
+kruskal.test(improvisation.sections$entropy~improvisation.sections$section)
+kruskal.test(composition.sections$flux~composition.sections$section)
+kruskal.test(composition.sections$entropy~composition.sections$section)
+
+#TukeyHSD(aov(flux~section, data=improvisation.sections))
+#TukeyHSD(aov(entropy~section, data=improvisation.sections))
+#pairwise.t.test(improvisation.sections$flux,improvisation.sections$section,p.adjust.method="bonferroni")
+                                        #pairwise.t.test(improvisation.sections$entropy,improvisation.sections$section,p.adjust.method="bonferroni")
+
+# Considering just performances:
+performances <- subset(improvisation.sections,performance_context == "performance")
+pairwise.t.test(performances$entropy,performances$section,p.adjust.method="bonferroni")
+summary(aov(entropy~section+Error(section/filename),data=performances))
+summary(performances)
+kruskal.test(performances$entropy~performances$section)
+
+#summary(aov(flux~section + Error(section/filename), data = tm))
+
+#TukeyHSD(aov(flux~section + Error(section/filename), data=improvisation.sections))
[…]

### 2015-09-01 flux-entropy-paper/flux-entropy-paper.tex
cleanup and new faceted plots

@@
 
-\section{A Corpus of Musical Touch-Screen Interactions}
+\section{A Corpus of Touch-Screen Interactions}
 
@@
 
+\begin{figure}
+  \centering
+  \includegraphics[width=\linewidth]{figures/context-flux-entropy-boxplot}
+  \caption{Distributions of flux and entropy values by performance context. Study
+    and rehearsal sessions had the highest flux for improvisations
+    which fell during performances. Compositions had lower flux in
+    rehearsals than performances. Demonstration sessions had much lower entropy
+  indicating data unevenly spread in the transition matrices. This
+  could be because untrained
+  participants did not explore the full gestural potential of the
+  instruments. 
+  \label{fig:flux-entropy-boxplot}}
+\end{figure}
+
+\begin{figure}
+  \centering
+  \includegraphics[width=\linewidth]{figures/flux-entropy-distribution}
+  \caption{Distribution of flux and entropy by performance type and
+    context. Flux and entropy tended to be linearly related in
+    improvisations, particularly in demonstrations and studies, while
+    compositions had a more scattered relationship.
+    \label{fig:flux-entropy-distribution}}
+\end{figure}
+
 The distribution of flux and entropy with respect to session contexts,
-types and instrumentation are shown in Figures \ref{fig:flux-boxplot},
-\ref{fig:entropy-boxplot}, and \ref{fig:flux-entropy-distribution}. A
-three-way ANOVA procedure was performed on the data set to find
-whether the metatdata could significantly predict the entropy and flux
-values. In the following analysis, only the 95 performative sessions
-were considered (i.e. not the demonstration or data collection
-contexts).
+types and instrumentation are shown in Figures
+\ref{fig:flux-entropy-boxplot} and
+\ref{fig:flux-entropy-distribution}. A three-way ANOVA procedure was
+performed on the data set to find whether the metatdata could
+significantly predict the entropy and flux values. In the following
+analysis, only the 95 performative sessions were considered (i.e. not
+the demonstration or data collection contexts).
 
@@
 
-\begin{figure}
-  \centering
-  \includegraphics[width=\linewidth]{figures/flux-boxplot}
-  \caption{Distributions of flux values by performance context. Study
-    and rehearsal sessions had the highest flux for improvisations
-    which fell during performances. Compositions had lower flux in
-    rehearsals than performances.
-    \label{fig:flux-boxplot}}
-\end{figure}
-
-\begin{figure}
-  \centering
-  \includegraphics[width=\linewidth]{figures/entropy-boxplot}
-  \caption{Distribution of entropy by
-  performance context. Demonstration sessions had much lower entropy
-  indicating data unevenly spread in the transition matrices. This
-  could be because untrained
-  participants did not explore the full gestural potential of the
-  instruments. 
-  \label{fig:entropy-boxplot} }
-\end{figure}
-
-\begin{figure}
-  \centering
-  \includegraphics[width=\linewidth]{figures/flux-entropy-distribution}
-  \caption{Distribution of flux and entropy by performance type and
-    context. Flux and entropy tended to be linearly related in
-    improvisations, particularly in demonstrations and studies, while
-    compositions had a more scattered relationship.
-    \label{fig:flux-entropy-distribution}}
-\end{figure}
-
 \subsection{Breaking performances into sections}
@@
 
-Figures \ref{fig:type-section-flux} and \ref{fig:type-section-entropy}
-show that the flux and entropy of compositions are much more stable
-with respect to section than for improvisations. The stability of
-compositions is to be expected since performers are expected to use
-the same gestures in each rehearsal and performance of a composition,
-it may be that different kinds of compositions could result in
-different structural differences. Both measures have a downward trend
-through sections in improvisations with Kruskal-Wallis tests showing a
-significant effect of section on flux ($\chi^2(2)=5.8, p=0.05$). This
-could be explained by performers entering an exploration stage at the
-start of an improvisation, where they change gesture frequently to
-experiment with different ideas. At the end of an improvisation,
-performers may change gesture less frequently while ``winding up'' the
-performance.
+Figures \ref{fig:type-section-flux-entropy} shows that the flux and
+entropy of compositions are much more stable with respect to section
+than for improvisations. The stability of compositions is to be
+expected since performers are expected to use the same gestures in
+each rehearsal and performance of a composition, it may be that
+different kinds of compositions could result in different structural
+differences. Both measures have a downward trend through sections in
+improvisations with Kruskal-Wallis tests showing a significant effect
+of section on flux ($\chi^2(2)=5.8, p=0.05$). This could be explained
+by performers entering an exploration stage at the start of an
+improvisation, where they change gesture frequently to experiment with
+different ideas. At the end of an improvisation, performers may change
+gesture less frequently while ``winding up'' the performance.
 
 The gestural change in improvisation can be further explored by
-dividing the sessions by performance context. Figures
-\ref{fig:context-section-flux}, and \ref{fig:context-section-entropy} show
-this division for improvisations only. In rehearsals and studies the
-distribution of flux and entropy over sections has a similar shape,
-both measures drop in the ending section of studies and after the
-beginning section in rehearsals. In performances however, flux drops
-after the beginning while entropy drops substantially in the ending
-section. This indicates a more constrained performance style in this
-closing section, where performers transition between fewer gestures in
-the final section of performances.
+dividing the sessions by performance context. Figure
+\ref{fig:context-section-flux-entropy} shows this division for
+improvisations only. In rehearsals and studies the distribution of
+flux and entropy over sections has a similar shape, both measures drop
+in the ending section of studies and after the beginning section in
+rehearsals. In performances however, flux drops after the beginning
+while entropy drops substantially in the ending section. This
+indicates a more constrained performance style in this closing
+section, where performers transition between fewer gestures in the
+final section of performances.
 
@@
   \includegraphics[width=\linewidth]{figures/type-section-flux}
+\includegraphics[width=\linewidth]{figures/type-section-entropy}
   \caption{The flux of compositions and improvisations divided into
-    beginning, middle, and end. Compositions remain stable throughout,
-    but improvisations show a downward trend.
-    \label{fig:type-section-flux}}
+    beginning, middle, and end. Composition flux remains stable throughout,
+    but improvisations show a downward trend and composition entropy tends to move up
+    at the end, while flux moves down.
+    \label{fig:type-section-flux-entropy}}
 \end{figure}
@@
   \centering
-  \includegraphics[width=\linewidth]{figures/type-section-entropy}
-  \caption{The entropy of compositions and improvisations divided by
-    beginning, middle, and end. Composition entropy tends to move up
-    at the end, while flux moves down.
-    \label{fig:type-section-entropy}}
-\end{figure}
-
-% todo add the results from the plots.
-
-\begin{figure}
-  \centering
   \includegraphics[width=\linewidth]{figures/context-section-flux}
-  \caption{The distribution in performances divided by ``beginnings,
+  \includegraphics[width=\linewidth]{figures/context-section-entropy}
+  \caption{The distribution of flux and entropy in performances divided by ``beginnings,
     middles, and ends''.
-    \label{fig:context-section-flux}}
-\end{figure}
-
-\begin{figure}
-  \centering
-  \includegraphics[width=\linewidth]{figures/context-section-entropy}
-  \caption{The distribution of entropy in performances divided into
-    sections and by performance context.
-    \label{fig:context-section-entropy}}
+    \label{fig:context-section-flux-entropy}}
 \end{figure}
@@
 contexts, they are all part of a program of artistic and HCI research
-that has evolved over time. Figures \ref{fig:flux-through-time} and
-\ref{fig:entropy-through-time} show the values of flux and entropy for
-each session in the corpus. Over the course of this research program,
-different ensembles and different activities have been emphasised,
-some of these differences can be seen as clusters of performances with
-similar flux and entropy in the two plots. For instance, the earliest
-series of rehearsals and performances had flux between 0.05 and 0.15,
-while the period of activity after July 2014 had flux between 0.1 and
-0.25. This change corresponded with an intensive series of study
-sessions and rehearsals where the participants developed a more
-fluctuating style of performance. The entropy between these two
-periods, however, remains at a similiar level, showing the the space
-of available gestures that the performers explored was more constant.
+that has evolved over time. Figure \ref{fig:flux-entropy-through-time}
+shows the values of flux and entropy for each session in the corpus.
+Over the course of this research program, different ensembles and
+different activities have been emphasised, some of these differences
+can be seen as clusters of performances with similar flux and entropy
+in the two plots. For instance, the earliest series of rehearsals and
+performances had flux between 0.05 and 0.15, while the period of
+activity after July 2014 had flux between 0.1 and 0.25. This change
+corresponded with an intensive series of study sessions and rehearsals
+where the participants developed a more fluctuating style of
+performance. The entropy between these two periods, however, remains
+at a similiar level, showing the the space of available gestures that
+the performers explored was more constant.
 
@@
   \centering
-  \includegraphics[width=\linewidth]{figures/flux-through-time}
-  \caption{Distribution of flux values of sessions through time.
+  \includegraphics[width=\linewidth]{figures/flux-entropy-through-time}
+  \caption{Distribution of flux and entropy values of sessions through time.
     Different performance styles are visible as clusters of similar
     flux regardless of performance context.
-    \label{fig:flux-through-time}}
+    \label{fig:flux-entropy-through-time}}
 \end{figure}
-
-\begin{figure}
-  \centering
-  \includegraphics[width=\linewidth]{figures/entropy-through-time}
-  \caption{Distribution of entropy values of sessions through time.
-    \label{fig:entropy-through-time}}
-\end{figure}
-
-% todo add anymore results from these plots
 

### 2015-09-01 R/analysis.R

@@ -1,6 +1,9 @@
 library(ggplot2)
 library(MASS)
 library(grid)
+library(gridExtra)
+library(reshape2)
+
 chifig.3colours <- c("#e41a1c", "#377eb8", "#4daf4a")
 chifig.2colours <- c("#984ea3", "#ff7f00")
 chifig.5colours <- c("#e41a1c", "#377eb8", "#4daf4a", "#984ea3", "#ff7f00")
@@ -26,6 +29,12 @@ message(paste(sum(valid.sessions$number_performers),min(valid.sessions$number_pe
                                         #
 message(paste(signif(sum(valid.sessions$flux),4),signif(min(valid.sessions$flux),4),signif(median(valid.sessions$flux),4),signif(max(valid.sessions$flux),4),sep = " & "))
 
+# New long data frame for initial plots.
+valid.sessions.long <- read.csv("../metatone-performance-data-normalmatrices.csv")
+valid.sessions.long <- melt(valid.sessions.long,id.vars=c("time","performance_context","performance_type","instruments"),variable.name = "measure",measure.vars = c("flux","entropy"))
+valid.sessions.long$time <- strptime(as.character(valid.sessions.long$time), "%Y-%m-%d %H:%M:%S.%OS")
+valid.sessions.long <-  subset(valid.sessions.long, performance_type %in% c("composition","improvisation"))
+summary(valid.sessions.long)
 
 ## Initial Plots
 # cut out data_collection
@@ -36,11 +45,14 @@ ggplot(valid.sessions, aes(performance_context)) + geom_bar(aes(fill = performan
 quartz.save("../flux-entropy-paper/figures/sessions-count.pdf", type = "pdf")
 #ggplot(perf.contexts, aes(performance_context)) + geom_bar(aes(fill = performance_type))
 # Flux boxplot
-ggplot(valid.sessions, aes(performance_context, flux)) + geom_boxplot(aes(fill = performance_type)) + theme(plot.margin=unit(rep(0,4), "cm"), legend.position = "top", legend.box = "horizontal") + scale_fill_manual(values=chifig.2colours) + scale_x_discrete("performance context")
-quartz.save("../flux-entropy-paper/figures/flux-boxplot.pdf", type = "pdf")
[…]

### 2015-09-01 R/analysis.R
shorter count plot

@@ -41,7 +41,7 @@ summary(valid.sessions.long)
 #valid.sessions <- (subset(valid.sessions, performance_context != "data_collection"))
 ## plots
 # Count of Session Data
-ggplot(valid.sessions, aes(performance_context)) + geom_bar(aes(fill = performance_type)) + theme(plot.margin=unit(rep(0,4), "cm"), legend.position = "top", legend.box = "horizontal") + scale_fill_manual(values=chifig.2colours)
+ggplot(valid.sessions, aes(performance_context)) + geom_bar(aes(fill = performance_type)) + theme(plot.margin=unit(rep(0,4), "cm"), legend.position = "right", legend.box = "vertical") + scale_fill_manual(values=chifig.2colours)
 quartz.save("../flux-entropy-paper/figures/sessions-count.pdf", type = "pdf")
 #ggplot(perf.contexts, aes(performance_context)) + geom_bar(aes(fill = performance_type))
 # Flux boxplot

### 2015-09-01 flux-entropy-paper/flux-entropy-paper.tex
-
fixed up a bit of the flux def

@@
 \begin{equation}
-  p_{ij} = \frac{N_{ij}}{\sum_j N_{ij}}
+  p_{ij} = \frac{N_{ij}}{\sum_i \sum_j N_{ij}}
 \end{equation}
@@
 be calculated by averaging the transition matrix for each performer.
+The matrix is normalised so that the sum of all elements will be one
+(rather than the sum of each row being one, as for a right-stochastic matrix).
 
@@
 two high-level quantities, ``flux'' and ``entropy''. The flux of a
-sequence is defined as the ratio of changed transitions (e.g. state $x$
-to state $y$), to self transitions (e.g. state $x$ to state $x$). The
-the flux of a transition matrix $P$ can be calculated as:
+sequence is defined as the ratio of changed transitions (e.g.
+$A \rightarrow B$), to self transitions (e.g. $A \rightarrow A $). The
+the flux of a transition matrix $P$ can easily be calculated:
 \begin{equation}
-  \mathrm{flux}(P) = 1 - tr(P)
+  \mathrm{flux}(P) = 1 - \mathrm{tr}(P)
 \end{equation}
-Where the trace of P, $tr(P)$ is given by the vector 1-norm of the
-diagonal entries of $P$. 
+Where the trace of P, $tr(P)$ is given by the sum of the diagonal
+entries of $P$.
 % Where $\|P\|_1 = \sum_{i,j}|p_{ij}|$ is the element-wise 1-norm of $P$
@@
 
-A related (but different) high-level derived quantity for the
-transition matrix is its entropy, defined in the
-information-theoretic~\cite{Shannon:1948rt} sense:
+Another useful measure that can be used on a transition matrix is its
+entropy, defined in the information-theoretic~\cite{Shannon:1948rt}
+sense:
 \begin{equation}
@@
 In the following sections, we will use transition matrices and the
-flux measure to analyse a corpus of data collected as part of our research
-into collaborative interactions with musical touch-screen tablet
-interfaces.
+flux measure to analyse a corpus of data collected as part of our
+research into collaborative interactions with musical touch-screen
+tablet interfaces.
 
@@
 
-% todo need to add overall contrasts column to do pairwise t-tests.
+\begin{itemize}
+\item todo: need to add overall contrasts column to do pairwise t-tests.
+\end{itemize}
 

### 2015-09-10 flux-entropy-paper/flux-entropy-paper.tex
ben edits, add some questions/comments to beginning

@@
   evaluated with user ratings and interviews. However, such data can
-  be impractical to collect in-the-wild, such as during musical
-  performances. In this paper we consider measures of gestural
-  transitions during collaborative interactions with musical touch
-  screen interfaces. The performers' interactions are characterised as
-  sequences of discrete gestures and summarised in a transition matrix
-  for each performance. Using a corpus of more than 140 performances,
-  rehearsals, and demonstrations, we find that two matrix measures,
-  flux and entropy, vary significantly with different kinds of
-  in-the-wild interactions. Further, using performers' ratings
-  collected during a lab-study we find that these measures have a
-  significant relationship with the perceived effect of the interface
-  on performance. Thus transition analysis, flux and entropy, can be
-  used as a proxy for performer ratings during in-the-wild interaction.
+  be impractical to collect in-the-wild, for example during musical
+  performances. In this paper we show how a transition matrix based
+  post-hoc analysis of logged interaction from a long program of
+  experiments in group musical creativity can shed light on the nature
+  of the group interaction in-the-wild. Using a corpus of more than
+  140 performances, rehearsals, and demonstrations over more than two
+  years, we find that two matrix measures, flux and entropy, can be
+  used as a proxy for performer ratings during in-the-wild
+  interaction.
 \end{abstract}
@@
 %   classifiers. This section is required.}{}{}
+
+\section{General comments}
+
+\begin{itemize}
+\item we should be clear (and careful) where we're talking about
+  groups TMs, and where we're talking about individual performers
+\item what's the terminology we're using for a ``performance''? Is it
+  ``performance'', or is that too music-specific? ``Session''?
+  Alternatives are ``activity'', or ``interaction'', but they're so
+  general\ldots This decision has implications for our other
+  terminology: participant vs performer vs musician, etc.
+\item Title: maybe ``Traces through time: making sense of longitudinal
+  open-ended interaction data'' or some such? I think that's our big
+  contribution: it takes the in-the-wild, practitioner-driven research
+  program stuff and tries to make some quantitative sense out of it
+  post-hoc.
+\end{itemize}
 
@@
 
-Finally, we report a quantitative analysis of the transition matrices
-from our experimental protocols, and describe the quantities
-\emph{flux} and \emph{entropy} which can be used to characterise the
-gestural activity of the musicians during a performance. We found a
-significant effect of the app on these measures, allowing us to
-observe quantitatively the performers' dissatisfaction with a
-particular interface. We also investigate the significant relationship
-between both of these measures and the survey responses, which
-suggests that performers rate interaction with our agent and apps more
-highly when engaged in adventurous, exploratory performance.
+\section{Introduction}
 
-\section{Related Work}
-
-\begin{itemize}
-\item Using Markov chains and transition matrices for usability
-  modelling~\cite{Thimbleby:2001kq,Thimbleby:2004fj}
-\item coding livecoding~\cite{Swift:2014tya}
-\end{itemize}
+Ideas: the in-the-wild nature, and the different contexts - both the
+fact that they are different, and the fact that their differences make
+collecting certain data/responses easier or harder
 
@@
 
-It has been recognised that collaborative interactions such as
-improvised musical performances can be modelled as a sequence of
-states. In the following section we present a method for analysing the
-transitions between adjacent states in such a sequence and derive two
-high-level quantities that we use to compare musical performances
-under different circumstances. Suppose that an collaborative
-interaction can be characterised by a set of interaction states $S$
-with $m$ members. Each participant's activity can then be represented
-as a sequence:
+Open-ended collaborative interactions such as improvised musical
+performances can be modelled as a sequence of
+% could also use Thimbleby:2004fj
+states\cite{Thimbleby:2001kq,Swift:2014tya}. In this section we use
+transition matrix (TM) statistics to analyse the transitions between
+adjacent states in such a sequence. We then present two high-level
+derived quantities---\emph{flux} and \emph{entropy}---which can be
+used to compare musical performances under different circumstances.
+
+Suppose that each participant's activity in a collaborative
+interaction can at any given time be classified as belonging to one of
+$m$ distinct \emph{states}. Over the course of the activity, each
+participant $X$ generates a length $N$ sequence of these interaction
+states
 \begin{equation}
@@
 \end{equation}
-Where each $X_i$ is a member of $G$. The transitions between different
-interaction states can be summarised in a transition matrix as if the
-sequence $X_n$ were a first-order Markov chain. Let $N_{ij}$ be the
-number of times that state $j$ follows state $i$. Each participant's
-transition matrix, $P$, will then be an $m \times m$ matrix with
-entries calculated thus
+The transitions between these interaction states can be summarised in
+a transition matrix (TM) as if the sequence $X_n$ were a first-order Markov
+chain. Let $N_{ij}$ be the number of times that state $j$ follows
+state $i$. Each participant's TM $P$, will then be an
+$m \times m$ matrix with entries
 \begin{equation}
@@
 the $j^{th}$ column. The transition activity of the whole ensemble can
-be calculated by averaging the transition matrix for each performer.
-The matrix is normalised so that the sum of all elements will be one
-(rather than the sum of each row being one, as for a right-stochastic matrix).
+be calculated by averaging the TM for each performer.
+The matrix can be normalised so that the sum of all elements is equal
+to one.
+% (rather than the sum of each row being one, as for a right-stochastic matrix)
 
-To compare the transition matrices of different interactions we use
-two high-level quantities, ``flux'' and ``entropy''. The flux of a
-sequence is defined as the ratio of changed transitions (e.g.
-$A \rightarrow B$), to self transitions (e.g. $A \rightarrow A $). The
-the flux of a transition matrix $P$ can easily be calculated:
+If there is an (automated) classification process for determining each
+participant's state sequence during an activity, then it is a simple
+process to generate a TM for each participant and for the group as a
+whole. The challenge is to understand what the TM reveals about the
+group interaction, and to allow for comparisons between sessions and
+groups.
+
+To compare the TMs of different interactions we use two high-level
+quantities, ``flux'' and ``entropy''. The flux of a sequence is
+defined as the ratio of change transitions (e.g. $A \rightarrow B$
+where $A \neq B$), to self transitions (e.g. $A \rightarrow A $). The
+flux of a TM $P$ is given by
 \begin{equation}
@@
 \end{equation}
-Where the trace of P, $tr(P)$ is given by the sum of the diagonal
-entries of $P$.
+where the trace $\mathrm{tr}(P)$ is the sum of the diagonal entries of
+$P$.
 % Where $\|P\|_1 = \sum_{i,j}|p_{ij}|$ is the element-wise 1-norm of $P$
@@
 % $P$.
-Intuitively, flux is a measure of how frequently a group changes state. Flux
-returns a value in the interval $[0,1]$ and will return 0 when
-participants never change state, and 1 when participants never stay on
-the same state for two measurements in a row.
+Intuitively, flux is a measure of how frequently the participant/group
+changes state. Flux returns a value in the interval $[0,1]$ and will
+return 0 when participants never change state, and 1 when participants
+never stay in the same state for two measurements in a row.
 
@@
 $ B \rightarrow A$) even though the participant has only used a small
-subset of the state space. The entropy of this interaction,
-on the other hand, will be low. Entropy, therefore, is a measure of
-how broad the participant's exploration of the state space is in a given
+subset of the state space. The entropy of this interaction, on the
+other hand, will be low. Entropy, therefore, is a measure of how broad
+the participant's exploration of the state space is in a given
 interaction.
 
-\section{A Corpus of Touch-Screen Interactions}
+\section{Making sense of two year's worth of touch-screen
+  interactions}
 
@@
 
-In the following sections, we will use transition matrices and the
-flux measure to analyse a corpus of data collected as part of our
-research into collaborative interactions with musical touch-screen
-tablet interfaces.
+% TODO maybe a ref to the CHI paper here?
+
+% TODO I think this ``scene-setting'' stuff could actually go in the
+% first section of the paper, with the TM stuff following.  But I'm
+% open to push-back on that.
 
 Over a period of more than two years (April 2013 - June 2015), we have
-collected data from $150$ sessions where various participants used
-musical touch-screen interfaces on Apple iPad devices. These
-interfaces all shared a common mode of interaction where the majority
-of the screen could be used for free form touch interactions; tapping
-produced short sounds, while swiping or swirling produced long sounds
-with volume proportional to the velocity of the moving touch point.
-All touch interactions from each participant in a session was recorded
-by sending data to a central server application. This data was stored
-in CSV files recording the date and time, X, and Y locations,
-velocity, and a unique device identifier for each touch event detected
-by the iOS operating system.
+collected data from $150$ group musicmaking sessions where
+participants used musical touch-screen interfaces on Apple iPad
+devices. These interfaces all shared a common mode of interaction
+where the majority of the screen could be used for free form touch
+interactions; tapping produced short sounds, while swiping or swirling
+produced long sounds with volume proportional to the velocity of the
+moving touch point. Through all of these performances every touch
+interaction---every touch down, dragged and up---was timestamped and
+logged to disk for later analysis
+
+% TODO say something about how it was both your research project and
+% your creative practice. That's why it encompassed all these
+% different contexts.
 
@@
 Figure \ref{fig:count-data}.
+
+% TODO again, I think this bit would be beefed-up if we included the
+% fact that it's already been published.  And it also means we don't
+% have to go into as much detail about how we did it - the main thing
+% is that we've got a sensible (and peer-reviewed!) way of taking the
+% touch interaction data and generating gesture sequences (and
+% therefore TMs)
 
@@
 performed on the data set to find whether the metatdata could
-significantly predict the entropy and flux values. In the following
-analysis, only the 95 performative sessions were considered (i.e. not
-the demonstration or data collection contexts).
+significantly predict the entropy and flux values.
+% TODO we haven't explained what ``the 95 performative sessions'' are,
+% have we?
+In the following analysis, only the 95 performative sessions were
+considered (i.e.~not the demonstration or data collection contexts).
 
@@
 These tests show that the different interaction styles that might be
-present in performance contexts, types and
-instrumentations are reflected in the values of flux and entropy of
-their transition matrices.
+present in performance contexts, types and instrumentations are
+reflected in the values of flux and entropy of their transition
+matrices.
 
 \begin{itemize}
-\item todo: need to add overall contrasts column to do pairwise t-tests.
+\item TODO: need to add overall contrasts column to do pairwise t-tests.
 \end{itemize}
@@
 performances? A simple model for understanding temporal art-forms is
-the ``beginning, middle, end'' structure. To investigate how the
-structure fits session in our corpus, we can divide the
-gestural-sequences of each session into thirds and calculate the
-transition matrices given by each section.
+the ``beginning, middle, end'' structure. To investigate how this
+structure fits our corpus, we can divide the gestural-sequences of
+each session into thirds and calculate the transition matrices given
+by each section.
 
-Figures \ref{fig:type-section-flux-entropy} shows that the flux and
+Figure~\ref{fig:type-section-flux-entropy} shows that the flux and
 entropy of compositions are much more stable with respect to section
@@
 corresponded with an intensive series of study sessions and rehearsals
-where the participants developed a more fluctuating style of
+where the participants developed a more ``fluctuating'' style of
 performance. The entropy between these two periods, however, remains
-at a similiar level, showing the the space of available gestures that
-the performers explored was more constant.
+at a similar level, showing the the space of available gestures that
+the performers explored was relatively stable.
 
@@
 
-In this paper we have extended the transition matrix methodology
-previously presented at CHI with two simple matrix measures, flux and
-entropy, both of which have been demonstrated to have significant
-discriminatory power with regards to a corpus of 150 collaborative
-touch-screen interactions. Not only do these measures differentiate
-between different types of sessions and the internal structure of
-these sessions, but they have explanatory power that has thrown new
-light on the musical interactions under examination. HCI researchers
-are increasingly concerned with interactions that happen in-the-wild,
+In this paper we have extended our previous work on transition
+matrices with two simple matrix measures, flux and entropy, both of
+which have been demonstrated to have significant discriminatory power
+with regards to a corpus of 150 collaborative touch-screen
+interactions. Not only do these measures differentiate between
+different types of sessions and the internal structure of these
+sessions, but they have explanatory power that has thrown new light on
+the musical interactions under examination. HCI researchers are
+increasingly concerned with interactions that happen in-the-wild,
 where the interactions of many users may be collected without the

### 2015-09-14 R/analysis.R
added shorter session count figure

@@ -41,8 +41,8 @@ summary(valid.sessions.long)
 #valid.sessions <- (subset(valid.sessions, performance_context != "data_collection"))
 ## plots
 # Count of Session Data
-ggplot(valid.sessions, aes(performance_context)) + geom_bar(aes(fill = performance_type)) + theme(plot.margin=unit(rep(0,4), "cm"), legend.position = "right", legend.box = "vertical") + scale_fill_manual(values=chifig.2colours)
-quartz.save("../flux-entropy-paper/figures/sessions-count.pdf", type = "pdf")
+ggplot(valid.sessions, aes(performance_context)) + geom_bar(aes(fill = performance_type)) + theme(plot.margin=unit(rep(0,4), "cm"), legend.position = "top", legend.box = "horizontal") + scale_fill_manual(values=chifig.2colours) + scale_x_discrete("performance context") + scale_y_continuous("no. of sessions")
+ggsave("../flux-entropy-paper/figures/sessions-count.pdf", width=5.4, height = 2.4)
 #ggplot(perf.contexts, aes(performance_context)) + geom_bar(aes(fill = performance_type))
 # Flux boxplot
 #ggplot(valid.sessions, aes(performance_context, flux)) + geom_boxplot(aes(fill = performance_type)) + theme(plot.margin=unit(rep(0,4), "cm"), legend.position = "top", legend.box = "horizontal") + scale_fill_manual(values=chifig.2colours) + scale_x_discrete("performance context")

### 2015-09-14 metatone_post_hoc_analysis.py
pushed figures together a bit

@@ -40,8 +40,8 @@ class MetatoneTouchLog:
             self.gestures = generate_posthoc_gesture_score.generate_gesture_frame(self.touches)
             self.gestures.to_csv(gestures_path)
         self.ensemble_transition_matrix = transitions.calculate_full_group_transition_matrix(self.gestures)
-        self.ensemble_transition_matrix = transitions.transition_matrix_to_stochastic_matrix(self.ensemble_transition_matrix)
-        #self.ensemble_transition_matrix = transitions.transition_matrix_to_normal_transition_matrix(self.ensemble_transition_matrix)
+        #self.ensemble_transition_matrix = transitions.transition_matrix_to_stochastic_matrix(self.ensemble_transition_matrix)
+        self.ensemble_transition_matrix = transitions.transition_matrix_to_normal_transition_matrix(self.ensemble_transition_matrix)
         self.longest_break = self.find_long_breaks()
 
     def first_touch_timestamp(self):

### 2015-09-14 flux-entropy-paper/flux-entropy-paper.tex
cleaned up figures a bit

@@
 \usepackage{todonotes}
+
+\usepackage{subcaption}
 
@@
 
-\begin{figure}
-  \centering
-  \includegraphics[width=\linewidth]{figures/flux-entropy-distribution}
-  \caption{Distribution of flux and entropy by performance type and
-    context. Flux and entropy tended to be linearly related in
-    improvisations, particularly in demonstrations and studies, while
-    compositions had a more scattered relationship.
-    \label{fig:flux-entropy-distribution}}
-\end{figure}
+% \begin{figure}
+%   \centering
+%   \includegraphics[width=\linewidth]{figures/flux-entropy-distribution}
+%   \caption{Distribution of flux and entropy by performance type and
+%     context. Flux and entropy tended to be linearly related in
+%     improvisations, particularly in demonstrations and studies, while
+%     compositions had a more scattered relationship.
+%     \label{fig:flux-entropy-distribution}}
+% \end{figure}
 
 The distribution of flux and entropy with respect to session contexts,
-types and instrumentation are shown in Figures
-\ref{fig:flux-entropy-boxplot} and
-\ref{fig:flux-entropy-distribution}. A three-way ANOVA procedure was
+types and instrumentation are shown in Figure
+\ref{fig:flux-entropy-boxplot}. A three-way ANOVA procedure was
 performed on the data set to find whether the metatdata could
@@
 \end{itemize}
+
+\begin{figure*}[t!]
+    \centering
+    \begin{subfigure}[t]{0.4\linewidth}
+        \centering
+          \includegraphics[width=\linewidth]{figures/type-section-flux}
+          \includegraphics[width=\linewidth]{figures/type-section-entropy}
+          \caption{Performances divided by performance type. \label{fig:type-section-flux-entropy}}
+    \end{subfigure}%
+    ~ 
+    \begin{subfigure}[t]{0.4\linewidth}
+        \centering
+          \includegraphics[width=\linewidth]{figures/context-section-flux}
+          \includegraphics[width=\linewidth]{figures/context-section-entropy}
+          \caption{Sessions divided by performance context. \label{fig:context-section-flux-entropy}}
+    \end{subfigure}
+    \caption{The flux and entropy of different interaction sessions
+      divided into beginning, middle, and end.}
+\end{figure*}
 
@@
 
-\begin{figure}
-  \centering
-  \includegraphics[width=\linewidth]{figures/type-section-flux}
-\includegraphics[width=\linewidth]{figures/type-section-entropy}
-  \caption{The flux of compositions and improvisations divided into
-    beginning, middle, and end. Composition flux remains stable throughout,
-    but improvisations show a downward trend and composition entropy tends to move up
-    at the end, while flux moves down.
-    \label{fig:type-section-flux-entropy}}
-\end{figure}
+% \begin{figure}
+%   \centering
+%   \includegraphics[width=\linewidth]{figures/type-section-flux}
+%   \includegraphics[width=\linewidth]{figures/type-section-entropy}
+%   \caption{The flux of compositions and improvisations divided into
+%     beginning, middle, and end. Composition flux remains stable throughout,
+%     but improvisations show a downward trend and composition entropy tends to move up
+%     at the end, while flux moves down.
+%     \label{fig:type-section-flux-entropy}}
+% \end{figure}
 
-\begin{figure}
-  \centering
-  \includegraphics[width=\linewidth]{figures/context-section-flux}
-  \includegraphics[width=\linewidth]{figures/context-section-entropy}
-  \caption{The distribution of flux and entropy in performances divided by ``beginnings,
-    middles, and ends''.
-    \label{fig:context-section-flux-entropy}}
-\end{figure}
+% \begin{figure}
+%   \centering
+%   \includegraphics[width=\linewidth]{figures/context-section-flux}
+%   \includegraphics[width=\linewidth]{figures/context-section-entropy}
+%   \caption{The distribution of flux and entropy in performances divided by ``beginnings,
+%     middles, and ends''.
+%     \label{fig:context-section-flux-entropy}}
+% \end{figure}
 

### 2015-09-14 flux-entropy-paper/flux-entropy-paper.tex
two faceted plots for beginning middle end

@@
 to one.
-% (rather than the sum of each row being one, as for a right-stochastic matrix)
 
@@
 where the trace $\mathrm{tr}(P)$ is the sum of the diagonal entries of
-$P$.
-% Where $\|P\|_1 = \sum_{i,j}|p_{ij}|$ is the element-wise 1-norm of $P$
-% and $\mathrm{diag}(P)$ is the vector of the main diagonal entries of
-% $P$.
-Intuitively, flux is a measure of how frequently the participant/group
+$P$. Intuitively, flux is a measure of how frequently the participant/group
 changes state. Flux returns a value in the interval $[0,1]$ and will
@@
 
-\section{Making sense of two year's worth of touch-screen
-  interactions}
+% \section{Making sense of two year's worth of touch-screen
+%   interactions}
+\section{A corpus of touch-screen interaction}
 
@@
 Over a period of more than two years (April 2013 - June 2015), we have
-collected data from $150$ group musicmaking sessions where
+collected data from $150$ group music-making sessions where
 participants used musical touch-screen interfaces on Apple iPad
@@
 interaction---every touch down, dragged and up---was timestamped and
-logged to disk for later analysis
+logged to disk for later analysis.
 
-% TODO say something about how it was both your research project and
-% your creative practice. That's why it encompassed all these
-% different contexts.
-
-Each session took place in a number of contexts including rehearsals,
+These sessions were part of a long-term research project into music
+touch-screen interaction, as well as an ongoing creative practice. As
+a result, the sessions had a variety of contexts including rehearsals,
 performances, lab study sessions, classes, and casual demonstrations,
@@
 second provide a chain of gesture-states for each participant in each
-session. These chains can be used to create $9 /times 9$
+session. These chains can be used to create $9 \times 9$
 gesture-transition matrices for individual performers, and the whole
@@
 In the following analysis, only the 95 performative sessions were
-considered (i.e.~not the demonstration or data collection contexts).
+considered, excluding the demonstration and data-collection contexts.
 
@@
     \centering
-    \begin{subfigure}[t]{0.4\linewidth}
+    \begin{subfigure}[t]{0.5\linewidth}
         \centering
-          \includegraphics[width=\linewidth]{figures/type-section-flux}
-          \includegraphics[width=\linewidth]{figures/type-section-entropy}
+          \includegraphics[width=\linewidth]{figures/type-section-flux-entropy}
           \caption{Performances divided by performance type. \label{fig:type-section-flux-entropy}}
@@
     ~ 
-    \begin{subfigure}[t]{0.4\linewidth}
+    \begin{subfigure}[t]{0.5\linewidth}
         \centering
-          \includegraphics[width=\linewidth]{figures/context-section-flux}
-          \includegraphics[width=\linewidth]{figures/context-section-entropy}
+          \includegraphics[width=\linewidth]{figures/context-section-flux-entropy}
           \caption{Sessions divided by performance context. \label{fig:context-section-flux-entropy}}

### 2015-09-14 R/analysis.R

@@ -206,6 +206,8 @@ ggsave("../flux-entropy-paper/figures/context-section-flux-lm.pdf")
 # Jitter Plots Section x Entropy and context
 ggplot(improvisation.sections,aes(section,entropy))  + facet_wrap(~performance_context) + geom_jitter(alpha=.5,size=3, position = position_jitter(w = 0.1, h = 0), aes(colour=section)) + scale_colour_manual(values=chifig.3colours) + stat_smooth(aes(group=1),method="lm",size=2) + theme(plot.margin=unit(rep(0,4), "cm"), legend.position = "none", legend.box = "horizontal")
 ggsave("../flux-entropy-paper/figures/context-section-entropy-lm.pdf")
+
+
 # Boxplot of Flux by section and context
 ggplot(improvisation.sections, aes(section, flux)) + geom_boxplot(aes(fill=section)) + facet_wrap(~performance_context) + scale_fill_manual(values=chifig.3colours) + theme(plot.margin=unit(rep(0,4), "cm"), legend.position = "none", legend.box = "horizontal")
 ggsave("../flux-entropy-paper/figures/context-section-flux.pdf")
@@ -219,3 +221,11 @@ ggsave("../flux-entropy-paper/figures/type-section-flux.pdf")
 ggplot(tm, aes(section, entropy)) + geom_boxplot(aes(fill=section)) + facet_wrap(~performance_type) + scale_fill_manual(values=chifig.3colours) + theme(plot.margin=unit(rep(0,4), "cm"), legend.position = "none", legend.box = "horizontal")
 ggsave("../flux-entropy-paper/figures/type-section-entropy.pdf")
 
+## Boxplots for flux and entropy by section.
+## faceted by section and measure (flux and entropy)
+improvisation.sections.long <- melt(improvisation.sections,id.vars=c("section","performance_context","performance_type","instruments"),variable.name = "measure",measure.vars = c("flux","entropy"))
+ggplot(improvisation.sections.long, aes(section, value)) + geom_boxplot(aes(fill=section)) + facet_grid(measure~performance_context, scales="free_y") + scale_fill_manual(values=chifig.3colours) + theme(plot.margin=unit(rep(0,4), "cm"), legend.position = "none", legend.box = "horizontal") + scale_x_discrete("") + scale_y_continuous("measure value")
+ggsave("../flux-entropy-paper/figures/context-section-flux-entropy.pdf", width=6, height = 3.6)
[…]

### 2015-09-14 R/analysis.R
milhouse, lower those eyebrows
and the other one.

@@ -52,7 +52,7 @@ ggsave("../flux-entropy-paper/figures/sessions-count.pdf", width=5.4, height = 2
 #quartz.save("../flux-entropy-paper/figures/entropy-boxplot.pdf", type = "pdf")
 # Both Flux and Entropy Boxplot
 ggplot(valid.sessions.long, aes(performance_context, value)) + geom_boxplot(aes(fill = performance_type)) + theme(plot.margin=unit(rep(0,4), "cm"), legend.position = "top", legend.box = "horizontal") + scale_fill_manual(values=chifig.2colours) + scale_x_discrete("performance context") + scale_y_continuous("measure value") + facet_grid(measure~.,scales="free_y")
-quartz.save("../flux-entropy-paper/figures/context-flux-entropy-boxplot.pdf", type = "pdf")
+ggsave("../flux-entropy-paper/figures/context-flux-entropy-boxplot.pdf",width=5.9,height=3.6)
 
 ggplot(valid.sessions, aes(performance_context, trace)) + geom_boxplot(aes(fill = performance_type))
 ggplot(valid.sessions, aes(performance_context, norm)) + geom_boxplot(aes(fill = performance_type))
@@ -74,11 +74,9 @@ quartz.save("../flux-entropy-paper/figures/flux-entropy-distribution.pdf", type 
 #ggplot(valid.sessions, aes(time, entropy)) + geom_point(aes(colour = performance_context,shape = performance_type), alpha = 0.8) + theme(plot.margin=unit(rep(0,4), "cm"), legend.position = "right", legend.box = "vertical") + scale_colour_manual(values=chifig.5colours)
 #quartz.save("../flux-entropy-paper/figures/entropy-through-time.pdf", type = "pdf")
                                         # Flux and Entropy through time.
-ggplot(valid.sessions.long, aes(time, value)) + geom_point(aes(colour = performance_context,shape = performance_type), alpha = 0.8) + theme(plot.margin=unit(rep(0,4), "cm"), legend.position = "right", legend.box = "vertical") + scale_colour_manual(values=chifig.5colours) + scale_y_continuous("measure value") + facet_grid(measure~.,scales="free_y")
-quartz.save("../flux-entropy-paper/figures/flux-entropy-through-time.pdf", type = "pdf")
[…]

### 2015-09-15 flux-entropy-paper/flux-entropy-paper.tex
boom
first complete draft

@@
 
-\title{Quantitative Evaluation of Touch-Screen Performance with
+\title{Quantitative Evaluation of Touch-Screen Group Musical Performance with
   Transition Matrix Measures}
+\title{Mining interaction log data in long-term practice-led research}
 
@@
 \begin{abstract}
-  HCI systems for supporting collaborative creativity are often
-  evaluated with user ratings and interviews. However, such data can
-  be impractical to collect in-the-wild, for example during musical
+  Interfaces for supporting collaborative creativity are often
+  evaluated with user ratings and interviews. Such data can be
+  impractical to collect in-the-wild, for example during musical
   performances. In this paper we show how a transition matrix based
-  post-hoc analysis of logged interaction from a long program of
+  post-hoc analysis of logged interaction data from a long program of
   experiments in group musical creativity can shed light on the nature
@@
 
-\begin{itemize}
-\item we should be clear (and careful) where we're talking about
-  groups TMs, and where we're talking about individual performers
-\item what's the terminology we're using for a ``performance''? Is it
-  ``performance'', or is that too music-specific? ``Session''?
-  Alternatives are ``activity'', or ``interaction'', but they're so
-  general\ldots This decision has implications for our other
-  terminology: participant vs performer vs musician, etc.
-\item Title: maybe ``Traces through time: making sense of longitudinal
-  open-ended interaction data'' or some such? I think that's our big
-  contribution: it takes the in-the-wild, practitioner-driven research
-  program stuff and tries to make some quantitative sense out of it
-  post-hoc.
-\end{itemize}
+% \begin{itemize}
+% \item we should be clear (and careful) where we're talking about
+%   groups TMs, and where we're talking about individual performers
+% \item what's the terminology we're using for a ``performance''? Is it
+%   ``performance'', or is that too music-specific? ``Session''?
+%   Alternatives are ``activity'', or ``interaction'', but they're so
+%   general\ldots This decision has implications for our other
+%   terminology: participant vs performer vs musician, etc.
+% \item Title: maybe ``Traces through time: making sense of longitudinal
+%   open-ended interaction data'' or some such? I think that's our big
+%   contribution: it takes the in-the-wild, practitioner-driven research
+%   program stuff and tries to make some quantitative sense out of it
+%   post-hoc.
+% \end{itemize}
 
@@
 
-Ideas: the in-the-wild nature, and the different contexts - both the
-fact that they are different, and the fact that their differences make
-collecting certain data/responses easier or harder
+Computing devices and interfaces continue to open up rich, open-ended
+interaction contexts such as group music-making. Understanding and
+evaluating interfaces in these contexts presents many challenges;
+making use of the easily-logged interaction data without descending
+into ``quantitative reductionism'', as well as balancing the desire to
+gather ``felt experience'' data through surveys/interviews without
+unnecessary interruptions to the participants.
 
-\section{Transition Matrices and Matrix Measures}
+In this paper we show how we have used post-hoc analyses of log data
+to support a long-term practice-led research program into iPad
+interfaces for group music-making. Over a period of more than two
+years (April 2013--June 2015), we have collected data from $150$ group
+music-making sessions where participants used musical touch-screen
+interfaces on Apple iPad devices. These interfaces (see\cite{TODO} for
+details) shared a common mode of interaction where the majority of
+the screen could be used for free form touch interactions; tapping
+produced short sounds, while swiping or swirling produced long sounds
+with volume proportional to the velocity of the moving touch point.
+Through all of these performances every touch interaction---every
+touch down, dragged and up---was timestamped and logged to disk for
+later analysis.
 
-Open-ended collaborative interactions such as improvised musical
-performances can be modelled as a sequence of
-% could also use Thimbleby:2004fj
-states\cite{Thimbleby:2001kq,Swift:2014tya}. In this section we use
-transition matrix (TM) statistics to analyse the transitions between
-adjacent states in such a sequence. We then present two high-level
-derived quantities---\emph{flux} and \emph{entropy}---which can be
-used to compare musical performances under different circumstances.
+These sessions were part of a long-term research project into music
+touch-screen interaction, as well as an ongoing creative practice. The
+log data collected over these sessions is homogeneous in form, but
+represents a wide variety of performance contexts: rehearsals and
+performances, written compositions and free improvisations, controlled
+lab studies and casual demonstrations, with and without additional
+acoustic instruments (some of these are shown in
+Figure~\ref{fig:three-performance-contexts}).
 
-Suppose that each participant's activity in a collaborative
-interaction can at any given time be classified as belonging to one of
-$m$ distinct \emph{states}. Over the course of the activity, each
-participant $X$ generates a length $N$ sequence of these interaction
-states
-\begin{equation}
- X_n \hskip 2em n = 1, \ldots, N
-\end{equation}
-The transitions between these interaction states can be summarised in
-a transition matrix (TM) as if the sequence $X_n$ were a first-order Markov
-chain. Let $N_{ij}$ be the number of times that state $j$ follows
-state $i$. Each participant's TM $P$, will then be an
-$m \times m$ matrix with entries
-\begin{equation}
-  p_{ij} = \frac{N_{ij}}{\sum_i \sum_j N_{ij}}
-\end{equation}
-Each entry of $P$ will be such that the transition probability from
-state $i$ to state $j$ is given by the entry in the $i^{th}$ row and
-the $j^{th}$ column. The transition activity of the whole ensemble can
-be calculated by averaging the TM for each performer.
-The matrix can be normalised so that the sum of all elements is equal
-to one.
-
-If there is an (automated) classification process for determining each
-participant's state sequence during an activity, then it is a simple
-process to generate a TM for each participant and for the group as a
-whole. The challenge is to understand what the TM reveals about the
-group interaction, and to allow for comparisons between sessions and
-groups.
-
-To compare the TMs of different interactions we use two high-level
-quantities, ``flux'' and ``entropy''. The flux of a sequence is
-defined as the ratio of change transitions (e.g. $A \rightarrow B$
-where $A \neq B$), to self transitions (e.g. $A \rightarrow A $). The
-flux of a TM $P$ is given by
-\begin{equation}
-  \mathrm{flux}(P) = 1 - \mathrm{tr}(P)
-\end{equation}
-where the trace $\mathrm{tr}(P)$ is the sum of the diagonal entries of
-$P$. Intuitively, flux is a measure of how frequently the participant/group
-changes state. Flux returns a value in the interval $[0,1]$ and will
-return 0 when participants never change state, and 1 when participants
-never stay in the same state for two measurements in a row.
-
-Another useful measure that can be used on a transition matrix is its
-entropy, defined in the information-theoretic~\cite{Shannon:1948rt}
-sense:
-\begin{equation}
-  H(P) = -\sum_{i,j}p_{ij}\log_2(p_{ij})
-\end{equation}
-This entropy measure is small when the matrix is sparse, and largest
-when each matrix element is equal. It offers a different perspective
-on collaborative interactions than the flux measure by capturing the
-breadth of the gestural space explored by the participants throughout
-the course of an interaction. Consider the degenerate case of a
-participant who alternates between two states for a whole performance:
-the flux in this case will be maximal ($\mathrm{flux}(P) = 1$) since
-there are no self-transitions, (only $A \rightarrow B$ or
-$ B \rightarrow A$) even though the participant has only used a small
-subset of the state space. The entropy of this interaction, on the
-other hand, will be low. Entropy, therefore, is a measure of how broad
-the participant's exploration of the state space is in a given
-interaction.
-
-% \section{Making sense of two year's worth of touch-screen
-%   interactions}
-\section{A corpus of touch-screen interaction}
+The median length of these sessions was 7m29s and the median number of
+participants was four; other descriptive statistics about the sessions
+are shown in Table \ref{corpus-table}. The session data includes
+metadata regarding the session \emph{context} (whether it was a
+performance, rehearsal, study, demonstration, or data collection
+session), session \emph{type} (either improvisation or composition),
+and the instruments used (iPads alone, or iPads with acoustic
+percussion). The distribution of recorded sessions by context and type
+is shown in Figure \ref{fig:count-data}, and we shall use these
+groupings for our analysis in
+Section~\ref{differentiating-interaction-sessions}.
 
@@
 
+In some of these sessions, surveys and open-ended interviews were
+conducted with the participating musicians in an effort to understand
+the interaction dynamics and felt experience of the participants.
+However, many of these sessions were in-the-wild musical performances,
+and in these cases no such subjective evaluation data was
+collected---only the log data. This raises the question: can we use
+the log data to fill in some of the missing ``subjective experience''
+data, and to understand broader trends in the interaction behaviour of
+the musicians over this period?
+
+\section{Transition Matrices and Matrix Measures}
+
+Open-ended collaborative interactions such as improvised musical
+performances can be modelled as a sequence of
+% could also use Thimbleby:2004fj
+states\cite{Thimbleby:2001kq,Swift:2014tya}. In this section we use
+transition matrix (TM) statistics to analyse the transitions between
+adjacent states in such a sequence. We then present two high-level
+derived quantities---\emph{flux} and \emph{entropy}---which can be
+used to compare musical performances under different circumstances.
+
+Suppose that each participant's activity in a collaborative
+interaction can at any given time be classified as belonging to one of
+$m$ distinct \emph{states}. Over the course of the activity, each
+participant $X$ generates a length $N$ sequence of these interaction
+states
+\begin{equation}
+ X_n \hskip 2em n = 1, \ldots, N
+\end{equation}
+which can be thought of as Markov process. If there is an
+(automated) classification process for extracting each participant's
+state sequence from (or during) their interaction data, then we can
+estimate the $m \times m$ transition matrix $P$ which characterises
+this process by simply counting the number of times state $j$ follows
+state $i$
+\begin{equation}
+  p_{ij} = \frac{N_{ij}}{\sum_i \sum_j N_{ij}}
+\end{equation}
+The transition activity of the whole ensemble can be calculated by
+averaging the TM for each performer. The matrix can be normalised so
+that the sum of all elements is equal to one.
+
+From previous work\cite{TODO} we have a robust technique for
+segmenting the raw touch data generated by the instrument into a
+sequence of high-level ``gestures'' using a Random
+Forest\cite{Breiman:2001kx} classifier\footnote{The classifier used
+  was from Python's Scikit-learn package\cite{scikit-learn}.}. The
+classifier was applied to 5-second windows of the touch-data to
+provide a chain of gesture-states for each participant in each
+session. These chains can be used to create $9 \times 9$ (the
+classifier recognises 9~distinct gestures) transition matrices for
+individual performers and the whole ensemble in each session.
+
+\section{Quantitative transition matrix statistics}
+\label{sec:underst-impr-group}
+
+While the interaction sessions recorded in our corpus were performed
+by many different participants with varying experience levels, in
+different contexts and on different instrumental setups, the common
+interface and method of recording the data allows all the sessions to
+be transformed into gesture sequences. These transition matrices have
+been used as a visualisation technique to reveal qualitative
+differences in structure in improvisational
+music-making\cite{Swift:2014tya}. In this paper, however, we wish to
+extract quantitative statistics from these TMs for comparing and
+tracking different sessions over time.
+
+To do this we calculate two high-level quantities from each session:
+``flux'' and ``entropy''. The flux of a sequence is defined as the
+ratio of state transitions (e.g. $A \rightarrow B$ where $A \neq B$),
+to self transitions (e.g. $A \rightarrow A $). The flux of a TM $P$ is
+given by $\mathrm{flux}(P) = 1 - \mathrm{tr}(P)$ where the trace
+$\mathrm{tr}(P)$ is the sum of the diagonal entries of $P$.
+Intuitively, flux is a measure of how frequently the participant/group
+changes state. Flux returns a value in the interval $[0,1]$ and will
+return 0 when participants never change state, and 1 when participants
+never stay in the same state for two measurements in a row.
+
+Another useful measure that can be used on a transition matrix is its
+entropy, defined in the information-theoretic\cite{Shannon:1948rt}
+sense:
+\begin{equation}
+  H(P) = -\sum_{i,j}p_{ij}\log_2(p_{ij})
+\end{equation}
+This entropy measure is small when the matrix is sparse, and largest
+when each matrix element is equal. It offers a different perspective
+on collaborative interactions than the flux measure by capturing the
+breadth of the gestural space explored by the participants throughout
+the course of an interaction. Consider the degenerate case of a
+participant who alternates between two states for a whole performance:
+the flux in this case will be maximal ($\mathrm{flux}(P) = 1$) since
+there are no self-transitions, (only $A \rightarrow B$ or
+$ B \rightarrow A$) even though the participant has only used a small
+subset of the state space. The entropy of this interaction, on the
+other hand, will be low. Entropy, therefore, is a measure of how broad
+the participant's exploration of the state space is in a given
+interaction.
+
+In the rest of this paper we will use these matrix measures to answer
+two questions. First, do the transition matrix measures allow us to
+distinguish between different performance contexts, types and
+instrumental setups? Secondly, do these performance measure allow us
+further insight into the corpus than the metadata can provide? TODO I
+think this second question is too broad?
+
 \begin{figure}
@@
 \end{figure}
-
-% TODO maybe a ref to the CHI paper here?
-
-% TODO I think this ``scene-setting'' stuff could actually go in the
-% first section of the paper, with the TM stuff following.  But I'm
-% open to push-back on that.
-
-Over a period of more than two years (April 2013 - June 2015), we have
-collected data from $150$ group music-making sessions where
-participants used musical touch-screen interfaces on Apple iPad
-devices. These interfaces all shared a common mode of interaction
-where the majority of the screen could be used for free form touch
-interactions; tapping produced short sounds, while swiping or swirling
-produced long sounds with volume proportional to the velocity of the
-moving touch point. Through all of these performances every touch
-interaction---every touch down, dragged and up---was timestamped and
-logged to disk for later analysis.
-
-These sessions were part of a long-term research project into music
-touch-screen interaction, as well as an ongoing creative practice. As
-a result, the sessions had a variety of contexts including rehearsals,
-performances, lab study sessions, classes, and casual demonstrations,
-some of which are shown in Figure
-\ref{fig:three-performance-contexts}. Participants performed free
-improvisations as well as written compositions and performed on iPads
-alone or using setups of iPads and acoustic percussion instruments.
-The median length of these sessions was 7m29s and the median number of
-participants was four, other descriptive statistics about the sessions
-is shown in Table \ref{corpus-table}. Each session was coded with
-metadata regarding the session context (whether it was a performance,
-rehearsal, study, demonstration, or data collection session), the
-session type (either improvisation or composition), and the
-instruments used (iPads alone, or iPads with acoustic percussion). The
-distribution of recorded sessions by context and type is shown in
-Figure \ref{fig:count-data}.
-
-% TODO again, I think this bit would be beefed-up if we included the
-% fact that it's already been published.  And it also means we don't
-% have to go into as much detail about how we did it - the main thing
-% is that we've got a sensible (and peer-reviewed!) way of taking the
-% touch interaction data and generating gesture sequences (and
-% therefore TMs)
-
-To analyse the gestural interactions present in each of these
-sessions, the sessions were automatically coded according to a
-vocabulary of nine simple, continuous, touch-screen gestures. Examples
-of these gestures was generated by an expert user in formal process
-where each gesture was performed for 1-minute and feature vectors of
-discriptive statistics on these examples was used to train a Random
-Forest~\cite{Breiman:2001kx} classifier\footnote{The classifier used
-  was from Python's Scikit-learn package~\cite{scikit-learn}.}. The
-classifier was applied to 5-second windows of the touch-data each
-second provide a chain of gesture-states for each participant in each
-session. These chains can be used to create $9 \times 9$
-gesture-transition matrices for individual performers, and the whole
-ensemble in each session.
-
-While the interaction sessions recorded in our corpus were performed
-by many different participants with varying experience levels, in
-different contexts and on different instrumental setups, the common
-interface and method of recording the data allows all the sessions to
-be transformed into gesture sequences. This common format allows us to
-ask if performance styles vary in between these different situations.
-The sequences can be plotted and inspected qualitatively, but
-transition matrices present a quantitative approach to analysing such
-interactions.
-
-In the rest of this paper we will use our matrix measure techniques to
-interrogate this dataset and to answer two questions. First, do the
-transition matrix measures allow us to distinguish between different
-performance contexts, types and instrumental setups? Secondly, do
-these performance measure allow us further insight into the corpus
-than the metadata can provide?
 
@@
 % have we?
-In the following analysis, only the 95 performative sessions were
-considered, excluding the demonstration and data-collection contexts.
+In the following analysis, only the 95~``performance context''
+sessions were considered, excluding the demonstration and
+data-collection contexts. TODO can we justify this?
 
@@
 matrices.
-
-\begin{itemize}
-\item TODO: need to add overall contrasts column to do pairwise t-tests.
-\end{itemize}
 
@@
 
-It is clear that flux and entropy can detect the variation in
-performance styles in between performances of different types, but can
-they be used to understand some of the internal structure of
-performances? A simple model for understanding temporal art-forms is
-the ``beginning, middle, end'' structure. To investigate how this
-structure fits our corpus, we can divide the gestural-sequences of
-each session into thirds and calculate the transition matrices given
-by each section.
+It is clear from these results that flux and entropy can be used to
+differentiate between performance styles and types, but can they be
+used to understand some of the internal structure of performances? A
+simple model for understanding temporal art-forms is the ``beginning,
+middle, end'' structure. To investigate how this structure fits our
+corpus, we can divide the gestural-sequences of each session into
+thirds and calculate the transition matrices given by each section.
 
@@
 % difficult, both due to the problem of treating ordinal Likert
-% responses as numeric~\cite{Gardner:2007fj} and also
+% responses as numeric\cite{Gardner:2007fj} and also
 % due to the unknown distribution of the entropy and flux statistics. As

### 2015-09-20 flux-entropy-paper/flux-entropy-paper.tex
cleaned up beginning section headers
Henry's updates (2015-09-18)

@@
 
-\title{Quantitative Evaluation of Touch-Screen Group Musical Performance with
-  Transition Matrix Measures}
-\title{Mining interaction log data in long-term practice-led research}
+%\title{Quantitative Evaluation of Touch-Screen Group Musical Performance with
+ % Transition Matrix Measures}
+%\title{Mining interaction log data in long-term practice-led research}
+
+\title{Mining Interaction Log Data in a Creative Arts Practice using Macro Gestures and Transition Matrices}
+
+%in a Creative Arts Practice}
 
@@
 \begin{abstract}
-  Interfaces for supporting collaborative creativity are often
-  evaluated with user ratings and interviews. Such data can be
-  impractical to collect in-the-wild, for example during musical
-  performances. In this paper we show how a transition matrix based
-  post-hoc analysis of logged interaction data from a long program of
-  experiments in group musical creativity can shed light on the nature
-  of the group interaction in-the-wild. Using a corpus of more than
-  140 performances, rehearsals, and demonstrations over more than two
-  years, we find that two matrix measures, flux and entropy, can be
-  used as a proxy for performer ratings during in-the-wild
-  interaction.
+
+%condiser the following? .............
+
+  %Digital arts practices can allow for the archival of interaction log data.
+  %Such log data can capture and preserve performances of new-media artists and musicians and
+  %and can facilitate post-hoc data processing and data mining.
+
+  %In this note we describe a data archive for an iPad musical-ensemble performance practice which has included 
+  %concerts, demonstrations and formal HCI studies of groups of musicians over more than two years. Our corpus contains data for 
+  %150 performances, rehearsals, and demonstrations, involving some 100 different musicians and several different iPad apps.
+  %At the lowest level, data for each musician in a particular performance was archived as Apple iOS callback methods for touching and swiping.
+Subsequent data analysis has been undertaken by first classifying these data into higher level gesture types and then describing the transitions between these gestures for each artist as a function of time.
+We
+%find that a transition-matrix measure, which we call ``flux'', can be
+%used to distinguish different performance types and contexts, as well as different phases (beginning, middle and end) of all of the performances in our corpus.
+
 \end{abstract}
@@
 
-\section{Introduction}
+Modern computing devices and interfaces continue to open up rich, open-ended
+creative opportunities in the new media arts. 
+Understanding and evaluating interfaces in these contexts presents many challenges
+[refs...Resnick, Mitchel, Brad Myers, Kumiyo Nakakoji, Ben Shneiderman, Randy Pausch, Ted Selker, and Mike Eisenberg.
+"Design principles for tools to support creative thinking." (2005) http://repository.emu.edu/isr; Shneiderman, Ben.
+"Creativity support tools: Accelerating discovery and innovation." Communications of the ACM 50.12 (2007): 20-32.].
+In the context of live performance, interface evaluation, using, for example, post-performance questionnaires, can omit consideration of the dynamic needs of artists in the moment of the performance and in the wild of the performance context.
+In contrast to traditional art forms, logged interaction data from a new media arts performance can provide an enduring record of that performance for post-hoc quantitative investigation as well as for curation and preservation[ref: England, David, et al.
+"Curating the digital: Spaces for art and interaction." CHI'14 Extended Abstracts on Human Factors in Computing Systems.
+ACM, 2014.].
+Over time, the accumulation of new-media interaction log data might allow data-mining techniques to be deployed to potentially describe universal features of such interaction. 
+%Additionally, a wider dissemination of such log data will better support the replication of findings in HCI research [ref:Wilson, Max L., et al.
+"RepliCHI-CHI should be replicating and validating results more: discuss." CHI'11 Extended Abstracts on Human Factors in Computing Systems.
+ACM, 2011.]
 
-Computing devices and interfaces continue to open up rich, open-ended
-interaction contexts such as group music-making. Understanding and
-evaluating interfaces in these contexts presents many challenges;
-making use of the easily-logged interaction data without descending
-into ``quantitative reductionism'', as well as balancing the desire to
-gather ``felt experience'' data through surveys/interviews without
-unnecessary interruptions to the participants.
+It is a feature of creative  expression in the arts that highly skilled practitioners exploit simple tools to produce works of beauty whether it be static paintings or dynamic performance.
+In previous work [ref: charles at CHI] researchers have shown that trained percussionists can exploit a very simple iPad interface to discover new categories of musical gestures.
+In that paper, iPad apps were studied where tapping was mapped to produce produced short sounds, while swiping produced long sounds
+with the volume proportional to the velocity of the moving touch point.
+It turned out that trained percussionists working together in an ensemble were able to discover and repeatedly make use of five macro gestures made up of taps and swipes: fast taps, slow taps, random taps, short swipes and swirls.
+This process of discovery of interesting musical gestures from a simple interface is analogous to what percussionists do with acoustic instruments -- drums and bowls and bells are often tapped, stroked and hit in novel ways to communicate meaning and emotion through music [ref to Schick ].
 
-In this paper we show how we have used post-hoc analyses of log data
-to support a long-term practice-led research program into iPad
-interfaces for group music-making. Over a period of more than two
-years (April 2013--June 2015), we have collected data from $150$ group
-music-making sessions where participants used musical touch-screen
-interfaces on Apple iPad devices. These interfaces (see\cite{TODO} for
-details) shared a common mode of interaction where the majority of
-the screen could be used for free form touch interactions; tapping
+In this note we describe a log data archive that has been accumulated to 
+support a long-term practice-led research program into simple iPad
+interfaces for group music-making similar to those described in [ref: charles at CHI] . Over a period of more than two
+years (April 2013--June 2015), we collected data from $....$ group
+music-making sessions where the participants used custom-built, touch-screen
+apps on Apple iPad devices. Although several different apps have been included in our corpus, (see\cite{anon} for
+details) they all shared the simple mode of interaction where the majority of
+the screen could be used for free-form touch interactions; tapping
 produced short sounds, while swiping or swirling produced long sounds
-with volume proportional to the velocity of the moving touch point.
-Through all of these performances every touch interaction---every
-touch down, dragged and up---was timestamped and logged to disk for
+with the volume proportional to the velocity of the moving touch point.
+Apart from this commonality, our apps featured different palettes of available sounds and modes of synthesis, different arrangements of sounds on the iPad screens, different kinds of visual display and feedback, different kinds of networked interactions (between the iPads in an ensemble and with a central server), as well as a variety of special features for each app.
+
+In all of  our data-collection sessions,  every touch interaction---every
+touch down, dragging movement and touch up---was time-stamped and logged to disk for
 later analysis.
+These sessions included rehearsals without an audience as well as live performances at art festivals and other events.
+They also included two formally-structured HCI experiments in order to understand and iterate on the system and interface design.
+Some of the performances were iPad-only and others used iPads together with other acoustic percussion instruments.
+Although the majority of the sessions were free-form improvisations, a number of rehearsals and performances involved composed pieces where the musicians read and interpreted percussion notation as they used the apps.
+Representative photographs of some of these session contexts (rehearsal, performance including other acoustic instruments and demonstrations) are are shown
+Figure~\ref{fig:three-performance-contexts}) and other descriptive statistics about the sessions
+are shown in Table \ref{corpus-table}. In our corpus, the median length of sessions is 7m29s and the median number of
+participants per ensemble is four; .
+Metadata is included together with the interaction data for each session to record the \emph{context}  (performance, rehearsal, study), session \emph{type}  (improvisation or composition), and  \emph{instrumentation} (iPads-only, or iPads with acoustic
+percussion). The distribution of recorded sessions by context and type
+is shown in Figure \ref{fig:count-data}.
 
-These sessions were part of a long-term research project into music
-touch-screen interaction, as well as an ongoing creative practice. The
-log data collected over these sessions is homogeneous in form, but
-represents a wide variety of performance contexts: rehearsals and
-performances, written compositions and free improvisations, controlled
-lab studies and casual demonstrations, with and without additional
-acoustic instruments (some of these are shown in
-Figure~\ref{fig:three-performance-contexts}).
+NOTE: FORGET ABOUT THE DEMONSTRATIONS AND THE DATA COLLECTION PARTS OF THE CORPUS. THEY JUST MAKE THE STORY VAGUE. YOU DON'T INCLUDE THEM IN YOUR ANALYSIS AT ALL.
 
-The median length of these sessions was 7m29s and the median number of
-participants was four; other descriptive statistics about the sessions
-are shown in Table \ref{corpus-table}. The session data includes
-metadata regarding the session \emph{context} (whether it was a
-performance, rehearsal, study, demonstration, or data collection
-session), session \emph{type} (either improvisation or composition),
-and the instruments used (iPads alone, or iPads with acoustic
-percussion). The distribution of recorded sessions by context and type
-is shown in Figure \ref{fig:count-data}, and we shall use these
-groupings for our analysis in
-Section~\ref{differentiating-interaction-sessions}.
+NOTE: NEED TO MODIFY ALL OF THESE GRAPHS, TABLES AND FIGURES.
 
@@
   interaction sessions used in this study. These sessions were
-  rehearsals, performances, studies, and demonstration sessions and the
-  participants improvised or performed written compositions. Each record
+  rehearsals, performances, studies, and demonstration sessions and 
+  participants improvised or performed compositions. Each record
   consisted of touch-data recorded in CSV format and was later analysed
@@
 
-In some of these sessions, surveys and open-ended interviews were
-conducted with the participating musicians in an effort to understand
-the interaction dynamics and felt experience of the participants.
-However, many of these sessions were in-the-wild musical performances,
-and in these cases no such subjective evaluation data was
-collected---only the log data. This raises the question: can we use
-the log data to fill in some of the missing ``subjective experience''
-data, and to understand broader trends in the interaction behaviour of
-the musicians over this period?
+\section{Macro Gestures, Transition Matrices and Matrix Measures}
 
-\section{Transition Matrices and Matrix Measures}
+A useful way of modelling user interfaces is to consider their configurations as state vectors and to deploy the methods of matrix algebra to describe transitions between these states during an interaction sequence\cite{Thimbleby:2001kq, Thimbleby, Harold, Paul Cairns, and Matt Jones.
+"Usability analysis with Markov models." ACM Transactions on Computer-Human Interaction (TOCHI) 8.2 (2001): 99-132.}.
+Data obtained from usability analysis of quite complex systems (such as a resource planning application for anti-terrorism planning [ref: Kannampallil, Thomas, and Steven Haynes.
+"Using Event Based Markov Model Simulation to Analyze Interactive Human Behavior." AMCIS 2007 Proceedings (2007): 142.
+]) have been undertaken by coding sequences of fine-detail human actions as states and then to model data from usability walkthroughs as sequences of state transitions.
+The transition probably matrix for transitions between states can then be derived from such user data and conclusions drawn about the nature of interaction.
+In the creative arts, editing commands at a computer interface for the ``live coding of computer music'' have also been coded coded as states and analysed using transition matrices, in both the textual and musical dimensions, in order to arrive at conclusions regarding artistic style\cite{Swift:2014tya}.
 
-Open-ended collaborative interactions such as improvised musical
-performances can be modelled as a sequence of
-% could also use Thimbleby:2004fj
-states\cite{Thimbleby:2001kq,Swift:2014tya}. In this section we use
-transition matrix (TM) statistics to analyse the transitions between
-adjacent states in such a sequence. We then present two high-level
-derived quantities---\emph{flux} and \emph{entropy}---which can be
-used to compare musical performances under different circumstances.
+In our corpus of iPad musical performances, rehearsals and studies, the macro gestures identified in [ref: charlesHCI] provide a basis to identify and code macro gestures.
+Each performer in each muscial piece can then be thought of as following a ``score'' of state transitions and an overall transition matrix for the entire ensemble can be derived by averaging the individual transition matrices.
 
@@
 
-From previous work\cite{TODO} we have a robust technique for
-segmenting the raw touch data generated by the instrument into a
-sequence of high-level ``gestures'' using a Random
+Previous work showed that the raw touch data obtained from interaction with simple iPad interfaces similar to ours could
+be segmented  using a Random
 Forest\cite{Breiman:2001kx} classifier\footnote{The classifier used
@@
 
+Mention Manovich transmutation stuff
+
 % todo blah de blah

### 2015-09-20 flux-entropy-paper/flux-entropy-paper.tex
updates to refs so paper will compile

@@
 
-  %Digital arts practices can allow for the archival of interaction log data.
-  %Such log data can capture and preserve performances of new-media artists and musicians and
-  %and can facilitate post-hoc data processing and data mining.
+%   %Digital arts practices can allow for the archival of interaction
+% log data. Such log data can capture and preserve performances of
+% new-media artists and musicians and %and can facilitate post-hoc data
+% processing and data mining.
 
-  %In this note we describe a data archive for an iPad musical-ensemble performance practice which has included 
-  %concerts, demonstrations and formal HCI studies of groups of musicians over more than two years. Our corpus contains data for 
-  %150 performances, rehearsals, and demonstrations, involving some 100 different musicians and several different iPad apps.
-  %At the lowest level, data for each musician in a particular performance was archived as Apple iOS callback methods for touching and swiping.
-Subsequent data analysis has been undertaken by first classifying these data into higher level gesture types and then describing the transitions between these gestures for each artist as a function of time.
-We
-%find that a transition-matrix measure, which we call ``flux'', can be
-%used to distinguish different performance types and contexts, as well as different phases (beginning, middle and end) of all of the performances in our corpus.
+%   %In this note we describe a data archive for an iPad
+% musical-ensemble performance practice which has included %concerts,
+% demonstrations and formal HCI studies of groups of musicians over more
+% than two years. Our corpus contains data for %150 performances,
+% rehearsals, and demonstrations, involving some 100 different musicians
+% and several different iPad apps. %At the lowest level, data for each
+% musician in a particular performance was archived as Apple iOS
+% callback methods for touching and swiping. Subsequent data analysis
+% has been undertaken by first classifying these data into higher level
+% gesture types and then describing the transitions between these
+% gestures for each artist as a function of time. We %find that a
+% transition-matrix measure, which we call ``flux'', can be %used to
+% distinguish different performance types and contexts, as well as
+% different phases (beginning, middle and end) of all of the
+% performances in our corpus.
 
@@
 
-Modern computing devices and interfaces continue to open up rich, open-ended
-creative opportunities in the new media arts. 
-Understanding and evaluating interfaces in these contexts presents many challenges
-[refs...Resnick, Mitchel, Brad Myers, Kumiyo Nakakoji, Ben Shneiderman, Randy Pausch, Ted Selker, and Mike Eisenberg.
-"Design principles for tools to support creative thinking." (2005) http://repository.emu.edu/isr; Shneiderman, Ben.
-"Creativity support tools: Accelerating discovery and innovation." Communications of the ACM 50.12 (2007): 20-32.].
-In the context of live performance, interface evaluation, using, for example, post-performance questionnaires, can omit consideration of the dynamic needs of artists in the moment of the performance and in the wild of the performance context.
-In contrast to traditional art forms, logged interaction data from a new media arts performance can provide an enduring record of that performance for post-hoc quantitative investigation as well as for curation and preservation[ref: England, David, et al.
-"Curating the digital: Spaces for art and interaction." CHI'14 Extended Abstracts on Human Factors in Computing Systems.
-ACM, 2014.].
-Over time, the accumulation of new-media interaction log data might allow data-mining techniques to be deployed to potentially describe universal features of such interaction. 
+Modern computing devices and interfaces continue to open up rich,
+open-ended creative opportunities in the new media arts. Understanding
+and evaluating interfaces in these contexts presents many challenges~\cite{Resnick:2005yu,Shneiderman:2007qv}. In the context of live performance,
+interface evaluation, using, for example, post-performance
+questionnaires, can omit consideration of the dynamic needs of artists
+in the moment of the performance and in the wild of the performance
+context. In contrast to traditional art forms, logged interaction data
+from a new media arts performance can provide an enduring record of
+that performance for post-hoc quantitative investigation as well as
+for curation and preservation~\cite{England:2014ys}. Over
+time, the accumulation of new-media interaction log data might allow
+data-mining techniques to be deployed to potentially describe
+universal features of such interaction.
+
 %Additionally, a wider dissemination of such log data will better support the replication of findings in HCI research [ref:Wilson, Max L., et al.
@@
 
-It is a feature of creative  expression in the arts that highly skilled practitioners exploit simple tools to produce works of beauty whether it be static paintings or dynamic performance.
-In previous work [ref: charles at CHI] researchers have shown that trained percussionists can exploit a very simple iPad interface to discover new categories of musical gestures.
-In that paper, iPad apps were studied where tapping was mapped to produce produced short sounds, while swiping produced long sounds
-with the volume proportional to the velocity of the moving touch point.
-It turned out that trained percussionists working together in an ensemble were able to discover and repeatedly make use of five macro gestures made up of taps and swipes: fast taps, slow taps, random taps, short swipes and swirls.
-This process of discovery of interesting musical gestures from a simple interface is analogous to what percussionists do with acoustic instruments -- drums and bowls and bells are often tapped, stroked and hit in novel ways to communicate meaning and emotion through music [ref to Schick ].
+It is a feature of creative expression in the arts that highly skilled
+practitioners exploit simple tools to produce works of beauty whether
+it be static paintings or dynamic performance. In previous
+work~\cite{Martin:2014cr} researchers have shown that trained
+percussionists can exploit a very simple iPad interface to discover
+new categories of musical gestures. In that paper, iPad apps were
+studied where tapping was mapped to produce produced short sounds,
+while swiping produced long sounds with the volume proportional to the
+velocity of the moving touch point. It turned out that trained
+percussionists working together in an ensemble were able to discover
+and repeatedly make use of five macro gestures made up of taps and
+swipes: fast taps, slow taps, random taps, short swipes and swirls.
+This process of discovery of interesting musical gestures from a
+simple interface is analogous to what percussionists do with acoustic
+instruments -- drums and bowls and bells are often tapped, stroked and
+hit in novel ways to communicate meaning and emotion through music~\cite{Schick:2006fk}.
 
-In this note we describe a log data archive that has been accumulated to 
-support a long-term practice-led research program into simple iPad
-interfaces for group music-making similar to those described in [ref: charles at CHI] . Over a period of more than two
-years (April 2013--June 2015), we collected data from $....$ group
-music-making sessions where the participants used custom-built, touch-screen
-apps on Apple iPad devices. Although several different apps have been included in our corpus, (see\cite{anon} for
-details) they all shared the simple mode of interaction where the majority of
-the screen could be used for free-form touch interactions; tapping
-produced short sounds, while swiping or swirling produced long sounds
-with the volume proportional to the velocity of the moving touch point.
-Apart from this commonality, our apps featured different palettes of available sounds and modes of synthesis, different arrangements of sounds on the iPad screens, different kinds of visual display and feedback, different kinds of networked interactions (between the iPads in an ensemble and with a central server), as well as a variety of special features for each app.
+In this note we describe a log data archive that has been accumulated
+to support a long-term practice-led research program into simple iPad
+interfaces for group music-making similar to those described in [ref:
+charles at CHI] . Over a period of more than two years (April
+2013--June 2015), we collected data from $....$ group music-making
+sessions where the participants used custom-built, touch-screen apps
+on Apple iPad devices. Although several different apps have been
+included in our corpus,
+% TODO: should there be a citation here?
+ they all shared
+the simple mode of interaction where the majority of the screen could
+be used for free-form touch interactions; tapping produced short
+sounds, while swiping or swirling produced long sounds with the volume
+proportional to the velocity of the moving touch point. Apart from
+this commonality, our apps featured different palettes of available
+sounds and modes of synthesis, different arrangements of sounds on the
+iPad screens, different kinds of visual display and feedback,
+different kinds of networked interactions (between the iPads in an
+ensemble and with a central server), as well as a variety of special
+features for each app.
 
-In all of  our data-collection sessions,  every touch interaction---every
-touch down, dragging movement and touch up---was time-stamped and logged to disk for
-later analysis.
-These sessions included rehearsals without an audience as well as live performances at art festivals and other events.
-They also included two formally-structured HCI experiments in order to understand and iterate on the system and interface design.
-Some of the performances were iPad-only and others used iPads together with other acoustic percussion instruments.
-Although the majority of the sessions were free-form improvisations, a number of rehearsals and performances involved composed pieces where the musicians read and interpreted percussion notation as they used the apps.
-Representative photographs of some of these session contexts (rehearsal, performance including other acoustic instruments and demonstrations) are are shown
-Figure~\ref{fig:three-performance-contexts}) and other descriptive statistics about the sessions
-are shown in Table \ref{corpus-table}. In our corpus, the median length of sessions is 7m29s and the median number of
-participants per ensemble is four; .
-Metadata is included together with the interaction data for each session to record the \emph{context}  (performance, rehearsal, study), session \emph{type}  (improvisation or composition), and  \emph{instrumentation} (iPads-only, or iPads with acoustic
-percussion). The distribution of recorded sessions by context and type
-is shown in Figure \ref{fig:count-data}.
+In all of our data-collection sessions, every touch
+interaction---every touch down, dragging movement and touch up---was
+time-stamped and logged to disk for later analysis. These sessions
+included rehearsals without an audience as well as live performances
+at art festivals and other events. They also included two
+formally-structured HCI experiments in order to understand and iterate
+on the system and interface design. Some of the performances were
+iPad-only and others used iPads together with other acoustic
+percussion instruments. Although the majority of the sessions were
+free-form improvisations, a number of rehearsals and performances
+involved composed pieces where the musicians read and interpreted
+percussion notation as they used the apps. Representative photographs
+of some of these session contexts (rehearsal, performance including
+other acoustic instruments and demonstrations) are are shown
+Figure~\ref{fig:three-performance-contexts}) and other descriptive
+statistics about the sessions are shown in Table \ref{corpus-table}.
+In our corpus, the median length of sessions is 7m29s and the median
+number of participants per ensemble is four; . Metadata is included
+together with the interaction data for each session to record the
+\emph{context} (performance, rehearsal, study), session \emph{type}
+(improvisation or composition), and \emph{instrumentation}
+(iPads-only, or iPads with acoustic percussion). The distribution of
+recorded sessions by context and type is shown in Figure
+\ref{fig:count-data}.
 
@@
 
-A useful way of modelling user interfaces is to consider their configurations as state vectors and to deploy the methods of matrix algebra to describe transitions between these states during an interaction sequence\cite{Thimbleby:2001kq, Thimbleby, Harold, Paul Cairns, and Matt Jones.
-"Usability analysis with Markov models." ACM Transactions on Computer-Human Interaction (TOCHI) 8.2 (2001): 99-132.}.
+A useful way of modelling user interfaces is to consider their configurations as state vectors and to deploy the methods of matrix algebra to describe transitions between these states during an interaction sequence\cite{Thimbleby:2001kq, Thimbleby:2004fj}.
 Data obtained from usability analysis of quite complex systems (such as a resource planning application for anti-terrorism planning [ref: Kannampallil, Thomas, and Steven Haynes.
@@
 
-Mention Manovich transmutation stuff
-
-% todo blah de blah
+%TODO Mention Manovich transmutation stuff
 

### 2015-09-20 flux-entropy-paper/flux-entropy-paper.tex
cut out some commented material at the end
minor edits and bib refs.

@@
 
-%condiser the following? .............
+%consider the following? .............
 
@@
 %   %In this note we describe a data archive for an iPad
-% musical-ensemble performance practice which has included %concerts,
+% musical-ensemble performance practice which has included concerts,
 % demonstrations and formal HCI studies of groups of musicians over more
-% than two years. Our corpus contains data for %150 performances,
+% than two years. Our corpus contains data for 150 performances,
 % rehearsals, and demonstrations, involving some 100 different musicians
-% and several different iPad apps. %At the lowest level, data for each
+% and several different iPad apps. At the lowest level, data for each
 % musician in a particular performance was archived as Apple iOS
@@
 % gesture types and then describing the transitions between these
-% gestures for each artist as a function of time. We %find that a
-% transition-matrix measure, which we call ``flux'', can be %used to
+% gestures for each artist as a function of time. We find that a
+% transition-matrix measure, which we call ``flux'', can be used to
 % distinguish different performance types and contexts, as well as
@@
 
-\section{General comments}
+%\section{General comments}
 
@@
 
-%Additionally, a wider dissemination of such log data will better support the replication of findings in HCI research [ref:Wilson, Max L., et al.
-"RepliCHI-CHI should be replicating and validating results more: discuss." CHI'11 Extended Abstracts on Human Factors in Computing Systems.
-ACM, 2011.]
+% Additionally, a wider dissemination of such log data will better
+% support the replication of findings in HCI
+% research~\cite{Wilson:2011ve}
 
@@
 % TODO: should there be a citation here?
- they all shared
-the simple mode of interaction where the majority of the screen could
-be used for free-form touch interactions; tapping produced short
-sounds, while swiping or swirling produced long sounds with the volume
-proportional to the velocity of the moving touch point. Apart from
-this commonality, our apps featured different palettes of available
-sounds and modes of synthesis, different arrangements of sounds on the
-iPad screens, different kinds of visual display and feedback,
-different kinds of networked interactions (between the iPads in an
-ensemble and with a central server), as well as a variety of special
-features for each app.
+they all shared the simple mode of interaction where the majority of
+the screen could be used for free-form touch interactions; tapping
+produced short sounds, while swiping or swirling produced long sounds
+with the volume proportional to the velocity of the moving touch
+point. Apart from this commonality, our apps featured different
+palettes of available sounds and modes of synthesis, different
+arrangements of sounds on the iPad screens, different kinds of visual
+display and feedback, different kinds of networked interactions
+(between the iPads in an ensemble and with a central server), as well
+as a variety of special features for each app.
 
@@
 
-NOTE: FORGET ABOUT THE DEMONSTRATIONS AND THE DATA COLLECTION PARTS OF THE CORPUS. THEY JUST MAKE THE STORY VAGUE. YOU DON'T INCLUDE THEM IN YOUR ANALYSIS AT ALL.
+% TODO: NOTE: FORGET ABOUT THE DEMONSTRATIONS AND THE DATA COLLECTION
+% PARTS OF THE CORPUS. THEY JUST MAKE THE STORY VAGUE. YOU DON'T
+% INCLUDE THEM IN YOUR ANALYSIS AT ALL.
 
-NOTE: NEED TO MODIFY ALL OF THESE GRAPHS, TABLES AND FIGURES.
+% TODO: NOTE: NEED TO MODIFY ALL OF THESE GRAPHS, TABLES AND FIGURES.
 
@@
 %TODO Mention Manovich transmutation stuff
+The macro-gestures that these processes reveal could also be used as
+the basis for new works of new media art~\cite{Manovich:2002ly}.
 
-% \section{Modelling Performer Ratings with Flux and Entropy}
-
-% \begin{figure}
-%   \centering
-%   \includegraphics[width=\linewidth]{figures/stat-response-fit.pdf}
-%   \caption{Transition matrix statistics (flux above and entropy below) vs Likert responses for
-%     questions 3, 6 and 7, all of which exhibited significant
-%     relationships under a proportional odds logistic regression model.
-%     A least-squares linear model is shown with a dotted line.
-%     \label{fig:stat-response-fit}}
-% \end{figure}
-
-% One final question to explore is how these flux and entropy measures
-% in the interaction data relate to the performers' perceptions of their
-% performance?
-
-% To answer this question, we construct models with the flux and entropy
-% measures as independent variables and the Likert responses (1 to 5) as
-% the dependent variable. These measures exhibited clear relationships
-% for three of the survey questions (Questions 3, 6 and 7), which are
-% plotted in Figure~\ref{fig:stat-response-fit} where the least-squares
-% line of best fit has a positive gradient for both flux and entropy.
-
-% Determining the strengths of these positive relationships is
-% difficult, both due to the problem of treating ordinal Likert
-% responses as numeric\cite{Gardner:2007fj} and also
-% due to the unknown distribution of the entropy and flux statistics. As
-% a result, we use a conservative proportional odds logistic regression
-% (POLR)~(using the \texttt{polr} function in \texttt{R}, from the MASS
-% package \cite{Venables:2002qv}) to treat the Likert responses as an
-% ordered factor response.
-
-% The POLR results indicate a significant ($p<0.05$) relationship
-% between the user-reported Likert response and the flux statistic for
-% Question 3 ($t = 2.77, df = 4, p = 0.025$), Question 6
-% ($t = 2.41, df = 4, p = 0.037$), and Question 7
-% ($t = 2.49, df = 4, p = 0.034$). The same questions displayed a
-% significant relationship for the entropy statistic: Question 3
-% ($t = 2.93, df = 4, p = 0.022$), Question 6
-% ($t = 2.69, df = 4, p = 0.028$), and Question 7
-% ($t = 2.89, df = 4, p = 0.022$). These relationships are shown in
-% Figure~\ref{fig:stat-response-fit}.
-
-% It is notable that these questions (3, 6, and 7) all pertain to the
-% \emph{perceived effect of the agent} on the performers. As the agent
-% uses the flux measure to trigger new-idea messages, it appears that
-% performers appreciated its intervention more in particularly creative
-% performances where flux and entropy were high.
-
+\pagebreak
 \bibliographystyle{SIGCHI-Reference-Format}

### 2015-09-20 flux-entropy-paper/flux-entropy-paper.tex
chucked out nonperformative data from plots

@@
 
-%   %In this note we describe a data archive for an iPad
-% musical-ensemble performance practice which has included concerts,
-% demonstrations and formal HCI studies of groups of musicians over more
-% than two years. Our corpus contains data for 150 performances,
-% rehearsals, and demonstrations, involving some 100 different musicians
-% and several different iPad apps. At the lowest level, data for each
-% musician in a particular performance was archived as Apple iOS
-% callback methods for touching and swiping. Subsequent data analysis
-% has been undertaken by first classifying these data into higher level
-% gesture types and then describing the transitions between these
-% gestures for each artist as a function of time. We find that a
-% transition-matrix measure, which we call ``flux'', can be used to
-% distinguish different performance types and contexts, as well as
-% different phases (beginning, middle and end) of all of the
-% performances in our corpus.
+  In this note we describe the analysis of a data archive for an iPad
+  musical-ensemble performance practice where transition matrix
+  measures are able to distinguish between sessions of different
+  contexts, and the internal structure of these sessions. The archive
+  includes 95 touch data logs of ensemble touch-screen interactions in
+  concerts, rehearslas and formal HCI studies of groups of musicians
+  over more than two years.
+
+  % At the lowest level, data for each musician in a particular
+  % performance was archived as Apple iOS callback methods for
+  % touching and swiping.
+
+  The data analysis has been undertaken by first classifying
+  these data into higher level gesture types and then describing the
+  transitions between these gestures for each artist as a function of
+  time. We find that a transition-matrix measure, which we call
+  ``flux'', can be used to distinguish different performance types and
+  contexts, as well as different phases (beginning, middle and end) of
+  all of the performances in our corpus.
 
@@
 \category{H.5.5.}{Information Interfaces and Presentation
-  (e.g. HCI)}{Sound and Music Computing}{Systems}
+  (e.g. HCI)}{Sound and Music Computing}{ Systems}
 
@@
 open-ended creative opportunities in the new media arts. Understanding
-and evaluating interfaces in these contexts presents many challenges~\cite{Resnick:2005yu,Shneiderman:2007qv}. In the context of live performance,
-interface evaluation, using, for example, post-performance
-questionnaires, can omit consideration of the dynamic needs of artists
-in the moment of the performance and in the wild of the performance
-context. In contrast to traditional art forms, logged interaction data
-from a new media arts performance can provide an enduring record of
-that performance for post-hoc quantitative investigation as well as
-for curation and preservation~\cite{England:2014ys}. Over
-time, the accumulation of new-media interaction log data might allow
-data-mining techniques to be deployed to potentially describe
-universal features of such interaction.
+and evaluating interfaces in these contexts presents many
+challenges~\cite{Resnick:2005yu,Shneiderman:2007qv}. In the context of
+live performance, interface evaluation, using, for example,
+post-performance questionnaires, can omit consideration of the dynamic
+needs of artists in the moment of the performance and in the wild of
+the performance context. In contrast to traditional art forms, logged
+interaction data from a new media arts performance can provide an
+enduring record of that performance for post-hoc quantitative
+investigation as well as for curation and
+preservation~\cite{England:2014ys}. Over time, the accumulation of
+new-media interaction log data might allow data-mining techniques to
+be deployed to potentially describe universal features of such
+interaction.
 
@@
 instruments -- drums and bowls and bells are often tapped, stroked and
-hit in novel ways to communicate meaning and emotion through music~\cite{Schick:2006fk}.
+hit in novel ways to communicate meaning and emotion through
+music~\cite{Schick:2006fk}.
 
@@
 
-A useful way of modelling user interfaces is to consider their configurations as state vectors and to deploy the methods of matrix algebra to describe transitions between these states during an interaction sequence\cite{Thimbleby:2001kq, Thimbleby:2004fj}.
-Data obtained from usability analysis of quite complex systems (such as a resource planning application for anti-terrorism planning [ref: Kannampallil, Thomas, and Steven Haynes.
-"Using Event Based Markov Model Simulation to Analyze Interactive Human Behavior." AMCIS 2007 Proceedings (2007): 142.
-]) have been undertaken by coding sequences of fine-detail human actions as states and then to model data from usability walkthroughs as sequences of state transitions.
-The transition probably matrix for transitions between states can then be derived from such user data and conclusions drawn about the nature of interaction.
-In the creative arts, editing commands at a computer interface for the ``live coding of computer music'' have also been coded coded as states and analysed using transition matrices, in both the textual and musical dimensions, in order to arrive at conclusions regarding artistic style\cite{Swift:2014tya}.
+A useful way of modelling user interfaces is to consider their
+configurations as state vectors and to deploy the methods of matrix
+algebra to describe transitions between these states during an
+interaction sequence\cite{Thimbleby:2001kq, Thimbleby:2004fj}. Data
+obtained from usability analysis of quite complex systems (such as 
+resource planning software for security applications~\cite{Kannampallil:2007fp})
 
-In our corpus of iPad musical performances, rehearsals and studies, the macro gestures identified in [ref: charlesHCI] provide a basis to identify and code macro gestures.
-Each performer in each muscial piece can then be thought of as following a ``score'' of state transitions and an overall transition matrix for the entire ensemble can be derived by averaging the individual transition matrices.
+have been undertaken by coding sequences of fine-detail human actions
+as states and then to model data from usability walkthroughs as
+sequences of state transitions. The transition probably matrix for
+transitions between states can then be derived from such user data and
+conclusions drawn about the nature of interaction. In the creative
+arts, editing commands at a computer interface for the ``live coding
+of computer music'' have also been coded coded as states and analysed
+using transition matrices, in both the textual and musical dimensions,
+in order to arrive at conclusions regarding artistic
+style\cite{Swift:2014tya}.
+
+In our corpus of iPad musical performances, rehearsals and studies,
+the macro gestures identified by Martin et al~\cite{Martin:2014cr}
+provide a basis to identify and code macro gestures. Each performer in
+each muscial piece can then be thought of as following a ``score'' of
+state transitions and an overall transition matrix for the entire
+ensemble can be derived by averaging the individual transition
+matrices.
 
@@
 
-Previous work showed that the raw touch data obtained from interaction with simple iPad interfaces similar to ours could
-be segmented  using a Random
-Forest\cite{Breiman:2001kx} classifier\footnote{The classifier used
-  was from Python's Scikit-learn package\cite{scikit-learn}.}. The
-classifier was applied to 5-second windows of the touch-data to
+Previous work showed that the raw touch data obtained from interaction
+with simple iPad interfaces similar to ours could be segmented using a
+Random Forest\cite{Breiman:2001kx} classifier\footnote{The classifier
+  used was from Python's Scikit-learn package\cite{scikit-learn}.}.
+The classifier was applied to 5-second windows of the touch-data to
 provide a chain of gesture-states for each participant in each
@@
 instrumental setups? Secondly, do these performance measure allow us
-further insight into the corpus than the metadata can provide? TODO I
-think this second question is too broad?
+further insight into the corpus than the metadata can provide?
+% TODO I think this second question is too broad?
 
@@
 
-\begin{figure}
-  \centering
+\begin{figure} \centering
   \includegraphics[width=\linewidth]{figures/context-flux-entropy-boxplot}
-  \caption{Distributions of flux and entropy values by performance context. Study
-    and rehearsal sessions had the highest flux for improvisations
-    which fell during performances. Compositions had lower flux in
-    rehearsals than performances. Demonstration sessions had much lower entropy
-  indicating data unevenly spread in the transition matrices. This
-  could be because untrained
-  participants did not explore the full gestural potential of the
-  instruments. 
+  \caption{Distributions of flux and entropy values by performance
+context. Study and rehearsal sessions had the highest flux for
+improvisations which fell during performances. Compositions had lower
+flux in rehearsals than performances. Demonstration sessions had much
+lower entropy indicating data unevenly spread in the transition
+matrices. This could be because untrained participants did not explore
+the full gestural potential of the instruments.
 \label{fig:flux-entropy-boxplot}}
 \end{figure}
-
-% \begin{figure}
-%   \centering
-%   \includegraphics[width=\linewidth]{figures/flux-entropy-distribution}
-%   \caption{Distribution of flux and entropy by performance type and
-%     context. Flux and entropy tended to be linearly related in
-%     improvisations, particularly in demonstrations and studies, while
-%     compositions had a more scattered relationship.
-%     \label{fig:flux-entropy-distribution}}
-% \end{figure}
 
@@
 All main effects on flux were found to be significant: performance
-context ($F(2,86) = 6.28, p < 0.01$), performance type ($F(1,86) =
-7.21, p < 0.01$), and instrumentation ($F(1,86) = 20.06, p < 0.001$).
-Significant interaction effects were also found for performance
-context and type ($F(1,86) = 6.18, p < 0.05$), and performance context and
+context ($F(2,86) = 6.28, p < 0.01$), performance type
+($F(1,86) = 7.21, p < 0.01$), and instrumentation
+($F(1,86) = 20.06, p < 0.001$). Significant interaction effects were
+also found for performance context and type
+($F(1,86) = 6.18, p < 0.05$), and performance context and
 instrumentation ($F(1,86) = 10.48, p < 0.01$).
@@
 For entropy, significant main effects were found due to performance
-context ($F(2,86) = 12.10, p < 0.001$), and instrumentation ($F(1,86)
-= 25.86$). The interaction of performance type and instrumentation was
-also found to be significant ($F(1,86) = 9.93, p<0.01$).
+context ($F(2,86) = 12.10, p < 0.001$), and instrumentation
+($F(1,86) = 25.86$). The interaction of performance type and
+instrumentation was also found to be significant
+($F(1,86) = 9.93, p<0.01$).
 
@@
 
-\pagebreak
+%\pagebreak
 \bibliographystyle{SIGCHI-Reference-Format}

### 2015-09-20 R/analysis.R

@@ -36,13 +36,21 @@ valid.sessions.long$time <- strptime(as.character(valid.sessions.long$time), "%Y
 valid.sessions.long <-  subset(valid.sessions.long, performance_type %in% c("composition","improvisation"))
 summary(valid.sessions.long)
 
+perf.sessions.long <- read.csv("../metatone-performance-data-normalmatrices.csv")
+perf.sessions.long <- subset(perf.sessions.long, performance_context %in% c("rehearsal", "performance", "study"))
+perf.sessions.long <- melt(perf.sessions.long,id.vars=c("time","performance_context","performance_type","instruments"),variable.name = "measure",measure.vars = c("flux","entropy"))
+perf.sessions.long$time <- strptime(as.character(perf.sessions.long$time), "%Y-%m-%d %H:%M:%S.%OS")
+perf.sessions.long <-  subset(perf.sessions.long, performance_type %in% c("composition","improvisation"))
+summary(perf.sessions.long)
+
+
 ## Initial Plots
 # cut out data_collection
 #valid.sessions <- (subset(valid.sessions, performance_context != "data_collection"))
 ## plots
 # Count of Session Data
-ggplot(valid.sessions, aes(performance_context)) + geom_bar(aes(fill = performance_type)) + theme(plot.margin=unit(rep(0,4), "cm"), legend.position = "top", legend.box = "horizontal") + scale_fill_manual(values=chifig.2colours) + scale_x_discrete("performance context") + scale_y_continuous("no. of sessions")
-ggsave("../flux-entropy-paper/figures/sessions-count.pdf", width=5.4, height = 2.4)
+ggplot(perf.contexts, aes(performance_context)) + geom_bar(aes(fill = performance_type)) + theme(plot.margin=unit(rep(0,4), "cm"), legend.position = "right", legend.box = "vertical") + scale_fill_manual(values=chifig.2colours) + scale_x_discrete("performance context") + scale_y_continuous("no. of sessions")
+ggsave("../flux-entropy-paper/figures/sessions-count.pdf", width=5.4, height = 2)
 #ggplot(perf.contexts, aes(performance_context)) + geom_bar(aes(fill = performance_type))
 # Flux boxplot
[…]

### 2015-09-21 flux-entropy-paper/flux-entropy-paper.tex
lots of introduction edits

@@
 
-Modern computing devices and interfaces continue to open up rich,
-open-ended creative opportunities in the new media arts. Understanding
-and evaluating interfaces in these contexts presents many
-challenges~\cite{Resnick:2005yu,Shneiderman:2007qv}. In the context of
-live performance, interface evaluation, using, for example,
-post-performance questionnaires, can omit consideration of the dynamic
-needs of artists in the moment of the performance and in the wild of
-the performance context. In contrast to traditional art forms, logged
-interaction data from a new media arts performance can provide an
+Modern computing devices and interfaces continue to open up open-ended
+creative opportunities for new artistic experiences. Understanding and
+evaluating interfaces in these contexts presents many
+challenges~\cite{Resnick:2005yu,Shneiderman:2007qv}.
+
+In the context of live performance, interface evaluation, using, for
+example, post-performance questionnaires, can omit consideration of
+the dynamic needs of artists in the moment of the performance and in
+the wild of the performance context.
+
+In contrast to traditional art forms, logged interaction data from
+interaction with a computer-based artistic interface can provide an
 enduring record of that performance for post-hoc quantitative
 investigation as well as for curation and
-preservation~\cite{England:2014ys}. Over time, the accumulation of
-new-media interaction log data might allow data-mining techniques to
-be deployed to potentially describe universal features of such
-interaction.
+preservation~\cite{England:2014ys}.
+
+Over time, the accumulation of
+interaction log data might allow data-mining techniques to be deployed
+to potentially describe universal features of such interaction.
 
@@
 practitioners exploit simple tools to produce works of beauty whether
-it be static paintings or dynamic performance. In previous
-work~\cite{Martin:2014cr} researchers have shown that trained
-percussionists can exploit a very simple iPad interface to discover
-new categories of musical gestures. In that paper, iPad apps were
-studied where tapping was mapped to produce produced short sounds,
-while swiping produced long sounds with the volume proportional to the
-velocity of the moving touch point. It turned out that trained
-percussionists working together in an ensemble were able to discover
-and repeatedly make use of five macro gestures made up of taps and
-swipes: fast taps, slow taps, random taps, short swipes and swirls.
-This process of discovery of interesting musical gestures from a
-simple interface is analogous to what percussionists do with acoustic
-instruments -- drums and bowls and bells are often tapped, stroked and
-hit in novel ways to communicate meaning and emotion through
-music~\cite{Schick:2006fk}.
+it be static paintings or dynamic performance.
+
+In previous work~\cite{Martin:2014cr} researchers have shown that
+trained percussionists can exploit a very simple iPad interface to
+discover new categories of musical gestures.
+
+In that paper, iPad apps
+were studied where tapping was mapped to produce produced short
+sounds, while swiping produced long sounds with the volume
+proportional to the velocity of the moving touch point.
+
+It turned out that trained percussionists working together in an
+ensemble were able to discover and repeatedly make use of five macro
+gestures made up of taps and swipes: fast taps, slow taps, random
+taps, short swipes and swirls. This process of discovery of
+interesting musical gestures from a simple interface is analogous to
+what percussionists do with acoustic instruments -- drums and bowls
+and bells are often tapped, stroked and hit in novel ways to
+communicate meaning and emotion through music~\cite{Schick:2006fk}.
 
@@
 to support a long-term practice-led research program into simple iPad
-interfaces for group music-making similar to those described in [ref:
-charles at CHI] . Over a period of more than two years (April
-2013--June 2015), we collected data from $....$ group music-making
+interfaces for group music-making similar to those described in Martin
+et al~\cite{Martin:2014cr}. Over a period of more than two years
+(April 2013--June 2015), we collected data from group music-making
 sessions where the participants used custom-built, touch-screen apps
-on Apple iPad devices. Although several different apps have been
-included in our corpus,
-% TODO: should there be a citation here?
-they all shared the simple mode of interaction where the majority of
-the screen could be used for free-form touch interactions; tapping
-produced short sounds, while swiping or swirling produced long sounds
-with the volume proportional to the velocity of the moving touch
-point. Apart from this commonality, our apps featured different
-palettes of available sounds and modes of synthesis, different
-arrangements of sounds on the iPad screens, different kinds of visual
-display and feedback, different kinds of networked interactions
-(between the iPads in an ensemble and with a central server), as well
-as a variety of special features for each app.
+on Apple iPad devices. 
+Although several different apps have been
+included in our corpus, they all shared the simple mode of interaction
+where the majority of the screen could be used for free-form touch
+interactions; tapping produced short sounds, while swiping or swirling
+produced long sounds with the volume proportional to the velocity of
+the moving touch point. Apart from this commonality, our apps featured
+different palettes of available sounds and modes of synthesis,
+different arrangements of sounds on the iPad screens, different kinds
+of visual display and feedback, different kinds of networked
+interactions (between the iPads in an ensemble and with a central
+server), as well as a variety of special features for each app.
 
@@
 
-% TODO: NOTE: FORGET ABOUT THE DEMONSTRATIONS AND THE DATA COLLECTION
-% PARTS OF THE CORPUS. THEY JUST MAKE THE STORY VAGUE. YOU DON'T
-% INCLUDE THEM IN YOUR ANALYSIS AT ALL.
-
-% TODO: NOTE: NEED TO MODIFY ALL OF THESE GRAPHS, TABLES AND FIGURES.
-
+% TODO fix this table to remove descriptive data about the
+% demonstration sessions.
 \begin{table}
-%\centering
+\centering
 \begin{tabular}{l|llll}
@@
 N           & 150       &         &        &  \\
-Record size & 306MB     &         &        &   \\
+%Record size & 306MB     &         &        &   \\
 Length      & 20H34M22S & 1M0S    & 7M29S  & 26M0S\\
@@
 Flux        &           & 0.02239 & 0.1445 & 0.3246\\
+%Entropy     &           &  &  & \\          
 \hline
@@
 algebra to describe transitions between these states during an
-interaction sequence\cite{Thimbleby:2001kq, Thimbleby:2004fj}. Data
-obtained from usability analysis of quite complex systems (such as 
-resource planning software for security applications~\cite{Kannampallil:2007fp})
+interaction sequence~\cite{Thimbleby:2001kq, Thimbleby:2004fj}.
+An opposite approach is to record users' interactions with an
+interface, manually coding interaction states. 
+This method has been applied to usability analysis of resource
+planning software for security applications~\cite{Kannampallil:2007fp}
+where the transition probability matrix of these state sequences
+was used to draw conclusions about the nature of this interactions.
+In the creative arts, editing commands at a computer interface for the
+``live coding'' of computer music has also been coded as states and
+analysed using transition matrices~\cite{Swift:2014tya}, in both the
+textual and musical dimensions, in order to arrive at conclusions
+regarding artistic style.
 
-have been undertaken by coding sequences of fine-detail human actions
-as states and then to model data from usability walkthroughs as
-sequences of state transitions. The transition probably matrix for
-transitions between states can then be derived from such user data and
-conclusions drawn about the nature of interaction. In the creative
-arts, editing commands at a computer interface for the ``live coding
-of computer music'' have also been coded coded as states and analysed
-using transition matrices, in both the textual and musical dimensions,
-in order to arrive at conclusions regarding artistic
-style\cite{Swift:2014tya}.
-
-In our corpus of iPad musical performances, rehearsals and studies,
-the macro gestures identified by Martin et al~\cite{Martin:2014cr}
-provide a basis to identify and code macro gestures. Each performer in
-each muscial piece can then be thought of as following a ``score'' of
-state transitions and an overall transition matrix for the entire
-ensemble can be derived by averaging the individual transition
-matrices.
+In our corpus of iPad musical performances, rehearsals and studies, we
+have used the continuous touch-screen gestures identified by Martin et
+al~\cite{Martin:2014cr} as a basis for a musical state-space.
+Each performer in each musical piece can then be thought of as
+producing a ``score'' of state transitions and an overall transition
+matrix for the entire ensemble can be derived by averaging the
+individual transition matrices.
 
@@
 
-Previous work showed that the raw touch data obtained from interaction
-with simple iPad interfaces similar to ours could be segmented using a
-Random Forest\cite{Breiman:2001kx} classifier\footnote{The classifier
-  used was from Python's Scikit-learn package\cite{scikit-learn}.}.
-The classifier was applied to 5-second windows of the touch-data to
-provide a chain of gesture-states for each participant in each
-session. These chains can be used to create $9 \times 9$ (the
-classifier recognises 9~distinct gestures) transition matrices for
-individual performers and the whole ensemble in each session.
+Previous work~\cite{Martin:2015jk} showed that the raw touch data
+obtained from interaction with simple iPad interfaces similar to ours
+could be segmented using a Random Forest\cite{Breiman:2001kx}
+classifier\footnote{The classifier used was from Python's Scikit-learn
+  package\cite{scikit-learn}.}. The classifier was applied to 5-second
+windows of the touch-data to provide a chain of gesture-states for
+each participant in each session. These chains can be used to create
+$9 \times 9$ (the classifier recognises 9~distinct gestures)
+transition matrices for individual performers and the whole ensemble
+in each session.
 
@@
 distinguish between different performance contexts, types and
-instrumental setups? Secondly, do these performance measure allow us
-further insight into the corpus than the metadata can provide?
-% TODO I think this second question is too broad?
+instrumental setups? Secondly, do these measures allow us further
+insight into the internal structure of performances?
+
+%  Secondly, do these performance measure allow us
+% further insight into the corpus than the metadata can provide?
+% % TODO I think this second question is too broad?
 
@@
 context. Study and rehearsal sessions had the highest flux for
-improvisations which fell during performances. Compositions had lower
-flux in rehearsals than performances. Demonstration sessions had much
-lower entropy indicating data unevenly spread in the transition
-matrices. This could be because untrained participants did not explore
-the full gestural potential of the instruments.
+improvisations. Compositions had lower
+flux in rehearsals than performances.
 \label{fig:flux-entropy-boxplot}}

### 2015-09-21 flux-entropy-paper/flux-entropy-paper.tex
intro updates

@@
 \begin{abstract}
-
-%consider the following? .............
-
-%   %Digital arts practices can allow for the archival of interaction
-% log data. Such log data can capture and preserve performances of
-% new-media artists and musicians and %and can facilitate post-hoc data
-% processing and data mining.
-
-  In this note we describe the analysis of a data archive for an iPad
-  musical-ensemble performance practice where transition matrix
-  measures are able to distinguish between sessions of different
-  contexts, and the internal structure of these sessions. The archive
-  includes 95 touch data logs of ensemble touch-screen interactions in
-  concerts, rehearslas and formal HCI studies of groups of musicians
-  over more than two years.
-
-  % At the lowest level, data for each musician in a particular
-  % performance was archived as Apple iOS callback methods for
-  % touching and swiping.
-
-  The data analysis has been undertaken by first classifying
-  these data into higher level gesture types and then describing the
-  transitions between these gestures for each artist as a function of
-  time. We find that a transition-matrix measure, which we call
-  ``flux'', can be used to distinguish different performance types and
-  contexts, as well as different phases (beginning, middle and end) of
-  all of the performances in our corpus.
-
+  In this note we describe the analysis of a data archive for an
+  ensemble musical performance practice on touch screens where
+  transition matrix measures are able to distinguish between sessions
+  of different contexts, and the internal structure of these sessions.
+  The archive includes 95 touch data logs of ensemble touch-screen
+  interactions in concerts, rehearslas and formal HCI studies of
+  groups of musicians over more than two years.
+  The data analysis has been undertaken by first classifying these
+  data into higher level gesture types and summarising the performers'
+  transitions between these gestures.
+  We find that two transition-matrix measure, flux, and entropy,
+  can be used to distinguish different performance types and contexts,
+  as well as different phases (beginning, middle and end) of all of
+  the performances in our corpus.
 \end{abstract}
 
-\keywords{  Creativity Support Tools;
+\keywords{  
+  Creativity Support Tools;
   Agent; 
@@
 %\section{General comments}
-
 % \begin{itemize}
@@
     in a studio,
-    performance in front of an audience, and demonstration at a
-    research poster session. All of these session were improvised,
-    and in the performance session, the musicians used percussion
-    instruments as well as iPads.
-    \label{fig:three-performance-contexts}}
+    performance with percussion instruments, and performance with just
+    iPads.\label{fig:three-performance-contexts}}
 \end{figure*}
 
-Modern computing devices and interfaces continue to open up open-ended
-creative opportunities for new artistic experiences. Understanding and
-evaluating interfaces in these contexts presents many
-challenges~\cite{Resnick:2005yu,Shneiderman:2007qv}.
+Modern computing devices and interfaces continue to present
+opportunities to support new kinds of creative expression and
+collaborative interaction~\cite{Resnick:2005yu}. However, it has been
+acknowledged that understanding and evaluating interfaces in creative
+contexts can be challenging~\cite{Shneiderman:2007qv}. In the context
+of live performance, interface evaluation using questionnaires, or
+ethnographic data gathering can miss the dynamic need of artists in
+the moment of performance and may be impractical in live performative
+contexts. When performers use a combuter-based artistic interface,
+logged interaction data which has been collected for curation and
+preservation~\cite{England:2014ys}, might also allow quantitative
+investigation into these performances.
 
-In the context of live performance, interface evaluation, using, for
-example, post-performance questionnaires, can omit consideration of
-the dynamic needs of artists in the moment of the performance and in
-the wild of the performance context.
+In this note we describe an archive of interaction data logs that has
+been accumulated to support a long-term practice-led research program
+into simple iPad interfaces for group music-making similar to those
+described in Martin et al~\cite{Martin:2014cr}. Over a period of more
+than two years (April 2013--June 2015), we collected data from group
+music-making sessions where the participants used custom-built,
+touch-screen apps on Apple iPad devices. Although several different
+apps have been included in our corpus, they shared a common mode of
+free-form touch interaction where tapping produced short sounds and
+swirling or swiping produced long sounds. Apart from this commonality,
+the apps featured a variety of sound palettes, different kinds of
+visual feedback, and different kinds of networked interactions
+(between the iPads in an ensemble and with a central server).
 
-In contrast to traditional art forms, logged interaction data from
-interaction with a computer-based artistic interface can provide an
-enduring record of that performance for post-hoc quantitative
-investigation as well as for curation and
-preservation~\cite{England:2014ys}.
+In all of our interaction sessions, every touch interaction---every
+touch down, dragging movement and release---was time-stamped and
+logged to disk for later analysis. These sessions included rehearsals
+without an audience as well as live performances at arts festivals and
+other events. They also included interactions during
+formally-structured HCI experiments where participant compared
+multiple apps and interfaces. Some of the performances were iPad-only
+and others used iPads together with other acoustic percussion
+instruments. Although the majority of the sessions were free-form
+improvisations, a number of rehearsals and performances involved
+composed pieces where the musicians read and performed percussion
+notation using the apps.
 
-Over time, the accumulation of
-interaction log data might allow data-mining techniques to be deployed
-to potentially describe universal features of such interaction.
-
-% Additionally, a wider dissemination of such log data will better
-% support the replication of findings in HCI
-% research~\cite{Wilson:2011ve}
-
-It is a feature of creative expression in the arts that highly skilled
-practitioners exploit simple tools to produce works of beauty whether
-it be static paintings or dynamic performance.
-
-In previous work~\cite{Martin:2014cr} researchers have shown that
-trained percussionists can exploit a very simple iPad interface to
-discover new categories of musical gestures.
-
-In that paper, iPad apps
-were studied where tapping was mapped to produce produced short
-sounds, while swiping produced long sounds with the volume
-proportional to the velocity of the moving touch point.
-
-It turned out that trained percussionists working together in an
-ensemble were able to discover and repeatedly make use of five macro
-gestures made up of taps and swipes: fast taps, slow taps, random
-taps, short swipes and swirls. This process of discovery of
-interesting musical gestures from a simple interface is analogous to
-what percussionists do with acoustic instruments -- drums and bowls
-and bells are often tapped, stroked and hit in novel ways to
-communicate meaning and emotion through music~\cite{Schick:2006fk}.
-
-In this note we describe a log data archive that has been accumulated
-to support a long-term practice-led research program into simple iPad
-interfaces for group music-making similar to those described in Martin
-et al~\cite{Martin:2014cr}. Over a period of more than two years
-(April 2013--June 2015), we collected data from group music-making
-sessions where the participants used custom-built, touch-screen apps
-on Apple iPad devices. 
-Although several different apps have been
-included in our corpus, they all shared the simple mode of interaction
-where the majority of the screen could be used for free-form touch
-interactions; tapping produced short sounds, while swiping or swirling
-produced long sounds with the volume proportional to the velocity of
-the moving touch point. Apart from this commonality, our apps featured
-different palettes of available sounds and modes of synthesis,
-different arrangements of sounds on the iPad screens, different kinds
-of visual display and feedback, different kinds of networked
-interactions (between the iPads in an ensemble and with a central
-server), as well as a variety of special features for each app.
-
-In all of our data-collection sessions, every touch
-interaction---every touch down, dragging movement and touch up---was
-time-stamped and logged to disk for later analysis. These sessions
-included rehearsals without an audience as well as live performances
-at art festivals and other events. They also included two
-formally-structured HCI experiments in order to understand and iterate
-on the system and interface design. Some of the performances were
-iPad-only and others used iPads together with other acoustic
-percussion instruments. Although the majority of the sessions were
-free-form improvisations, a number of rehearsals and performances
-involved composed pieces where the musicians read and interpreted
-percussion notation as they used the apps. Representative photographs
-of some of these session contexts (rehearsal, performance including
-other acoustic instruments and demonstrations) are are shown
+Representative photographs of some of these session contexts
+(rehearsal, performance including other acoustic instruments and iPad
+only performance) are are shown
 Figure~\ref{fig:three-performance-contexts}) and other descriptive
@@
 In our corpus, the median length of sessions is 7m29s and the median
-number of participants per ensemble is four; . Metadata is included
+number of participants per ensemble is four. Metadata was included
 together with the interaction data for each session to record the
@@
 \ref{fig:count-data}.
+
+% It is a feature of creative expression in the arts that highly skilled
+% practitioners exploit simple tools to produce works of beauty whether
+% it be static paintings or dynamic performance. 
+% In previous work~\cite{Martin:2014cr} researchers have shown that
+% trained percussionists can exploit a very simple iPad interface to
+% discover new categories of musical gestures.
+% In that paper, iPad apps
+% were studied where tapping was mapped to produce produced short
+% sounds, while swiping produced long sounds with the volume
+% proportional to the velocity of the moving touch point. 
+% It turned out that trained percussionists working together in an
+% ensemble were able to discover and repeatedly make use of five macro
+% gestures made up of taps and swipes: fast taps, slow taps, random
+% taps, short swipes and swirls. This process of discovery of
+% interesting musical gestures from a simple interface is analogous to
+% what percussionists do with acoustic instruments -- drums and bowls
+% and bells are often tapped, stroked and hit in novel ways to
+% communicate meaning and emotion through music~\cite{Schick:2006fk}.
 
@@
 Flux        &           & 0.02239 & 0.1445 & 0.3246\\
-%Entropy     &           &  &  & \\          
+Entropy     &           &  &  & \\          
 \hline
@@
   interaction sessions used in this study. These sessions were
-  rehearsals, performances, studies, and demonstration sessions and 
+  rehearsals, performances, or studies and 
   participants improvised or performed compositions. Each record
@@
 \end{equation}
-which can be thought of as Markov process. If there is an
-(automated) classification process for extracting each participant's
-state sequence from (or during) their interaction data, then we can
-estimate the $m \times m$ transition matrix $P$ which characterises
-this process by simply counting the number of times state $j$ follows
-state $i$
-\begin{equation}
-  p_{ij} = \frac{N_{ij}}{\sum_i \sum_j N_{ij}}
-\end{equation}
+which can be thought of as Markov process. If there is an (automated)
+classification process for extracting each participant's state
+sequence from (or during) their interaction data, then we can estimate
+the $m \times m$ transition matrix $P$ which characterises this
+process by simply counting the number of times state $j$ follows state
+$i$. The entries of $P$ are then given by:
+$p_{ij} = \frac{N_{ij}}{\sum_i \sum_j N_{ij}}$.
+% \begin{equation}
+%   p_{ij} = \frac{N_{ij}}{\sum_i \sum_j N_{ij}}
+% \end{equation}
 The transition activity of the whole ensemble can be calculated by
@@
 to self transitions (e.g. $A \rightarrow A $). The flux of a TM $P$ is
-given by $\mathrm{flux}(P) = 1 - \mathrm{tr}(P)$ where the trace
+given by: $\mathrm{flux}(P) = 1 - \mathrm{tr}(P)$. Where the trace
 $\mathrm{tr}(P)$ is the sum of the diagonal entries of $P$.
@@
 entropy, defined in the information-theoretic\cite{Shannon:1948rt}
-sense:
-\begin{equation}
-  H(P) = -\sum_{i,j}p_{ij}\log_2(p_{ij})
-\end{equation}
+sense: $H(P) = -\sum_{i,j}p_{ij}\log_2(p_{ij})$.
+% \begin{equation}
+%   H(P) = -\sum_{i,j}p_{ij}\log_2(p_{ij})
+% \end{equation}
 This entropy measure is small when the matrix is sparse, and largest
@@
 insight into the internal structure of performances?
-
-%  Secondly, do these performance measure allow us
-% further insight into the corpus than the metadata can provide?
-% % TODO I think this second question is too broad?
 
@@
 significantly predict the entropy and flux values.
-% TODO we haven't explained what ``the 95 performative sessions'' are,
-% have we?
-In the following analysis, only the 95~``performance context''
-sessions were considered, excluding the demonstration and
-data-collection contexts. TODO can we justify this?
 
@@
 
+% Additionally, a wider dissemination of such log data will better
+% support the replication of findings in HCI
+% research~\cite{Wilson:2011ve}
+
 %\pagebreak

### 2015-09-21 R/analysis.R
updates to calculate descriptive data more cleverly

@@ -17,17 +17,20 @@ flux.entropy.ratings <- read.csv("../flux_entropy_ratings.csv")
 #summary(raw.performance.data)
 summary(valid.sessions)
 ## rehearsals/performances/studies only
-perf.contexts <- subset(raw.performance.data, performance_context %in% c("rehearsal", "performance", "study"))
-perf.contexts <- subset(perf.contexts, performance_type %in% c("composition","improvisation"))
-summary(perf.contexts$performance_type) # Composition: 23 - Improvisation: 72
-summary(perf.contexts)
-                                        # Calculate total recorded time.
-total.seconds <- max(valid.sessions$length_seconds)
-message(paste(total.seconds %/% 3600, "H", (total.seconds %% 3600) %/% 60, "M", signif(total.seconds %% 60, 4), "S", sep = "" ))
-                                        # performer data
-message(paste(sum(valid.sessions$number_performers),min(valid.sessions$number_performers),median(valid.sessions$number_performers),max(valid.sessions$number_performers),sep = " & "))
-                                        #
-message(paste(signif(sum(valid.sessions$flux),4),signif(min(valid.sessions$flux),4),signif(median(valid.sessions$flux),4),signif(max(valid.sessions$flux),4),sep = " & "))
+perf.sessions <- subset(raw.performance.data, performance_context %in% c("rehearsal", "performance", "study"))
+perf.sessions <- subset(perf.sessions, performance_type %in% c("composition","improvisation"))
+summary(perf.sessions$performance_type) # Composition: 23 - Improvisation: 72
+summary(perf.sessions)
+
+format.seconds <- function(seconds){
+return(paste(seconds %/% 3600, "H", (seconds %% 3600) %/% 60, "M", signif(seconds %% 60, 4), "S", sep = ""))
+}
+                                        # length data - total min median max
+lengths <- perf.sessions$length_seconds
+message(paste("Length",format.seconds(sum(lengths)),format.seconds(min(lengths)),format.seconds(median(lengths)),format.seconds(max(lengths)),sep = " & "))
+message(paste("Participants",sum(perf.sessions$number_performers),min(perf.sessions$number_performers),median(perf.sessions$number_performers),max(perf.sessions$number_performers),sep = " & "))
[…]

### 2015-09-21 flux-entropy-paper/flux-entropy-paper.tex
new descriptive data table
update length median
intro fixed up

@@
   of different contexts, and the internal structure of these sessions.
-  The archive includes 95 touch data logs of ensemble touch-screen
-  interactions in concerts, rehearslas and formal HCI studies of
-  groups of musicians over more than two years.
-  The data analysis has been undertaken by first classifying these
-  data into higher level gesture types and summarising the performers'
-  transitions between these gestures.
-  We find that two transition-matrix measure, flux, and entropy,
-  can be used to distinguish different performance types and contexts,
-  as well as different phases (beginning, middle and end) of all of
-  the performances in our corpus.
+  Traditional methodologies for studying interfaces that support
+  collaborative creativity can be time-consuming and intrusive to
+  implement but our methods may provide a practical alternative for
+  large scale analysis of creative interactions.
+
+  Our archive includes 95 touch data logs of ensemble touch-screen
+  interactions in concerts, rehearsals and formal HCI studies of
+  groups of musicians over more than two years. The data analysis has
+  been undertaken by first classifying these data into higher level
+  gesture types and summarising the performers' transitions between
+  these gestures. We find that two transition-matrix measure, flux,
+  and entropy, can be used to distinguish different performance types
+  and contexts, as well as the internal structure (beginning, middle
+  and end) of the performances in our corpus.
 \end{abstract}
@@
 of live performance, interface evaluation using questionnaires, or
-ethnographic data gathering can miss the dynamic need of artists in
-the moment of performance and may be impractical in live performative
-contexts. When performers use a combuter-based artistic interface,
-logged interaction data which has been collected for curation and
-preservation~\cite{England:2014ys}, might also allow quantitative
-investigation into these performances.
+ethnographic data gathering can miss the dynamic needs of artists
+during collaborative interactions and may be impractical in live
+performative contexts. When performers use a combuter-based artistic
+interface, logged interaction data which has been collected for
+curation and preservation~\cite{England:2014ys}, might also allow
+quantitative investigation into these performances. In this note we
+will show how an archive of collaborative performance data logs can be
+subjected to a quantitative analysis based on transition probabilities
+between gestural states. We will demonstrate that two measures on
+transition matrices, flux and entropy, show significant differences
+between different kinds of performances, and give insight into the
+internal structure of these performances.
 
-In this note we describe an archive of interaction data logs that has
-been accumulated to support a long-term practice-led research program
-into simple iPad interfaces for group music-making similar to those
-described in Martin et al~\cite{Martin:2014cr}. Over a period of more
-than two years (April 2013--June 2015), we collected data from group
-music-making sessions where the participants used custom-built,
-touch-screen apps on Apple iPad devices. Although several different
-apps have been included in our corpus, they shared a common mode of
-free-form touch interaction where tapping produced short sounds and
-swirling or swiping produced long sounds. Apart from this commonality,
-the apps featured a variety of sound palettes, different kinds of
-visual feedback, and different kinds of networked interactions
-(between the iPads in an ensemble and with a central server).
+Our archive of interaction data logs has been accumulated to support a
+long-term practice-led research program into simple iPad interfaces
+for group music-making similar to those described in Martin et
+al~\cite{Martin:2014cr}. Over a period of more than two years (April
+2013--June 2015), we collected data from group music-making sessions
+where the participants used our custom-built, touch-screen apps on
+Apple iPad devices. Although several different apps have been included
+in our corpus, they shared a common mode of free-form touch
+interaction where tapping produced short sounds and swirling or
+swiping produced long sounds. Apart from this commonality, the apps
+featured a variety of sound palettes, different kinds of visual
+feedback, and different kinds of networked interactions (between the
+iPads in an ensemble and with a central server).
 
@@
 touch down, dragging movement and release---was time-stamped and
-logged to disk for later analysis. These sessions included rehearsals
-without an audience as well as live performances at arts festivals and
-other events. They also included interactions during
-formally-structured HCI experiments where participant compared
-multiple apps and interfaces. Some of the performances were iPad-only
-and others used iPads together with other acoustic percussion
-instruments. Although the majority of the sessions were free-form
-improvisations, a number of rehearsals and performances involved
-composed pieces where the musicians read and performed percussion
-notation using the apps.
+logged to disk for later analysis. These sessions included
+formally-structured HCI experiments, where participants were
+interviewed or surveyed. They also included rehearsals without an
+audience and performances at arts festivals where creative goals were
+paramount; in these situations surveys or interviews would have been
+inappropriate and possibly disruptive. Some of the performances were
+iPad-only and others used iPads together with other acoustic
+percussion instruments. Although the majority of the sessions were
+free-form improvisations, a number of rehearsals and performances
+involved composed pieces where the musicians read and performed
+percussion notation using the apps.
 
-Representative photographs of some of these session contexts
-(rehearsal, performance including other acoustic instruments and iPad
-only performance) are are shown
-Figure~\ref{fig:three-performance-contexts}) and other descriptive
-statistics about the sessions are shown in Table \ref{corpus-table}.
-In our corpus, the median length of sessions is 7m29s and the median
-number of participants per ensemble is four. Metadata was included
-together with the interaction data for each session to record the
+Representative photographs of some of these session contexts are are
+shown Figure~\ref{fig:three-performance-contexts}) and other
+descriptive statistics about the sessions are shown in Table
+\ref{corpus-table}. In our corpus, the median length of sessions is
+7m26s and the median number of participants per ensemble is four. Each
+session record is labelled with metadata about the session
 \emph{context} (performance, rehearsal, study), session \emph{type}
@@
 
-% TODO fix this table to remove descriptive data about the
-% demonstration sessions.
 \begin{table}
@@
 \hline
-N           & 150       &         &        &  \\
+N           & 95    &      &          &         \\
 %Record size & 306MB     &         &        &   \\
-Length      & 20H34M22S & 1M0S    & 7M29S  & 26M0S\\
-Participants& 524       & 1       & 4      & 9    \\
+Length & 13H9M19S & 1M19S & 7M26S & 22M20S \\
+Participants& 353   & 2    & 4        & 9        \\
 Flux &   & 0.02239 & 0.1445 & 0.3246\\
-Entropy     &           &  &  & \\          
+Entropy &   & 0.7683 & 3.342 & 4.442\\          
 \hline
@@
 algebra to describe transitions between these states during an
-interaction sequence~\cite{Thimbleby:2001kq, Thimbleby:2004fj}.
-An opposite approach is to record users' interactions with an
-interface, manually coding interaction states. 
-This method has been applied to usability analysis of resource
-planning software for security applications~\cite{Kannampallil:2007fp}
-where the transition probability matrix of these state sequences
-was used to draw conclusions about the nature of this interactions.
-In the creative arts, editing commands at a computer interface for the
-``live coding'' of computer music has also been coded as states and
-analysed using transition matrices~\cite{Swift:2014tya}, in both the
-textual and musical dimensions, in order to arrive at conclusions
-regarding artistic style.
+interaction sequence~\cite{Thimbleby:2001kq, Thimbleby:2004fj}. An
+inverted approach is to record users' interactions with an interface,
+manually coding interaction states  This method has been applied to
+usability analysis of resource planning software for security
+applications~\cite{Kannampallil:2007fp} where the transition
+probability matrix of these state sequences was used to draw
+conclusions about the nature of this interactions. In the creative
+arts, editing commands at a computer interface for the ``live coding''
+of computer music has also been coded as states and analysed using
+transition matrices~\cite{Swift:2014tya}, in both the textual and
+musical dimensions, in order to arrive at conclusions regarding
+artistic style.
 

### 2015-09-21 flux-entropy-paper/flux-entropy-paper.tex
better macro gestures section

@@
 A useful way of modelling user interfaces is to consider their
-configurations as state vectors and to deploy the methods of matrix
-algebra to describe transitions between these states during an
-interaction sequence~\cite{Thimbleby:2001kq, Thimbleby:2004fj}. An
-inverted approach is to record users' interactions with an interface,
-manually coding interaction states  This method has been applied to
-usability analysis of resource planning software for security
-applications~\cite{Kannampallil:2007fp} where the transition
+theoretical configurations as state vectors and to deploy the methods
+of matrix algebra to describe transitions between these states during
+an interaction sequence~\cite{Thimbleby:2001kq, Thimbleby:2004fj}. An
+inverted approach is to record users' interactions and manually code
+interaction states to produce a transition model. This method has been
+applied to usability analysis of resource planning software for
+security applications~\cite{Kannampallil:2007fp} where the transition
 probability matrix of these state sequences was used to draw
 conclusions about the nature of this interactions. In the creative
-arts, editing commands at a computer interface for the ``live coding''
-of computer music has also been coded as states and analysed using
+arts, editing commands at a computer interface to ``live code''
+computer music has also been coded as states and analysed using
 transition matrices~\cite{Swift:2014tya}, in both the textual and
@@
 In our corpus of iPad musical performances, rehearsals and studies, we
-have used the continuous touch-screen gestures identified by Martin et
-al~\cite{Martin:2014cr} as a basis for a musical state-space.
-Each performer in each musical piece can then be thought of as
-producing a ``score'' of state transitions and an overall transition
-matrix for the entire ensemble can be derived by averaging the
-individual transition matrices.
+have used an automated method to classify logs of touch interaction,
+at a regular interval, into nine continuous touch-screen gestures, as
+described by Martin et al~\cite{Martin:2015jk}. Each performer in each
+musical piece can then be thought of as producing a ``score'' of state
+transitions and an overall transition matrix for the entire ensemble
+can be derived by averaging the individual transition matrices.
 
@@
 participant $X$ generates a length $N$ sequence of these interaction
-states
-\begin{equation}
- X_n \hskip 2em n = 1, \ldots, N
-\end{equation}
-which can be thought of as Markov process. If there is an (automated)
-classification process for extracting each participant's state
-sequence from (or during) their interaction data, then we can estimate
-the $m \times m$ transition matrix $P$ which characterises this
-process by simply counting the number of times state $j$ follows state
-$i$. The entries of $P$ are then given by:
-$p_{ij} = \frac{N_{ij}}{\sum_i \sum_j N_{ij}}$.
-% \begin{equation}
-%   p_{ij} = \frac{N_{ij}}{\sum_i \sum_j N_{ij}}
-% \end{equation}
-The transition activity of the whole ensemble can be calculated by
-averaging the TM for each performer. The matrix can be normalised so
-that the sum of all elements is equal to one.
+states $X_n$ ($n = 1, \ldots, N$) which can be thought of as Markov
+process.
 
-Previous work~\cite{Martin:2015jk} showed that the raw touch data
-obtained from interaction with simple iPad interfaces similar to ours
-could be segmented using a Random Forest\cite{Breiman:2001kx}
-classifier\footnote{The classifier used was from Python's Scikit-learn
-  package\cite{scikit-learn}.}. The classifier was applied to 5-second
-windows of the touch-data to provide a chain of gesture-states for
-each participant in each session. These chains can be used to create
+If there is an (automated) classification process for identifying each
+participant's state sequence from their interaction data, then we can
+estimate the $m \times m$ transition matrix $P$ which characterises
+this process by setting each entry $p_{ij}$ to be the number of times
+state $j$ follows state $i$ in the sequence. The transition activity
+of the whole ensemble can be calculated by averaging the TM for each
+performer. The matrix can be normalised so that the sum of all
+elements is equal to one.
+
+We applied a Random Forest~\cite{Breiman:2001kx}
+classifier\footnote{The Random Forest classifier used was from
+  Python's Scikit-learn package\cite{scikit-learn}.} to 5-second
+windows of touch-data in our corpus to provide chains of
+gesture-states for each participant in each session following previous
+work~\cite{Martin:2015jk}. These chains can be used to create
 $9 \times 9$ (the classifier recognises 9~distinct gestures)

### 2015-09-21 flux-entropy-paper/flux-entropy-paper.tex
a few more macro gesture fixes

@@
 
-In our corpus of iPad musical performances, rehearsals and studies, we
-have used an automated method to classify logs of touch interaction,
-at a regular interval, into nine continuous touch-screen gestures, as
-described by Martin et al~\cite{Martin:2015jk}. Each performer in each
-musical piece can then be thought of as producing a ``score'' of state
-transitions and an overall transition matrix for the entire ensemble
-can be derived by averaging the individual transition matrices.
+In our corpus of iPad musical performances, logs of low-level touch
+data record each performers micro-gestures. However, it has been
+established that musical performances on touch-screens can be seen in
+terms of vocabularies of continuous macro gestures that unfold over
+several seconds of touch interaction~\cite{Martin:2014cr}. Improvised
+performances can be then thought of as a sequence of such
+states~\cite{Pressing:1988uo} for each performer.
 
-Suppose that each participant's activity in a collaborative
-interaction can at any given time be classified as belonging to one of
-$m$ distinct \emph{states}. Over the course of the activity, each
+If the performers' touch interactions can be classified as belonging
+to one of $m$ distinct states, then over the course of a session each
 participant $X$ generates a length $N$ sequence of these interaction
@@
 
-We applied a Random Forest~\cite{Breiman:2001kx}
+To generate state sequences for our corpus we have used an automated
+method to classify logs of touch interaction, at a regular interval,
+into nine continuous touch-screen gestures, as described by Martin et
+al~\cite{Martin:2015jk}. A Random Forest~\cite{Breiman:2001kx}
 classifier\footnote{The Random Forest classifier used was from
-  Python's Scikit-learn package\cite{scikit-learn}.} to 5-second
-windows of touch-data in our corpus to provide chains of
-gesture-states for each participant in each session following previous
-work~\cite{Martin:2015jk}. These chains can be used to create
-$9 \times 9$ (the classifier recognises 9~distinct gestures)
-transition matrices for individual performers and the whole ensemble
-in each session.
+  Python's Scikit-learn package\cite{scikit-learn}.} was applied to
+5-second windows of touch-data in our corpus to provide chains of
+gesture-states for each participant in each session. These chains can
+be used to create $9 \times 9$ transition matrices for individual
+performers and the whole ensemble in each session.
 
@@
 interface and method of recording the data allows all the sessions to
-be transformed into gesture sequences. These transition matrices have
-been used as a visualisation technique to reveal qualitative
-differences in structure in improvisational
+be transformed into gesture sequences, and transition matrices.
+Transition matrices have been used as a visualisation technique to
+reveal qualitative differences in structure in improvisational
 music-making\cite{Swift:2014tya}. In this paper, however, we wish to
-extract quantitative statistics from these TMs for comparing and
+extract quantitative measures from these TMs for comparing and
 tracking different sessions over time.

### 2015-09-21 flux-entropy-paper/flux-entropy-paper.tex
up to conclusion

@@
 
-In all of our interaction sessions, every touch interaction---every
+In all of our 95 interaction sessions, every touch interaction---every
 touch down, dragging movement and release---was time-stamped and
@@
 interface and method of recording the data allows all the sessions to
-be transformed into gesture sequences, and transition matrices.
-Transition matrices have been used as a visualisation technique to
-reveal qualitative differences in structure in improvisational
-music-making\cite{Swift:2014tya}. In this paper, however, we wish to
-extract quantitative measures from these TMs for comparing and
-tracking different sessions over time.
+be transformed into gesture sequences, and transition matrices. While
+these matrices can be used to visualise the structure of individual
+performances~\cite{Swift:2014tya}, we wish to extract quantitative
+measures from these TMs to compare the large number of sessions in our
+corpus.
 
@@
 to self transitions (e.g. $A \rightarrow A $). The flux of a TM $P$ is
-given by: $\mathrm{flux}(P) = 1 - \mathrm{tr}(P)$. Where the trace
+given by: $\mathrm{flux}(P) = 1 - \mathrm{tr}(P)$ where the trace
 $\mathrm{tr}(P)$ is the sum of the diagonal entries of $P$.
@@
 entropy, defined in the information-theoretic\cite{Shannon:1948rt}
-sense: $H(P) = -\sum_{i,j}p_{ij}\log_2(p_{ij})$.
-% \begin{equation}
-%   H(P) = -\sum_{i,j}p_{ij}\log_2(p_{ij})
-% \end{equation}
-This entropy measure is small when the matrix is sparse, and largest
-when each matrix element is equal. It offers a different perspective
-on collaborative interactions than the flux measure by capturing the
-breadth of the gestural space explored by the participants throughout
-the course of an interaction. Consider the degenerate case of a
-participant who alternates between two states for a whole performance:
-the flux in this case will be maximal ($\mathrm{flux}(P) = 1$) since
-there are no self-transitions, (only $A \rightarrow B$ or
-$ B \rightarrow A$) even though the participant has only used a small
-subset of the state space. The entropy of this interaction, on the
-other hand, will be low. Entropy, therefore, is a measure of how broad
-the participant's exploration of the state space is in a given
-interaction.
+sense: $H(P) = -\sum_{i,j}p_{ij}\log_2(p_{ij})$. This measure
+is small when the matrix is sparse, and largest when each matrix
+element is equal. It offers a different perspective on collaborative
+interactions than the flux measure by capturing the breadth of the
+gestural space explored by the participants throughout the course of
+an interaction. Consider the degenerate case of a participant who
+alternates between two states for a whole performance: the flux in
+this case will be maximal ($\mathrm{flux}(P) = 1$) since there are no
+self-transitions, (only $A \rightarrow B$ or $ B \rightarrow A$) even
+though the participant has only used a small subset of the state
+space. The entropy of this interaction, on the other hand, will be
+low. Entropy, therefore, is a measure of how broad the participant's
+exploration of the state space is in a given interaction.
 
-In the rest of this paper we will use these matrix measures to answer
+In the rest of this note we will use these matrix measures to answer
 two questions. First, do the transition matrix measures allow us to
@@
 \ref{fig:flux-entropy-boxplot}. A three-way ANOVA procedure was
-performed on the data set to find whether the metatdata could
+performed on the data set to find whether the metadata could
 significantly predict the entropy and flux values.
@@
 used to understand some of the internal structure of performances? A
-simple model for understanding temporal art-forms is the ``beginning,
-middle, end'' structure. To investigate how this structure fits our
-corpus, we can divide the gestural-sequences of each session into
-thirds and calculate the transition matrices given by each section.
+classical model for understanding temporal art-forms is the
+``beginning, middle, end'' structure~\cite{Aristotle:350rt}. To
+investigate how this structure fits our corpus, we can divide the
+gestural-sequences of each session into thirds and calculate the
+transition matrices given by each section.
 
@@
 than for improvisations. The stability of compositions is to be
-expected since performers are expected to use the same gestures in
-each rehearsal and performance of a composition, it may be that
-different kinds of compositions could result in different structural
+expected since performers use the same gestures in each rehearsal and
+performance of these written scores, however it may be that different
+kinds of compositions could result in different structural
 differences. Both measures have a downward trend through sections in
@@
 
-% todo add the plots here.
-
-% \begin{figure}
-%   \centering
-%   \includegraphics[width=\linewidth]{figures/type-section-flux}
-%   \includegraphics[width=\linewidth]{figures/type-section-entropy}
-%   \caption{The flux of compositions and improvisations divided into
-%     beginning, middle, and end. Composition flux remains stable throughout,
-%     but improvisations show a downward trend and composition entropy tends to move up
-%     at the end, while flux moves down.
-%     \label{fig:type-section-flux-entropy}}
-% \end{figure}
-
-% \begin{figure}
-%   \centering
-%   \includegraphics[width=\linewidth]{figures/context-section-flux}
-%   \includegraphics[width=\linewidth]{figures/context-section-entropy}
-%   \caption{The distribution of flux and entropy in performances divided by ``beginnings,
-%     middles, and ends''.
-%     \label{fig:context-section-flux-entropy}}
-% \end{figure}
-
 \subsection{Performance styles over time}
@@
 Over the course of this research program, different ensembles and
-different activities have been emphasised, some of these differences
-can be seen as clusters of performances with similar flux and entropy
-in the two plots. For instance, the earliest series of rehearsals and
+activities have been emphasised and some of these differences can be
+seen as clusters of performances with similar flux and entropy in the
+two plots. For instance, the earliest series of rehearsals and
 performances had flux between 0.05 and 0.15, while the period of
@@
 
-In this paper we have extended our previous work on transition
+In this note we have extended previous work on transition
 matrices with two simple matrix measures, flux and entropy, both of
 which have been demonstrated to have significant discriminatory power
-with regards to a corpus of 150 collaborative touch-screen
+with regards to a corpus of 95 collaborative touch-screen
 interactions. Not only do these measures differentiate between

### 2015-09-21 flux-entropy-paper/flux-entropy-paper.tex
updates to conclusion

@@
 
-In this note we have extended previous work on transition
-matrices with two simple matrix measures, flux and entropy, both of
-which have been demonstrated to have significant discriminatory power
-with regards to a corpus of 95 collaborative touch-screen
-interactions. Not only do these measures differentiate between
-different types of sessions and the internal structure of these
-sessions, but they have explanatory power that has thrown new light on
-the musical interactions under examination. HCI researchers are
-increasingly concerned with interactions that happen in-the-wild,
-where the interactions of many users may be collected without the
-opportunity to survey or interview users. We suggest that analysing
-interaction-state transitions using matrices and simple measures can
-lead to new insights into HCI interactions outside the lab.
+In this note we have extended previous work using transition matrices
+to understand HCI processes with the application of two simple matrix
+measures, flux and entropy, both of which have been demonstrated to
+have significant discriminatory power with regards to a corpus of 95
+collaborative touch-screen interactions. Not only do these measures
+differentiate between different types of sessions and the internal
+structure of these sessions, but they have explanatory power that has
+thrown new light on the musical interactions under examination. It is
+notable that, for improvised performances, live performances had the
+lowest measure in both flux and entropy, while lab-studies had the
+highest. This may be due to the increased pressure on performers when
+improvising in front of an audience. The internal investigation of
+performances session revealed more structural variety in
+improvisations than compositions. In response, new works could be
+composed that specifically explore different gestural structures in
+touch-screen performance
 
-%TODO Mention Manovich transmutation stuff
-The macro-gestures that these processes reveal could also be used as
-the basis for new works of new media art~\cite{Manovich:2002ly}.
+HCI researchers are increasingly concerned with interactions that
+happen in-the-wild, where the creative collaborations of many users
+may be recorded, but where conducting surveys or interviews would be
+innapropriate. We suggest that analysing interaction-state transitions
+using matrices and simple measures can lead to new insights into HCI
+interactions outside the lab.
 
-% Additionally, a wider dissemination of such log data will better
-% support the replication of findings in HCI
-% research~\cite{Wilson:2011ve}
+Data-mining logs of interactions with creative interfaces sugggest
+further approaches for future work. The macro-gestures that these
+processes reveal could be used as the basis for new works of new media
+art~\cite{Manovich:2002ly}. Additionally, a wider dissemination of
+such log data could better support the replication of findings in HCI
+research~\cite{Wilson:2011ve}
 

### 2015-09-22 flux-entropy-paper/flux-entropy-paper.tex
Added apostrophe
I added an apostrophe to show that I can use github! It took me ages to
find the commit button! I am still having troubles running latex; the
github directory is not self contained; you guys have different latex
configurations to me.
finished more Ben edits

@@
 \begin{abstract}
-  In this note we describe the analysis of a data archive for an
-  ensemble musical performance practice on touch screens where
-  transition matrix measures are able to distinguish between sessions
-  of different contexts, and the internal structure of these sessions.
-  Traditional methodologies for studying interfaces that support
-  collaborative creativity can be time-consuming and intrusive to
-  implement but our methods may provide a practical alternative for
-  large scale analysis of creative interactions.
 
-  Our archive includes 95 touch data logs of ensemble touch-screen
-  interactions in concerts, rehearsals and formal HCI studies of
-  groups of musicians over more than two years. The data analysis has
-  been undertaken by first classifying these data into higher level
-  gesture types and summarising the performers' transitions between
-  these gestures. We find that two transition-matrix measure, flux,
-  and entropy, can be used to distinguish different performance types
-  and contexts, as well as the internal structure (beginning, middle
-  and end) of the performances in our corpus.
+  In this note we describe the analysis of an interaction data archive
+  associated with a long-term practice-led research program
+  investigating ensemble musical performance practice on touch
+  screens. We use transition matrix measures to distinguish between
+  sessions of different contexts, and the internal structure of these
+  sessions. The data analysis proceeds by first classifying these data
+  into higher level gesture types and summarising the performers'
+  transitions between these gestures. We find that two
+  transition-matrix measure, flux, and entropy, can be used to
+  distinguish different performance types and contexts, as well as the
+  internal structure of the performances in our corpus.
 \end{abstract}
@@
   Creativity Support Tools;
-  Agent; 
-  Design;
+  Collaborative Interaction;
   Methodology}
@@
 \category{H.5.5.}{Information Interfaces and Presentation
-  (e.g. HCI)}{Sound and Music Computing}{ Systems}
+  (e.g. HCI)}{Sound and Music Computing}{Systems}
 
@@
 acknowledged that understanding and evaluating interfaces in creative
-contexts can be challenging~\cite{Shneiderman:2007qv}. In the context
-of live performance, interface evaluation using questionnaires, or
-ethnographic data gathering can miss the dynamic needs of artists
-during collaborative interactions and may be impractical in live
-performative contexts. When performers use a combuter-based artistic
-interface, logged interaction data which has been collected for
-curation and preservation~\cite{England:2014ys}, might also allow
-quantitative investigation into these performances. In this note we
-will show how an archive of collaborative performance data logs can be
-subjected to a quantitative analysis based on transition probabilities
-between gestural states. We will demonstrate that two measures on
-transition matrices, flux and entropy, show significant differences
-between different kinds of performances, and give insight into the
-internal structure of these performances.
+contexts can be challenging~\cite{Shneiderman:2007qv}. This problem is
+even more pronounced in artistic \emph{performance}, where subjective
+interface evaluation through questionnaires or interviews may be both
+practically difficult and culturally inappropriate. When the interface
+is digital, however, it is still usually trivial to log interaction in
+these situations, which is useful for curation and
+preservation~\cite{England:2014ys}. The question is: can this log data
+be used not just as a record for posterity, but in support of a
+broader creative-practice led research program?
+
+In this note we will show how an archive of collaborative performance
+data logs can be subjected to a quantitative analysis based on
+transition probabilities between gestural states. We will demonstrate
+that two measures on transition matrices, flux and entropy, show
+significant differences between different kinds of performances, and
+give insight into the internal structure of these performances.
 
@@
 interaction where tapping produced short sounds and swirling or
-swiping produced long sounds. Apart from this commonality, the apps
+swiping produced long sounds. Apart from these similarities, the apps
 featured a variety of sound palettes, different kinds of visual
@@
 
-Representative photographs of some of these session contexts are are
-shown Figure~\ref{fig:three-performance-contexts}) and other
-descriptive statistics about the sessions are shown in Table
-\ref{corpus-table}. In our corpus, the median length of sessions is
-7m26s and the median number of participants per ensemble is four. Each
-session record is labelled with metadata about the session
-\emph{context} (performance, rehearsal, study), session \emph{type}
-(improvisation or composition), and \emph{instrumentation}
-(iPads-only, or iPads with acoustic percussion). The distribution of
-recorded sessions by context and type is shown in Figure
-\ref{fig:count-data}.
+Some of these session contexts are are shown
+Figure~\ref{fig:three-performance-contexts}) and other descriptive
+statistics about the sessions are shown in Table \ref{corpus-table}.
+In our corpus, the median length of sessions is 7m26s and the median
+number of participants per ensemble is four. Each session record is
+labelled with metadata about the session \emph{context} (performance,
+rehearsal, study), session \emph{type} (improvisation or composition),
+and \emph{instrumentation} (iPads-only, or iPads with acoustic
+percussion). The distribution of recorded sessions by context and type
+is shown in Figure \ref{fig:count-data}.
 
@@
 In our corpus of iPad musical performances, logs of low-level touch
-data record each performers micro-gestures. However, it has been
+data record each performer's micro-gestures. However, it has been
 established that musical performances on touch-screens can be seen in
@@
 differentiate between different types of sessions and the internal
-structure of these sessions, but they have explanatory power that has
-thrown new light on the musical interactions under examination. It is
-notable that, for improvised performances, live performances had the
-lowest measure in both flux and entropy, while lab-studies had the
-highest. This may be due to the increased pressure on performers when
-improvising in front of an audience. The internal investigation of
-performances session revealed more structural variety in
+structure of these sessions, they provide qualitative insight into the
+performances as well. For improvised performances, live performances
+had the lowest measure in both flux and entropy, while lab-studies had
+the highest. This may be due to the increased pressure on performers
+when improvising in front of an audience. The internal investigation
+of performances session revealed more structural variety in
 improvisations than compositions. In response, new works could be
 composed that specifically explore different gestural structures in
-touch-screen performance
+touch-screen performance.
 
-HCI researchers are increasingly concerned with interactions that
-happen in-the-wild, where the creative collaborations of many users
-may be recorded, but where conducting surveys or interviews would be
-innapropriate. We suggest that analysing interaction-state transitions
-using matrices and simple measures can lead to new insights into HCI
-interactions outside the lab.
-
-Data-mining logs of interactions with creative interfaces sugggest
+Data-mining logs of interactions with creative interfaces suggest
 further approaches for future work. The macro-gestures that these
@@
 such log data could better support the replication of findings in HCI
-research~\cite{Wilson:2011ve}
+research~\cite{Wilson:2011ve}. Finally, flux and entropy proved useful
+in the analysis of our data corpus, but other transition matrix
+metrics may be more suitable for different interaction data.
+
+HCI researchers are increasingly concerned with interactions that
+happen in-the-wild, where the creative collaborations of many users
+may be recorded, but where conducting surveys or interviews is
+innapropriate. We suggest that analysing interaction-state transitions
+using matrices and simple measures can lead to new insights into HCI
+interactions outside the lab.
 

### 2015-09-23 flux-entropy-paper/flux-entropy-paper.tex
changed ACM category
not sure how to deal with the sub sub category descriptor…
revised classification keywords again
mystery error is caused by references: hyper refs going over page
boundaries.
major edit from Ben's version
note terminology for context: rehearsal, recording, concert; many other
changes; hopefully getting closer

@@
 
-\pagenumbering{arabic}
+%\pagenumbering{arabic}
 
@@
 
-\title{Mining Interaction Log Data in a Creative Arts Practice using Macro Gestures and Transition Matrices}
+\title{Mining Interaction Log Data in a Creative Arts Practice using Transition Matrices}
 
@@
     \email{e-mail address}}\\
+     \alignauthor{4th Author Name\\
+    \affaddr{Affiliation}\\
+    \affaddr{City, Country}\\
+    \email{e-mail address}}\\
 }
@@
 
-  In this note we describe the analysis of an interaction data archive
+  In this note we present an analysis of an interaction data archive
   associated with a long-term practice-led research program
@@
   transitions between these gestures. We find that two
-  transition-matrix measure, flux, and entropy, can be used to
+  transition-matrix measures, flux, and entropy, can be used to
   distinguish different performance types and contexts, as well as the
@@
 
-\category{H.5.5.}{Information Interfaces and Presentation
-  (e.g. HCI)}{Sound and Music Computing}{Systems}
-
-% \category{See
-%   \url{http://acm.org/about/class/1998/} for the full list of ACM
-%   classifiers. This section is required.}{}{}
-
-%\section{General comments}
-% \begin{itemize}
-% \item we should be clear (and careful) where we're talking about
-%   groups TMs, and where we're talking about individual performers
-% \item what's the terminology we're using for a ``performance''? Is it
-%   ``performance'', or is that too music-specific? ``Session''?
-%   Alternatives are ``activity'', or ``interaction'', but they're so
-%   general\ldots This decision has implications for our other
-%   terminology: participant vs performer vs musician, etc.
-% \item Title: maybe ``Traces through time: making sense of longitudinal
-%   open-ended interaction data'' or some such? I think that's our big
-%   contribution: it takes the in-the-wild, practitioner-driven research
-%   program stuff and tries to make some quantitative sense out of it
-%   post-hoc.
-% \end{itemize}
+\category{H.5.5.}{Information Interfaces and Presentation (e.g.
+  HCI)}{Sound and Music Computing}[Methodologies and techniques]
 
@@
   \caption{Collaborative musical interaction with our touch-screen
-    interfaces in three different contexts: (left to right) rehearsal
-    in a studio,
-    performance with percussion instruments, and performance with just
-    iPads.\label{fig:three-performance-contexts}}
+    interfaces in two different {\em contexts} and two different {\em instrumentations}: (left to right) rehearsal
+    in a studio with iPads only,
+    concert performance with iPads and percussion instruments, and concert performance with 
+    iPads only.\label{fig:three-performance-contexts}}
 \end{figure*}
@@
 acknowledged that understanding and evaluating interfaces in creative
-contexts can be challenging~\cite{Shneiderman:2007qv}. This problem is
+contexts can be very challenging~\cite{Shneiderman:2007qv}. This problem is
 even more pronounced in artistic \emph{performance}, where subjective
 interface evaluation through questionnaires or interviews may be both
-practically difficult and culturally inappropriate. When the interface
-is digital, however, it is still usually trivial to log interaction in
-these situations, which is useful for curation and
-preservation~\cite{England:2014ys}. The question is: can this log data
-be used not just as a record for posterity, but in support of a
-broader creative-practice led research program?
+practically difficult and culturally inappropriate. However, when an artistic interface
+is digital, it becomes possible to log artistic performances and these
+logs can be used for curation and
+preservation~\cite{England:2014ys}. A question we address is: can this log data
+be used to understand and support a
+broader creative-practice-led research program?
 
 In this note we will show how an archive of collaborative performance
-data logs can be subjected to a quantitative analysis based on
+logs can be subjected to a quantitative analysis based on
 transition probabilities between gestural states. We will demonstrate
@@
 
-Our archive of interaction data logs has been accumulated to support a
-long-term practice-led research program into simple iPad interfaces
+Our archive of interaction data-logs has been accumulated to support a
+long-term, practice-led, research program into simple iPad interfaces
 for group music-making similar to those described in Martin et
@@
 touch down, dragging movement and release---was time-stamped and
-logged to disk for later analysis. These sessions included
-formally-structured HCI experiments, where participants were
-interviewed or surveyed. They also included rehearsals without an
-audience and performances at arts festivals where creative goals were
-paramount; in these situations surveys or interviews would have been
-inappropriate and possibly disruptive. Some of the performances were
+logged to disk for later analysis. These sessions included {\em rehearsals} for upcoming concerts, without an
+audience, as well as curated {\em concerts}, in front of a live audience, where 
+performative goals were paramount. They also included a number of {\em recording sessions}
+for the purposes of formally-structured HCI experiments. In the recording
+sessions, the performers were also interviewed and surveyed-- something which would have 
+been disruptive and inappropriate for the concerts. Survey and interview data were not part of the
+corpus. 
+Some of the sessions were
 iPad-only and others used iPads together with other acoustic
@@
 Some of these session contexts are are shown
-Figure~\ref{fig:three-performance-contexts}) and other descriptive
-statistics about the sessions are shown in Table \ref{corpus-table}.
-In our corpus, the median length of sessions is 7m26s and the median
+Figure~\ref{fig:three-performance-contexts} and other descriptive
+statistics about the corpus are shown in Table \ref{corpus-table}.
+In our corpus, the median length of sessions is 7m26s (IS THIS STILL THE MEDIAN IGNORING THE
+DEMONSTRATION AND DATA GATHERING SESSIONS?) and the median
 number of participants per ensemble is four. Each session record is
-labelled with metadata about the session \emph{context} (performance,
-rehearsal, study), session \emph{type} (improvisation or composition),
-and \emph{instrumentation} (iPads-only, or iPads with acoustic
+labelled with metadata about its \emph{context} (concert,
+rehearsal, recording), its \emph{type} (improvisation or composition),
+and its \emph{instrumentation} (iPads-only, or iPads with acoustic
 percussion). The distribution of recorded sessions by context and type
@@
 \caption{
-  Representative data about the corpus of musical touch-screen
-  interaction sessions used in this study. These sessions were
-  rehearsals, performances, or studies and
-  participants improvised or performed compositions. Each record
-  consisted of touch-data recorded in CSV format and was later analysed
-  with a gestural classification system. Some participants were
-  present in multiple sessions, however this data was not tracked in
-  the corpus.\label{corpus-table}}
+  Representative statistics from the corpus of musical touch-screen
+  interaction sessions described in this note.  Each record in the corpus
+  consists of low level touch-data (such as touches begin, touches move, touches end) recorded in CSV format. 
+  Many of the participants were
+  present in multiple sessions, but this data was not tracked in
+  our (deidentified) corpus.\label{corpus-table}}
 \end{table}
@@
 interaction states to produce a transition model. This method has been
-applied to usability analysis of resource planning software for
-security applications~\cite{Kannampallil:2007fp} where the transition
-probability matrix of these state sequences was used to draw
-conclusions about the nature of this interactions. In the creative
-arts, editing commands at a computer interface to ``live code''
-computer music has also been coded as states and analysed using
-transition matrices~\cite{Swift:2014tya}, in both the textual and
+applied to usability analysis of many applications outside of the creative arts, for example~\cite{Kannampallil:2007fp}
+studied resource-planning software for
+security applications and used the transition
+probability matrix of these state-sequences to draw
+conclusions about the nature of interactions with that software. As an example from the creative
+arts, protocols obtained from a computer interface for ``live coding''
+computer music have been  coded as states and analysed using
+transition matrices~\cite{Swift:2014tya} in both the textual (editing) and
 musical dimensions, in order to arrive at conclusions regarding
@@
 
-In our corpus of iPad musical performances, logs of low-level touch
-data record each performer's micro-gestures. However, it has been
-established that musical performances on touch-screens can be seen in
-terms of vocabularies of continuous macro gestures that unfold over
-several seconds of touch interaction~\cite{Martin:2014cr}. Improvised
-performances can be then thought of as a sequence of such
+In our corpus of iPad musical performances, logs of low-level touch-data record each performer's micro-gestures. 
+Because it has been
+established that musical performances on touch-screens can be chuncked into
+vocabularies of continuous macro gestures that unfold over
+several seconds of touch interaction~\cite{Martin:2014cr}, these logs
+ can be thought of as a sequence of such
 states~\cite{Pressing:1988uo} for each performer.
-
-If the performers' touch interactions can be classified as belonging
-to one of $m$ distinct states, then over the course of a session each
+If a performers' touch interactions can be classified as belonging
+to one of $m$ distinct states, then, over the course of a session, each
 participant $X$ generates a length $N$ sequence of these interaction
@@
 process.
-
 If there is an (automated) classification process for identifying each
 participant's state sequence from their interaction data, then we can
-estimate the $m \times m$ transition matrix $P$ which characterises
-this process by setting each entry $p_{ij}$ to be the number of times
+estimate the $m \times m$ transition matrix (TM), $P$, which characterises
+this process by setting each entry, $p_{ij}$, to be the number of times
 state $j$ follows state $i$ in the sequence. The transition activity
@@
 To generate state sequences for our corpus we have used an automated
-method to classify logs of touch interaction, at a regular interval,
-into nine continuous touch-screen gestures, as described by Martin et
+method to sample the data at regular intervals and to classify the logs of touch events
+into the nine continuous touch-screen gestures identified by Martin et
 al~\cite{Martin:2015jk}. A Random Forest~\cite{Breiman:2001kx}
@@
 these matrices can be used to visualise the structure of individual
-performances~\cite{Swift:2014tya}, we wish to extract quantitative
+performances, as has been done in ~\cite{Swift:2014tya}, we also wish to extract quantitative
 measures from these TMs to compare the large number of sessions in our
 corpus.
-
 To do this we calculate two high-level quantities from each session:
@@
 
-Another useful measure that can be used on a transition matrix is its
+Another, evocative measure that can be defined on a TM is its
 entropy, defined in the information-theoretic\cite{Shannon:1948rt}
@@
 gestural space explored by the participants throughout the course of
-an interaction. Consider the degenerate case of a participant who
-alternates between two states for a whole performance: the flux in
+an interaction. Consider a degenerate case of a participant who
+alternates between just two states for a whole performance: the flux in
 this case will be maximal ($\mathrm{flux}(P) = 1$) since there are no
@@
   \caption{Distribution of the sessions in our interaction corpus by
-    performance context. Improvisations or compositions were repeated
-    several times in rehearsals and study sessions but only performed
+    session {\em context}. Improvisations or compositions were repeated
+    several times in rehearsals and recording sessions but only performed
     once in concerts.
@@
 improvisations. Compositions had lower
-flux in rehearsals than performances.
+flux in rehearsals than concerts.
 \label{fig:flux-entropy-boxplot}}
@@
 
-The distribution of flux and entropy with respect to session contexts,
-types and instrumentation are shown in Figure
+The distribution of flux and entropy with respect to session {\em contexts} (concert,
+rehearsal, recording),
+{\em types} (improvisation or composition) and {\em instrumentation} (iPads-only, or iPads with acoustic
+percussion) are shown in Figure
 \ref{fig:flux-entropy-boxplot}. A three-way ANOVA procedure was
@@
 
-All main effects on flux were found to be significant: performance
-context ($F(2,86) = 6.28, p < 0.01$), performance type
+All main effects on flux were found to be significant: session
+context ($F(2,86) = 6.28, p < 0.01$), session type
 ($F(1,86) = 7.21, p < 0.01$), and instrumentation
 ($F(1,86) = 20.06, p < 0.001$). Significant interaction effects were
-also found for performance context and type
-($F(1,86) = 6.18, p < 0.05$), and performance context and
+also found for session context and type
+($F(1,86) = 6.18, p < 0.05$), and session context and
 instrumentation ($F(1,86) = 10.48, p < 0.01$).
 
-For entropy, significant main effects were found due to performance
+For entropy, significant main effects were found due to session
 context ($F(2,86) = 12.10, p < 0.001$), and instrumentation
-($F(1,86) = 25.86$). The interaction of performance type and
-instrumentation was also found to be significant
+($F(1,86) = 25.86$) (WHAT IS THE P VALUE HERE?). There was no main effect for session type although
+ the interaction of session type and
+instrumentation was found to be significant
 ($F(1,86) = 9.93, p<0.01$).
@@
 These tests show that the different interaction styles that might be
-present in performance contexts, types and instrumentations are
-reflected in the values of flux and entropy of their transition
+present in the various session contexts, types and instrumentations in our corpus
+appear to be descriminated  by values of the flux and entropy of their transition
 matrices.
@@
           \includegraphics[width=\linewidth]{figures/type-section-flux-entropy}
-          \caption{Performances divided by performance type. \label{fig:type-section-flux-entropy}}
+          \caption{Sessions divided by session type. \label{fig:type-section-flux-entropy}}
     \end{subfigure}%
@@
           \includegraphics[width=\linewidth]{figures/context-section-flux-entropy}
-          \caption{Sessions divided by performance context. \label{fig:context-section-flux-entropy}}
+          \caption{Sessions divided by session context. \label{fig:context-section-flux-entropy}}
     \end{subfigure}
@@
 
-\subsection{Breaking performances into sections}
+\subsection{Breaking Sessions into sections}
 
-It is clear from these results that flux and entropy can be used to
-differentiate between performance styles and types, but can they be
-used to understand some of the internal structure of performances? A
-classical model for understanding temporal art-forms is the
+It appears from our data-mining results that both flux and entropy can be used to
+differentiate between session styles and types, but can they be
+used to understand some of the internal structure of sessions? Since Aristotle, a
+classical model for understanding temporal art-forms is the three-phase
 ``beginning, middle, end'' structure~\cite{Aristotle:350rt}. To
-investigate how this structure fits our corpus, we can divide the
-gestural-sequences of each session into thirds and calculate the
+investigate how this structure fits our corpus, we have divided the
+gestural-sequences of each session into three equal sections and have calculated the
 transition matrices given by each section.
@@
 Figure~\ref{fig:type-section-flux-entropy} shows that the flux and
-entropy of compositions are much more stable with respect to section
-than for improvisations. The stability of compositions is to be
+entropy of compositions are much more stable across the three sections than  are
+improvisations. This stability of compositions is to be
 expected since performers use the same gestures in each rehearsal and
-performance of these written scores, however it may be that different
+concert of written scores, however it may be that different
 kinds of compositions could result in different structural
@@
 different ideas. At the end of an improvisation, performers may change
-gesture less frequently while ``winding up'' the performance.
+gesture less frequently while ``winding up'' the concert performance.
 
 The gestural change in improvisation can be further explored by
-dividing the sessions by performance context. Figure
+dividing the sessions by session context. Figure
 \ref{fig:context-section-flux-entropy} shows this division for
@@
 in the ending section of studies and after the beginning section in
-rehearsals. In performances however, flux drops after the beginning
+rehearsals. In concerts however, flux drops after the beginning
 while entropy drops substantially in the ending section. This
@@
 section, where performers transition between fewer gestures in the
-final section of performances.
+final section of concert performances.
 
@@
 
-While the interaction session in our corpus have many different
+While the interaction sessions in our corpus have many different
 contexts, they are all part of a program of artistic and HCI research
@@
 shows the values of flux and entropy for each session in the corpus.
-Over the course of this research program, different ensembles and
+Over the course of this data collection, different ensembles and
 activities have been emphasised and some of these differences can be
-seen as clusters of performances with similar flux and entropy in the
+seen as clusters of performances having similar fluxes and entropies in the
 two plots. For instance, the earliest series of rehearsals and
-performances had flux between 0.05 and 0.15, while the period of
-activity after July 2014 had flux between 0.1 and 0.25. This change
-corresponded with an intensive series of study sessions and rehearsals
-where the participants developed a more ``fluctuating'' style of
-performance. The entropy between these two periods, however, remains
-at a similar level, showing the the space of available gestures that
-the performers explored was relatively stable.
+concerts, from early 2013 until July 2014, 
+had fluxes that were bounded between approximately 0.05 and 0.17.
+Just before and after July 2014, fluxes were bounded between 0.1 and 0.26. Finally,
+between early and mid 2015, fluxes were bounded between 0.08 and 0.32. Within these 
+blocks of performances, distinct clusters can be identified. Clusters in the entropy plots
+sometimes show similar trends to the fluxes and sometimes different. The higher
+upper bounds of fluxes close to July 2015 relative to those a year earlier may indicate
+the development of a more ``fluctuating'' style of performance. In the same period, the bound
+of entropies remained more similar indicating that the total space of available gestures continued
+to be explored in a relatively stable way.
 
@@
     Different performance styles are visible as clusters of similar
-    flux regardless of performance context.
+    flux regardless of session context.
     \label{fig:flux-entropy-through-time}}
@@
 
-In this note we have extended previous work using transition matrices
-to understand HCI processes with the application of two simple matrix
-measures, flux and entropy, both of which have been demonstrated to
-have significant discriminatory power with regards to a corpus of 95
-collaborative touch-screen interactions. Not only do these measures
-differentiate between different types of sessions and the internal
-structure of these sessions, they provide qualitative insight into the
-performances as well. For improvised performances, live performances
-had the lowest measure in both flux and entropy, while lab-studies had
-the highest. This may be due to the increased pressure on performers
-when improvising in front of an audience. The internal investigation
-of performances session revealed more structural variety in
-improvisations than compositions. In response, new works could be
-composed that specifically explore different gestural structures in
-touch-screen performance.
+In this note we have applied transition matrices
+to a data corpus of 95 collaborative musical performances using touch screens.
+We have applied two simple matrix
+measures, flux and entropy, to the data and have shown that
+ both of these measures appear to 
+have discriminatory power. Not only do these measures
+differentiate between different types of sessions, and the internal
+structure of these sessions, but they provide qualitative insight into the
+performances. For improvised performances, it turns out that our results showed that concerts
+had the lowest values of both flux and entropy, while recording sessions had
+the highest. This may be due to the increased pressure on performers to take fewer risks
+when improvising in front of an audience at a concert and indicates that perhaps rehearsals should 
+emphasise the development of such risk-taking. 
+Our results for the three phases of performance (beginning, middle, end)
+revealed more variation in
+improvisations than compositions. As a creative
+ response to this finding, 
+new works might be
+composed that specifically explore and emphasise different gestural sequences across the three phases.
 
-Data-mining logs of interactions with creative interfaces suggest
-further approaches for future work. The macro-gestures that these
-processes reveal could be used as the basis for new works of new media
-art~\cite{Manovich:2002ly}. Additionally, a wider dissemination of
-such log data could better support the replication of findings in HCI
-research~\cite{Wilson:2011ve}. Finally, flux and entropy proved useful
-in the analysis of our data corpus, but other transition matrix
-metrics may be more suitable for different interaction data.
+Our data-mining of interaction-logs from creative interfaces suggests
+further approaches for future work.
+Our flux and entropy measures have proven useful
+in the analysis of our data corpus, but many other transition matrix
+metrics might be found to be useful for these and different interaction data.
+Our assumption of equal intervals for the beginning, middle and end of performances could be 
+optimised with a goal to identifying a common, or even optimal, apportionment of time to each phase.  
+A wide dissemination of corpuses such as ours could better support the replication of findings in HCI
+research~\cite{Wilson:2011ve}. The macro-gestures that these
+processes reveal could be used as the basis for ``transcoding'' new works of new media
+art~\cite{Manovich:2002ly} from these data corpuses.
 
-HCI researchers are increasingly concerned with interactions that
-happen in-the-wild, where the creative collaborations of many users
-may be recorded, but where conducting surveys or interviews is
-innapropriate. We suggest that analysing interaction-state transitions
-using matrices and simple measures can lead to new insights into HCI
-interactions outside the lab.
+%HCI researchers are increasingly concerned with interactions that
+%happen in-the-wild, where the creative collaborations of many users
+%may be recorded, but where conducting surveys or interviews is
+%innapropriate. We suggest that analysing interaction-state transitions
+%using matrices and simple measures can lead to new insights into HCI
+%interactions outside the lab.
 
-%\pagebreak
+\pagebreak
 \bibliographystyle{SIGCHI-Reference-Format}

### 2015-09-23 flux-entropy-paper/flux-entropy-paper.tex
yes, it's still the median

@@
 statistics about the corpus are shown in Table \ref{corpus-table}.
-In our corpus, the median length of sessions is 7m26s (IS THIS STILL THE MEDIAN IGNORING THE
-DEMONSTRATION AND DATA GATHERING SESSIONS?) and the median
+In our corpus, the median length of sessions is 7m26s and the median
 number of participants per ensemble is four. Each session record is

### 2015-09-23 flux-entropy-paper/flux-entropy-paper.tex
adjusted comments to be in comments - address these tonight

@@
 states~\cite{Pressing:1988uo} for each performer.
+% TODO need to refine the meaning of this sentence - now Pressing
+% doesn't make sense as a citation.
 If a performers' touch interactions can be classified as belonging
@@
 context ($F(2,86) = 12.10, p < 0.001$), and instrumentation
-($F(1,86) = 25.86$) (WHAT IS THE P VALUE HERE?). There was no main effect for session type although
+($F(1,86) = 25.86$).
+% TODO (WHAT IS THE P VALUE HERE?). 
+There was no main effect for session type although
  the interaction of session type and

### 2015-09-24 flux-entropy-paper/flux-entropy-paper.tex
cleanups and nailing todos for results section

--- 2015-09-23 flux-entropy-paper/flux-entropy-paper.tex
@@
 
-%\title{Quantitative Evaluation of Touch-Screen Group Musical Performance with
- % Transition Matrix Measures}
 %\title{Mining interaction log data in long-term practice-led research}
@@
 
-%in a Creative Arts Practice}
-
 \numberofauthors{3}
@@
   \alignauthor{1st Author Name\\
-    \affaddr{Affiliation}\\
-    \affaddr{City, Country}\\
-    \email{e-mail address}}\\
-  \alignauthor{2nd Author Name\\
     \affaddr{Affiliation}\\
@@
 Because it has been
-established that musical performances on touch-screens can be chuncked into
+established that musical performances on touch-screens can contain 
 vocabularies of continuous macro gestures that unfold over
-several seconds of touch interaction~\cite{Martin:2014cr}, these logs
- can be thought of as a sequence of such
-states~\cite{Pressing:1988uo} for each performer.
+several seconds of touch interaction~\cite{Martin:2014cr}, and that
+gestural performances can be divided into sequences of such states~\cite{Pressing:1988uo}, these logs
+ can be used identify a sequence of such states for each performer.
 If a performers' touch interactions can be classified as belonging
@@
 context ($F(2,86) = 12.10, p < 0.001$), and instrumentation
-($F(1,86) = 25.86$) (WHAT IS THE P VALUE HERE?). There was no main effect for session type although
+($F(1,86) = 25.86, p < 0.001$).
+There was no main effect for session type although
  the interaction of session type and
@@
 matrices.
+
+% R> summary(aov(flux~performance_context*performance_type*instruments, perf.sessions))
+%                                                  Df Sum Sq Mean Sq F value
+% performance_context                               2 0.0267  0.0133    6.28
+% performance_type                                  1 0.0153  0.0153    7.21
+% instruments                                       1 0.0426  0.0426   20.06
+% performance_context:performance_type              1 0.0131  0.0131    6.18
+% performance_context:instruments                   1 0.0000  0.0000    0.02
+% performance_type:instruments                      1 0.0223  0.0223   10.48
+% performance_context:performance_type:instruments  1 0.0034  0.0034    1.62
+% Residuals                                        86 0.1828  0.0021        
+%                                                   Pr(>F)
+% performance_context                               0.0028
+% performance_type                                  0.0087
+% instruments                                      2.3e-05
+% performance_context:performance_type              0.0149
+% performance_context:instruments                   0.8954
+% performance_type:instruments                      0.0017
+% performance_context:performance_type:instruments  0.2065
+% Residuals                                               
+% R> summary(aov(entropy~performance_context*performance_type*instruments, perf.sessions))
+%                                                  Df Sum Sq Mean Sq F value
+% performance_context                               2   5.10    2.55   12.10
+% performance_type                                  1   0.31    0.31    1.48
+% instruments                                       1   5.46    5.46   25.87
+% performance_context:performance_type              1   0.21    0.21    0.99
+% performance_context:instruments                   1   0.10    0.10    0.49
+% performance_type:instruments                      1   2.09    2.09    9.93
+% performance_context:performance_type:instruments  1   0.01    0.01    0.04
+% Residuals                                        86  18.14    0.21        
+%                                                   Pr(>F)
+% performance_context                              2.3e-05
+% performance_type                                  0.2267
+% instruments                                      2.1e-06
+% performance_context:performance_type              0.3224
+% performance_context:instruments                   0.4878
+% performance_type:instruments                      0.0022
+% performance_context:performance_type:instruments  0.8396
+% Residuals                                               
+% R>
 
@@
 
-\subsection{Breaking Sessions into sections}
+\subsection{Breaking sessions into sections}
 

### 2015-09-24 R/analysis.R
added better colours and relabelled data frames for plots

@@ -4,6 +4,17 @@ library(grid)
 library(gridExtra)
 library(reshape2)
 
+
+
+
+
+
+
+perf.type.colours <- c("#1b9e77","#d95f02")
+perf.context.colours <- c("#7570b3","#e7298a","#66a61e")
+perf.structure.colours <- c("#e6ab02","#a6761d","#666666")
+perf.structure.colours <-  c("#edf8b1","#7fcdbb","#2c7fb8") # sequential
+
 chifig.3colours <- c("#e41a1c", "#377eb8", "#4daf4a")
 chifig.2colours <- c("#984ea3", "#ff7f00")
 chifig.5colours <- c("#e41a1c", "#377eb8", "#4daf4a", "#984ea3", "#ff7f00")
@@ -47,12 +58,19 @@ perf.sessions.long <-  subset(perf.sessions.long, performance_type %in% c("compo
 summary(perf.sessions.long)
 
 
+# Re label performance contexts for paper
+perf.sessions$performance_context <- factor(perf.sessions$performance_context, levels=c("data_collection", "demonstration", "performance", "rehearsal", "study"), labels=c("data", "demo", "concert", "rehearsal", "recording"))
+perf.sessions.long$performance_context <- factor(perf.sessions.long$performance_context, levels=c("data_collection", "demonstration", "performance", "rehearsal", "study"), labels=c("data", "demo", "concert", "rehearsal", "recording"))
+summary(perf.sessions)
+
+
 ## Initial Plots
 # cut out data_collection
 #valid.sessions <- (subset(valid.sessions, performance_context != "data_collection"))
 ## plots
 # Count of Session Data
-ggplot(perf.sessions, aes(performance_context)) + geom_bar(aes(fill = performance_type)) + theme(plot.margin=unit(rep(0,4), "cm"), legend.position = "right", legend.box = "vertical") + scale_fill_manual(values=chifig.2colours) + scale_x_discrete("performance context") + scale_y_continuous("no. of sessions")
+ggplot(perf.sessions, aes(performance_context)) + geom_bar(aes(fill = performance_type)) + theme(plot.margin=unit(rep(0,4), "cm"), legend.position = "right", legend.box = "vertical") + scale_fill_manual("performance type",values=perf.type.colours) + scale_x_discrete("performance context") + scale_y_continuous("no. of sessions")
+
 ggsave("../flux-entropy-paper/figures/sessions-count.pdf", width=5.4, height = 2)
[…]

### 2015-09-24 R/analysis.R
back to more sensible colours for context and type

@@ -3,18 +3,12 @@ library(MASS)
 library(grid)
 library(gridExtra)
 library(reshape2)
-
-
-
-
-
-
-
-perf.type.colours <- c("#1b9e77","#d95f02")
-perf.context.colours <- c("#7570b3","#e7298a","#66a61e")
+#perf.type.colours <- c("#1b9e77","#d95f02")
+perf.type.colours <- c("#984ea3", "#ff7f00")
+#perf.context.colours <- c("#7570b3","#e7298a","#66a61e")
+perf.context.colours <- c("#e41a1c", "#377eb8", "#4daf4a")
 perf.structure.colours <- c("#e6ab02","#a6761d","#666666")
 perf.structure.colours <-  c("#edf8b1","#7fcdbb","#2c7fb8") # sequential
-
 chifig.3colours <- c("#e41a1c", "#377eb8", "#4daf4a")
 chifig.2colours <- c("#984ea3", "#ff7f00")
 chifig.5colours <- c("#e41a1c", "#377eb8", "#4daf4a", "#984ea3", "#ff7f00")

### 2015-09-24 flux-entropy-paper/flux-entropy-paper.tex
adjusted a few figure captions

@@
 \begin{abstract}
-
   In this note we present an analysis of an interaction data archive
@@
   \caption{Collaborative musical interaction with our touch-screen
-    interfaces in two different {\em contexts} and two different {\em instrumentations}: (left to right) rehearsal
-    in a studio with iPads only,
-    concert performance with iPads and percussion instruments, and concert performance with 
-    iPads only.\label{fig:three-performance-contexts}}
+    interfaces in two different {\em contexts} and two different {\em
+      instrumentations}: (left to right) a rehearsal with iPads only,
+    concert performance with iPads and percussion instruments, and
+    concert performance with iPads
+    only.\label{fig:three-performance-contexts}}
 \end{figure*}
@@
   Many of the participants were
-  present in multiple sessions, but this data was not tracked in
-  our (deidentified) corpus.\label{corpus-table}}
+  present in multiple sessions, but this data was not tracked.\label{corpus-table}}
 \end{table}
@@
   \caption{Distributions of flux and entropy values by performance
-context. Study and rehearsal sessions had the highest flux for
+context. Recording and rehearsal sessions had the highest flux for
 improvisations. Compositions had lower
@@
           \includegraphics[width=\linewidth]{figures/type-section-flux-entropy}
-          \caption{Sessions divided by session type. \label{fig:type-section-flux-entropy}}
+          \caption{Sessions divided by performance type. \label{fig:type-section-flux-entropy}}
     \end{subfigure}%
@@
           \includegraphics[width=\linewidth]{figures/context-section-flux-entropy}
-          \caption{Sessions divided by session context. \label{fig:context-section-flux-entropy}}
+          \caption{Improvisation performances divided by session context. \label{fig:context-section-flux-entropy}}
     \end{subfigure}

### 2015-09-24 flux-entropy-paper/flux-entropy-paper.tex
so close!

@@
 long-term, practice-led, research program into simple iPad interfaces
-for group music-making similar to those described in Martin et
-al~\cite{Martin:2014cr}. Over a period of more than two years (April
+for group music-making. Over a period of more than two years (April
 2013--June 2015), we collected data from group music-making sessions
@@
 
-In all of our 95 interaction sessions, every touch interaction---every
-touch down, dragging movement and release---was time-stamped and
-logged to disk for later analysis. These sessions included {\em rehearsals} for upcoming concerts, without an
-audience, as well as curated {\em concerts}, in front of a live audience, where 
-performative goals were paramount. They also included a number of {\em recording sessions}
+In all of our 95 interaction sessions, every touch interaction---
+touches begin, touches move, touches end ---was time-stamped and
+logged to disk for later analysis. These sessions included {\em rehearsals} without an
+audience, as well as {\em concerts}, in front of a live audience, where 
+performative goals were paramount. They also included a number of {\em
+  recording} sessions
 for the purposes of formally-structured HCI experiments. In the recording
@@
 percussion notation using the apps.
-
 Some of these session contexts are are shown
@@
   interaction sessions described in this note.  Each record in the corpus
-  consists of low level touch-data (such as touches begin, touches move, touches end) recorded in CSV format. 
+  consists of low level touch-data recorded in CSV format. 
   Many of the participants were
@@
 
-\section{Macro Gestures, Transition Matrices and Matrix Measures}
+\section{Macro Gestures and Transition Matrices}
 
@@
 
-\section{Quantitative transition matrix statistics}
+\section{Quantitative Transition Matrix Measures}
 \label{sec:underst-impr-group}

### 2015-09-24 flux-entropy-paper/flux-entropy-paper.tex
minor adjustment

@@
 
-In all of our 95 interaction sessions, every touch interaction---
-touches begin, touches move, touches end ---was time-stamped and
+In all of our 95 interaction sessions, every touch interaction 
+(touches begin, touches move, touches end) was time-stamped and
 logged to disk for later analysis. These sessions included {\em rehearsals} without an

### 2015-09-24 flux-entropy-paper/flux-entropy-paper.tex
now fits on 4 pages phew!

@@
 While the interaction sessions recorded in our corpus were performed
-by many different participants with varying experience levels, in
+by many different participants in
 different contexts and on different instrumental setups, the common
@@
 \label{differentiating-interaction-sessions}
-
-\begin{figure} \centering
-  \includegraphics[width=\linewidth]{figures/context-flux-entropy-boxplot}
-  \caption{Distributions of flux and entropy values by performance
-context. Recording and rehearsal sessions had the highest flux for
-improvisations. Compositions had lower
-flux in rehearsals than concerts.
-\label{fig:flux-entropy-boxplot}}
-\end{figure}
 
@@
 % R>
+
+\begin{figure} \centering
+  \includegraphics[width=\linewidth]{figures/context-flux-entropy-boxplot}
+  \caption{Distributions of flux and entropy values by performance
+context. Recording and rehearsal sessions had the highest flux for
+improvisations. Compositions had lower
+flux in rehearsals than concerts.
+\label{fig:flux-entropy-boxplot}}
+\end{figure}
 
@@
 
-%\pagebreak
+\pagebreak
 \bibliographystyle{SIGCHI-Reference-Format}

### 2015-09-24 flux-entropy-paper/flux-entropy-paper.tex
final updates from Michael and added DOIs

@@
 
-In this note we will show how an archive of collaborative performance
-logs can be subjected to a quantitative analysis based on
-transition probabilities between gestural states. We will demonstrate
-that two measures on transition matrices, flux and entropy, show
-significant differences between different kinds of performances, and
-give insight into the internal structure of these performances.
+In this note we show how an archive of collaborative performance logs
+can be quantitatively analysed using estimated transition
+probabilities between gestural states. We will demonstrate that two
+measures on transition matrices, flux and entropy, show significant
+differences between different kinds of performances, and yield insight
+into the internal structure of these performances.
 
@@
 2013--June 2015), we collected data from group music-making sessions
-where the participants used our custom-built, touch-screen apps on
-Apple iPad devices. Although several different apps have been included
-in our corpus, they shared a common mode of free-form touch
-interaction where tapping produced short sounds and swirling or
-swiping produced long sounds. Apart from these similarities, the apps
-featured a variety of sound palettes, different kinds of visual
-feedback, and different kinds of networked interactions (between the
-iPads in an ensemble and with a central server).
+where participants used our custom-built, touch-screen apps on Apple
+iPad devices. Although several different apps have been included in
+our corpus, they shared a common mode of free-form touch interaction
+where tapping produced short sounds and swirling or swiping produced
+long sounds. Apart from these similarities, the apps featured a
+variety of sound palettes, different kinds of visual feedback, and
+different kinds of networked interactions (between iPads in an
+ensemble and with a central server).
 
@@
   Representative statistics from the corpus of musical touch-screen
-  interaction sessions described in this note.  Each record in the corpus
+  interaction sessions.  Each record in the corpus
   consists of low level touch-data recorded in CSV format. 
@@
 
+\break 
 \section{Macro Gestures and Transition Matrices}
@@
 A useful way of modelling user interfaces is to consider their
-theoretical configurations as state vectors and to deploy the methods
-of matrix algebra to describe transitions between these states during
+theoretical configurations as state vectors and to describe transitions between these states during
 an interaction sequence~\cite{Thimbleby:2001kq, Thimbleby:2004fj}. An
@@
 Because it has been
-established that musical performances on touch-screens can contain 
+established that musical performances on touch-screens define  
 vocabularies of continuous macro gestures that unfold over
@@
 participant $X$ generates a length $N$ sequence of these interaction
-states $X_n$ ($n = 1, \ldots, N$) which can be thought of as Markov
+states $X_n$ ($n = 1, \ldots, N$) which can be modelled as a Markov
 process.
@@
 estimate the $m \times m$ transition matrix (TM), $P$, which characterises
-this process by setting each entry, $p_{ij}$, to be the number of times
+this process by setting each entry, $p_{ij}$, to be the proportion of times
 state $j$ follows state $i$ in the sequence. The transition activity
@@
 return 0 when participants never change state, and 1 when participants
-never stay in the same state for two measurements in a row.
+never stay in the same state for two consecutive measurements.
 
@@
 
-In the rest of this note we will use these matrix measures to answer
+We use these matrix measures to answer
 two questions. First, do the transition matrix measures allow us to
@@
 
-The distribution of flux and entropy with respect to session {\em contexts} (concert,
+The distributions of flux and entropy with respect to session {\em contexts} (concert,
 rehearsal, recording),
@@
 \ref{fig:flux-entropy-boxplot}. A three-way ANOVA procedure was
-performed on the data set to find whether the metadata could
-significantly predict the entropy and flux values.
+performed to find whether the metadata could 
+successfully predict the entropy and flux values.
 
@@
 
-These tests show that the different interaction styles that might be
+These tests suggest that the different interaction styles that might be
 present in the various session contexts, types and instrumentations in our corpus
-appear to be descriminated  by values of the flux and entropy of their transition
+appear to be discriminated  by values of the flux and entropy of their transition
 matrices.
@@
 
-It appears from our data-mining results that both flux and entropy can be used to
+It appears from our data-mining results that each of flux and entropy can be used to
 differentiate between session styles and types, but can they be
@@
 ``beginning, middle, end'' structure~\cite{Aristotle:350rt}. To
-investigate how this structure fits our corpus, we have divided the
-gestural-sequences of each session into three equal sections and have calculated the
+investigate how this structure fits our corpus, we divided the
+gestural-sequences of each session into three equal sections and calculated the
 transition matrices given by each section.
@@
 kinds of compositions could result in different structural
-differences. Both measures have a downward trend through sections in
+patterns. Both measures have a downward trend through sections in
 improvisations with Kruskal-Wallis tests showing a significant effect
@@
 \ref{fig:context-section-flux-entropy} shows this division for
-improvisations only. In rehearsals and studies the distribution of
-flux and entropy over sections has a similar shape, both measures drop
-in the ending section of studies and after the beginning section in
+improvisations only. In rehearsals and recordings the distributions of
+flux and entropy over sections has a similar shape, both measures dropping
+in the ending section of recordings and after the beginning section in
 rehearsals. In concerts however, flux drops after the beginning
@@
 seen as clusters of performances having similar fluxes and entropies in the
-two plots. For instance, the earliest series of rehearsals and
+two panels. For instance, the earliest series of rehearsals and
 concerts, from early 2013 until July 2014, 
-had fluxes that were bounded between approximately 0.05 and 0.17.
-Just before and after July 2014, fluxes were bounded between 0.1 and 0.26. Finally,
-between early and mid 2015, fluxes were bounded between 0.08 and 0.32. Within these 
-blocks of performances, distinct clusters can be identified. Clusters in the entropy plots
-sometimes show similar trends to the fluxes and sometimes different. The higher
+had fluxes that were between approximately 0.05 and 0.17.
+Just before and after July 2014, fluxes were between 0.1 and 0.26. Finally,
+between early and mid 2015, fluxes were between 0.08 and 0.32. Within these 
+blocks of performances, distinct clusters can be identified. The higher
 upper bounds of fluxes close to July 2015 relative to those a year earlier may indicate
-the development of a more ``fluctuating'' style of performance. In the same period, the bound
-of entropies remained more similar indicating that the total space of available gestures continued
+the development of a more ``fluctuating'' style of performance over time. In the same period, the bound
+of entropies remained more consistent indicating that the total space of available gestures continued
 to be explored in a relatively stable way.
@@
 
-In this note we have applied transition matrices
-to a data corpus of 95 collaborative musical performances using touch screens.
+In this note we estimated transition matrices 
+for a data corpus of 95 collaborative musical performances using touch screens.
 We have applied two simple matrix
@@
 structure of these sessions, but they provide qualitative insight into the
-performances. For improvised performances, it turns out that our results showed that concerts
+performances. For improvised performances, our results showed that concerts
 had the lowest values of both flux and entropy, while recording sessions had

### 2015-09-24 flux-entropy-paper/flux-entropy-paper.tex
cut some comments
final updates! uploading now.

@@
   \begin{document}
-
-%\title{Mining interaction log data in long-term practice-led research}
 
@@
 percussion notation using the apps.
-Some of these session contexts are are shown
+Some of these session contexts are are shown in 
 Figure~\ref{fig:three-performance-contexts} and other descriptive
@@
 is shown in Figure \ref{fig:count-data}.
-
-% It is a feature of creative expression in the arts that highly skilled
-% practitioners exploit simple tools to produce works of beauty whether
-% it be static paintings or dynamic performance. 
-% In previous work~\cite{Martin:2014cr} researchers have shown that
-% trained percussionists can exploit a very simple iPad interface to
-% discover new categories of musical gestures.
-% In that paper, iPad apps
-% were studied where tapping was mapped to produce produced short
-% sounds, while swiping produced long sounds with the volume
-% proportional to the velocity of the moving touch point. 
-% It turned out that trained percussionists working together in an
-% ensemble were able to discover and repeatedly make use of five macro
-% gestures made up of taps and swipes: fast taps, slow taps, random
-% taps, short swipes and swirls. This process of discovery of
-% interesting musical gestures from a simple interface is analogous to
-% what percussionists do with acoustic instruments -- drums and bowls
-% and bells are often tapped, stroked and hit in novel ways to
-% communicate meaning and emotion through music~\cite{Schick:2006fk}.
 
@@
 gestural performances can be divided into sequences of such states~\cite{Pressing:1988uo}, these logs
- can be used identify a sequence of such states for each performer.
+ can be used to identify a sequence of such states for each performer.
 If a performers' touch interactions can be classified as belonging
@@
 
-%HCI researchers are increasingly concerned with interactions that
-%happen in-the-wild, where the creative collaborations of many users
-%may be recorded, but where conducting surveys or interviews is
-%innapropriate. We suggest that analysing interaction-state transitions
-%using matrices and simple measures can lead to new insights into HCI
-%interactions outside the lab.
-
-\pagebreak
+\break
 \bibliographystyle{SIGCHI-Reference-Format}

### 2016-01-04 CHI2016-DataMiningPaper-Rebuttal.md
updates with CHI2016 rebuttal

Thanks to the reviewers and ACs for their careful consideration, in particular, we wish to thank R2 for extremely insightful and useful comments. We think that we can address R2's criticisms to allow our findings to be more confidently interpreted. We will also address suggested changes from all of the reviewers to improve clarity and add a citation of Ge Wang's related work.

In our note, we showed that interaction-protocol collection and analysis can yield interesting insights into artistic processes where other evaluations would be impractical. Studying a dataset from a real performance practice is called "significant and realistic" by R3 who noted that our findings are a "concrete research contribution". Other reviewers were also positive and we hope that given our response to the specific criticisms our note will be accepted.

It should be mentioned that most objections refer to compositions, but these data are less important than the improvisations in our analysis and in the note's contribution. We acknowledge that the research questions could be more prominently situated and we will move them from page 3 to the front matter of the note.

# Reviewers' major questions:

- Effect of Tempo (R2):

R2 is correct in suggesting that a kind of "gestural tempo" is encoded in our transition matrices by the timing of the gestural classifier (1s steps, 5s window), however we argue that this is not a significant problem for interpreting the data.

Most of our data relates to improvisations where tempo would affect the internal speed of gestures but not the rate of transition. In this case rate of transition through different gestures is equivalent to performers' "adventurousness" in the gestural space. The compositions used in our research were certainly played in different tempi which may have affected the gestural tempo, but our analysis suggests that the session-lengths of compositions did not significantly correlate with flux or entropy.

We will include these points in our discussion but our claim - that flux and entropy can differentiate between performance types and contexts – still holds.

- Distributions of Compositions (R2):

Three different compositions were present in the data. Out of 23 composition sessions, 17 were of one piece, with the other two accounting for 4 and 2 sessions. For the first two compositions, the distribution was close to 25% concerts, 75% rehearsals, and the last had one rehearsal and one performance. We will clarify that the distribution between concert and rehearsal sessions was quite similar across all three compositions, so that the change in session type is the most likely source of the change in Flux and Entropy between the performance contexts.

- Transition Matrices over Ensembles/Individuals (R2):

We agree that averaging matrices for ensembles results in an overview statistic. In Figure 3 and Figure 5, the statistics have been calculated for ensembles as this is more appropriate for the broader differentiation between performance contexts and types. In Figure 4, they are calculated for individual performers to analyse local behaviour inside each session. We will state this difference clearly in a revised note.

- Goal of the "Breaking Sessions into Sections" analysis (R2):

This section and Figure 4 are mainly focussed on improvisations. The goal is to investigate the internal structure of improvised playing and to see if these sessions do show evidence of a beginning/middle/end structure which we have shown statistically for flux. Compositions are included in Figure 4a to contrast with the clear structure of the improvisations. Even though, as R2 has suggested, the thirds division may not be optimal for identifying segments, structural effects can still be seen in the figures and in the results of a Kruskal-Wallis test for flux. Knowing more about the structure of improvisations will be useful for educational work with these apps, and for future app designs which might e.g. attempt to respond in particular ways in an "ending" phase. Future work could determine optimal divisions inside session.

- Why these Metrics? (R1,R3):

R1 has suggested some alternative metrics for measuring transitions and asks for further discussion, while R3 asks for more theoretical motivation. The advantage of flux and entropy is that they are simple to calculate and can be motivated with the intuitive performative examples described in the note. Simonton's "melodic originality" metric is 
defined by a "ground truth" corpus of 15000 melodies by 500 notable classical composers. It would be difficult to motivate such a definition from our more modest corpus. Kerne et al.'s "ideation metrics" require a classification of an "idea" in the data - single transitions in our data are too low level to be mapped into this system. Our metrics allow very low-level data to be investigated with minimal interpretation. We will cite these very interesting references to help situate our note.

### 2016-01-04 R/analysis.R

@@ -1,3 +1,13 @@
+install.packages("ggplot2")
+
+install.packages("MASS")
+
+install.packages("grid")
+
+install.packages("gridExtra")
+
+install.packages("reshape2")
+
 library(ggplot2)
 library(MASS)
 library(grid)
@@ -57,7 +67,15 @@ perf.sessions$performance_context <- factor(perf.sessions$performance_context, l
 perf.sessions.long$performance_context <- factor(perf.sessions.long$performance_context, levels=c("data_collection", "demonstration", "performance", "rehearsal", "study"), labels=c("data", "demo", "concert", "rehearsal", "recording"))
 summary(perf.sessions)
 
+ggplot((subset(perf.sessions, performance_type == "composition")), aes(length_seconds,flux)) + geom_point(aes(colour = performance_context, shape = instruments, size = number_performers)) + geom_smooth(method = "lm")
 
+ggplot((subset(perf.sessions, performance_type == "composition")), aes(length_seconds,entropy)) + geom_point(aes(colour = performance_context, shape = instruments, size = number_performers)) + geom_smooth(method = "lm")
+
+ggplot((subset(perf.sessions, performance_type == "composition")), aes(number_performers,flux)) + geom_point(aes(colour = performance_context, shape = instruments, size = length_seconds)) + geom_smooth(method = "lm")
+
+ggplot((subset(perf.sessions, performance_type == "composition")), aes(performance_context,entropy)) + geom_point(aes(colour = performance_context, shape = instruments, size = length_seconds)) + geom_smooth(method = "lm")
+
+ggplot((subset(perf.sessions, performance_type == "composition")), aes(performance_context,length_seconds)) + geom_boxplot() + geom_point(aes(colour = performance_context, shape = instruments, size = number_performers)) + geom_smooth(method = "lm") 
 ## Initial Plots
 # cut out data_collection
 #valid.sessions <- (subset(valid.sessions, performance_context != "data_collection"))
